---
title: "Regression to predict IUCN status"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(terra)
library(sf)
library(janitor)
```

This document will run a regression to predict AquaX species IUCN status on the number of EEZs its range intersects, whether its range is in the area beyond national jurisdictions (ABNJ), its total range area, its ecoregion, and taxon.

## Prep data 

The number of EEZs and IUCN status is prepared in other files. Here, we need to determine whether each species range intersects with the ABNJ, its total range area, and its ecoregin. The species taxon is included in the AquaX data.

```{r}
# Load ABNJ (high seas)
abnj <-  st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_High_Seas_v2_20241010')
```

```{r}
# Load MEOW (marine ecoregions of the world)
meow <-  st_read('/home/shares/usgs-cap-rangeshifts/spatial/MEOW-TNC')
```

```{r}
ggplot() +
  geom_sf(data = meow, aes(geometry = geometry, fill = REALM))
```

```{r}
meow_realm <- meow %>% 
  group_by(REALM) %>% 
  summarize(geometry = st_combine(geometry)) 
```

```{r}
ggplot() +
  geom_sf(data = meow_realm %>% filter(REALM == "Central Indo-Pacific"),
          aes(geometry = geometry, fill = REALM))
```

```{r}
load('/home/shares/usgs-cap-rangeshifts/data_aquax/2026_OHARA_TST_MORPHO/SDM/FINAL_EMSDM_EMMEAN_SP_550138.Rdata')
```

```{r}
f_test <- FINALEMMEAN %>% 
  select(x, y, Current, cutoff) %>% 
  mutate(historical = case_when(Current < cutoff ~ 0,
                                Current >= cutoff ~ 1)) %>% 
  # Filter for presence
  filter(historical == 1)

f_test_rast <- rast(f_test)
```

```{r}
# Extract maximum value from species range in ABNJ (would be 1 if present in ABNJ)
max_value_hs <- extract(f_test_rast$historical, abnj, fun = max, na.rm = TRUE)
```

```{r}
# Count species pixels in each MEOW realm
meow_count <- extract(f_test_rast$historical, meow_realm, fun = sum, na.rm= TRUE)

colnames(meow_count)[2] <- "presence_count"

meow_count_region <- cbind(meow_realm, meow_count)
```

```{r}
assigned_ecoregion <- meow_count_region[which.max(meow_count_region$presence_count), ]
```

## Build species information

```{r}
# Not on server
load('data/META_02122025.Rdata')
```

Associate SDM files with the metadata for the 564 sampled species
```{r}
spp_fs <- list.files('/home/shares/usgs-cap-rangeshifts/data_aquax/2026_OHARA_TST_MORPHO/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(META, by = c('aphiaid' = 'AphiaID'))
```

# Test sample 
A few issues to work out if regression is promising: global species, those species that might not fall into an eco realm, etc.

```{r}
#set.seed(123)
test_spp_30 <- df[1:15,]
files_30 <- test_spp_30$f
aphiaid_30 <- test_spp_30$aphiaid
group_30 <- test_spp_30$GROUP
#load(test_spp_30$f)

```

```{r}
extract_region <- function(id, df) {
  results_list <- list()
  
  spp_filter <- df %>% 
    filter(aphiaid == id) 
  
  load(spp_filter$f)
  
  spp_presence <- FINALEMMEAN %>% 
  mutate(aphiaid = id) %>% 
  select(x, y, aphiaid, Current, cutoff) %>% 
  mutate(current_presence = case_when(Current < cutoff ~ 0,
                                Current >= cutoff ~ 1)) %>%
  # filter only for presence (1)
  filter(current_presence == 1)
  
  r_spp <- rast(spp_presence, crs = "EPSG:4326")
  
  # high seas flag (abnj)
  hs_extract <- extract(r_spp$current_presence, abnj, fun = max, na.rm = TRUE)
  # if max is 1 then spp is in high seas
  # get 1 if max is 1, will give NA for NA (options are 1 and NA)
  in_high_seas <- ifelse(hs_extract[, 2] == 1, 1, 0)
  
  # ecoregion (max pixels in ecoregion)
  realm_extract <- extract(r_spp$current_presence, meow_realm, fun = sum, na.rm = TRUE)
  # get max
  colnames(realm_extract)[2] <- "presence_count"
  realm_count <- cbind(st_drop_geometry(meow_realm), realm_extract)
  
  # assign NA for no ecoregion (otherwise error)
  if (sum(realm_count$presence_count, na.rm = TRUE) > 0) {
    assigned_realm <- realm_count[which.max(realm_count$presence_count), 1]
  } else {
    assigned_realm <- NA # species range doesn't overlap with any MEOW realm
  }
  
  results_df <- data.frame(
    aphia_id = id,
    abnj_flag = in_high_seas,
    eco_realm = assigned_realm
  )
  
  #spp_region_df <- rbind(results_df)

  return(results_df)
}
```

```{r}
region_df_30 <- lapply(aphiaid_30, 
                       FUN = extract_region,
                       df = test_spp_30)
```


```{r}
region_df_30_df <- bind_rows(region_df_30)
```

## Run full sample, try parallel

```{r}
aphiaid <- df$aphiaid
```

```{r}
# Time (7 minutes)
system.time({
  out_list <- parallel::mclapply(aphiaid, FUN = extract_region, df = df, mc.cores = 10)
})
```

```{r}
out_list_df <- bind_rows(out_list) %>% 
  rename(aphiaid = aphia_id) %>% 
  # Replace NA with 0 in flag
  replace_na(list(abnj_flag = 0))
```

## Add in IUCN status, EEZ count, taxon, total pixels (could also calculate area)

```{r}
# Merge in group from df 
df_group <- df %>% 
  select(aphiaid, GROUP, ScientificName, range_pixels)

spp_df_group <- left_join(out_list_df, df_group)
```

# Read in EEZ data

```{r}
eez_list_status <- read.csv("/home/shares/usgs-cap-rangeshifts/data_aquax/outputs/spp_eez_pxl_count_041626.csv")
```

```{r}
# How many EEZs are each spp present in for current period?
eez_count_all <- eez_list_status %>% 
  group_by(aphiaid) %>% 
  summarize(current_eez_count = sum(n_current > 0)) %>% 
  ungroup() 
```

```{r}
# Merge in EEZ count
spp_df_group_eez <- left_join(spp_df_group, eez_count_all) %>% 
  mutate(worms_name = tolower(ScientificName)) %>% 
  select(!ScientificName)
```

```{r}
# IUCN threat info
iucn <- read.csv("/home/shares/usgs-cap-rangeshifts/data_iucn/iucn_to_worms_match_redlist_category.csv")
```

```{r}
# Merge in IUCN info based on the worms_name
spp_df_group_eez_iucn <- left_join(spp_df_group_eez, 
                                   # merge in IUCN data and filter for species (no subpops)
                                   iucn %>%  filter(species == TRUE))
```

```{r}
# Drop those without IUCN sid
spp_df_iucn <- spp_df_group_eez_iucn %>% 
  filter(!is.na(iucn_sid))
```

```{r}
# Assign numeric values to IUCN categories 
spp_df_iucn_num <- spp_df_iucn %>% 
  mutate(iucn_num = case_when(iucn_category_code == "DD" ~ NA,
                              iucn_category_code == "LC" ~ 0,
                              iucn_category_code == "NT" ~ 0.2,
                              iucn_category_code == "VU" ~ 0.4,
                              iucn_category_code == "EN" ~ 0.6,
                              iucn_category_code == "CR" ~ 0.8,
                              iucn_category_code == "EX" ~ 1))
```

```{r}
# Run regression 
iucn_reg <- lm(iucn_num ~ current_eez_count + abnj_flag + log(range_pixels) +
                 eco_realm + GROUP, data = spp_df_iucn_num)
```

```{r}
summary(iucn_reg)
```

