---
title: "EEZs and AquaX Redlist"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(terra)
library(sf)
library(here)
library(janitor)
library(arrow)
```

## EEZ 

```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
# Create a blank raster matching AquaX resolution (EEZs should be large enough to just do 0.05 degree raster)
eez_raster_blank <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05)
```

```{r}
eez_raster_output <- rasterize(eez, eez_raster_blank, field = "GEONAME")
```

## Prep AquaX redlist data

```{r}
# read in meta data on the aquax sample of redlist species
load("/home/shares/usgs-cap-rangeshifts/data_aquax/iucn/meta.Rdata")
```

```{r}
# IUCN threat info
iucn <- read.csv("/home/shares/usgs-cap-rangeshifts/data_iucn/iucn_to_worms_match_redlist_category.csv")
```

```{r}
meta_clean <- meta %>% 
  mutate(worms_name = tolower(ScientificName)) %>% 
  select(!ScientificName) %>% 
  # simplify for now for easier viewing
  select(AphiaID, SPID, worms_name, GROUP, NR_CR, range_pixels, SDM)
```

```{r}
# Merge in IUCN info based on the worms_name
meta_iucn <- left_join(meta_clean, 
                       # merge in IUCN data and filter for species (no subpops)
                       iucn %>%  filter(species == TRUE),
                       by = "worms_name")
```

```{r}
# Drop those without IUCN sid (get 13,132 species remaining)
meta_iucn_no_na <- meta_iucn %>% 
  filter(!is.na(iucn_sid))
```

```{r}
# May be duplicates (90 cases where AquaX/worms is two or more species in IUCN)
# Might just remove these for now
df_worms_iucn_dupes <- meta_iucn_no_na %>% 
  get_dupes(worms_name)

dupes <- unique(df_worms_iucn_dupes$worms_name)
```

```{r}
# Drop those 90 worms species that are representing 2+ IUCN species 
# 12,940 species remaining
meta_iucn_no_na_no_dupes <- meta_iucn_no_na %>% 
  filter(!worms_name %in% dupes)
```

```{r}
# For now, focusing on only threatened species (1,013 species)
meta_iucn_threatened <- meta_iucn_no_na_no_dupes %>% 
  filter(iucn_category %in% c("Critically Endangered",
                              "Endangered", "Vulnerable"))
```

```{r}
# Get AphiaID of threatened species
aphiaid_threatened <- meta_iucn_threatened$AphiaID
```

## Prep AquaX data
Here, there are just the 12,238 files -not totally sure why the files and meta data don't exactly line up. But when I filter for the 1,013 threatened AphiaIDs, I get 906 species.

```{r}
spp_fs <- list.files('/home/shares/usgs-cap-rangeshifts/data_aquax/iucn/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(meta, by = c('aphiaid' = 'AphiaID'))
```

```{r}
# Filter for the AphiaID of the threatend species (906)
df_threatened <- df %>% 
  filter(aphiaid %in% aphiaid_threatened)
```

```{r}
test_f <- read_parquet("/home/shares/usgs-cap-rangeshifts/data_aquax/iucn/SDM/FINAL_EMSDM_EMMEAN_SP_1006832.parquet")
```

```{r}
set.seed(123)
test_spp_30 <- slice_sample(df_threatened, n = 30)
files_30 <- test_spp_30$f
aphiaid_30 <- test_spp_30$aphiaid
```

```{r}
extract_eez <- function(id, df) {
  spp_filter <- df %>% 
    filter(aphiaid == id) 
  
  spp_f <- read_parquet(spp_filter$f)
  
  spp <- spp_f %>% 
    mutate(aphiaid = id) 
  
  r_spp <- rast(spp, crs = "EPSG:4326")
  
  # extend spp to the same extent as eez
  r_spp_ex <- extend(r_spp, ext(eez_raster_output))
  
  r_spp_eez <- c(r_spp_ex, eez_raster_output)
  
  spp_df <- as.data.frame(r_spp_eez, xy = TRUE) %>% 
    filter(!is.na(GEONAME)) %>%
    filter(!is.na(Current) | !is.na(RCP85_2100)) %>%
    group_by(GEONAME, aphiaid) %>%
    summarize(n_current = sum(Current >= cutoff),
              n_RCP85_2100 = sum(RCP85_2100 >= cutoff))
  
  return(spp_df)
}
```

```{r}
# Try // - took 2 min
system.time({
  eez_list <- parallel::mclapply(aphiaid_30, FUN = extract_eez, df = test_spp_30, mc.cores = 10)
})
```

```{r}
eez_list_all <- bind_rows(eez_list) %>% 
  mutate(status = case_when(n_current == 0 &  n_RCP85_2100 > 0 ~ 'entry into new',
                            n_current > 0  & n_RCP85_2100 == 0 ~ 'exit from historical',
                            n_current == 0 & n_RCP85_2100 == 0 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

## Test out full threatened sample

```{r}
files <- df_threatened$f
aphiaid <- df_threatened$aphiaid
```

```{r}
# Try // - took 31 min for 906 species
system.time({
  eez_list_threatened <- parallel::mclapply(aphiaid, FUN = extract_eez, df = df_threatened, mc.cores = 20)
})
```

```{r}
eez_list_threatened <- bind_rows(eez_list_threatened) %>% 
  mutate(status = case_when(n_current == 0 &  n_RCP85_2100 > 0 ~ 'entry into new',
                            n_current > 0  & n_RCP85_2100 == 0 ~ 'exit from historical',
                            n_current == 0 & n_RCP85_2100 == 0 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

```{r}
write_csv(eez_list_threatened, "/home/shares/usgs-cap-rangeshifts/data_aquax/outputs/iucn_threatened_spp_eez_pxl_count_050426.csv")
```

```{r}
# How many EEZs are each spp present in for current and future?
eez_count_threatened_all <- eez_list_threatened %>% 
  # Replace NA with 0 for sum
  replace_na(list(n_current = 0)) %>% 
  replace_na(list(n_RCP85_2100 = 0)) %>% 
  group_by(aphiaid) %>% 
  summarize(current_eez_count = sum(n_current > 0),
            RCP85_2100_eez_count = sum(n_RCP85_2100 > 0)) %>% 
  ungroup() #%>% 
  #pivot_longer(cols = c(current, RCP85_2100),
               #names_to = "timeframe", values_to = "eez_count")

```

```{r}
ggplot() +
  geom_point(data = eez_count_threatened_all, aes(x = current_eez_count, y = RCP85_2100_eez_count),
             alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "maroon") +
  labs(title = "IUCN Threatened Species EEZ count")
```

```{r}
ggplot() +
  geom_point(data = eez_count_threatened_iucn, aes(x = current_eez_count, y = RCP85_2100_eez_count, color = iucn_category_code),
             alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "maroon") +
  facet_wrap(~iucn_category_code) +
  theme(legend.position = "bottom")
```

```{r}
# Of the threatened species, which are maintaining EEZ count, which are increasing, which are decreasing
eez_count_threatend_summary <- eez_count_threatened_all %>% 
  mutate(eez_trend = case_when(current_eez_count == RCP85_2100_eez_count ~ "constant",
                   current_eez_count > RCP85_2100_eez_count ~ "decreases",
                   current_eez_count < RCP85_2100_eez_count ~ "increases")) %>% 
  rename(AphiaID = aphiaid)

```

```{r}
summary_count <- eez_count_threatend_summary %>% 
  group_by(eez_trend) %>% 
  summarize(n = n()) %>%
  mutate(n_perc = n/length(aphiaid)) %>% 
  ungroup() 

summary_count
```

```{r}
# Merge in species status
eez_count_threatened_iucn <- left_join(eez_count_threatend_summary, meta_iucn_no_na_no_dupes)

```

```{r}
summary_count_iucn <- eez_count_threatened_iucn %>% 
  group_by(eez_trend, iucn_category) %>% 
  summarize(n = n()) %>%
  mutate(n_perc = n/length(aphiaid)) %>% 
  ungroup() 
```

```{r}
summary_count_group <- eez_count_threatened_iucn %>% 
  group_by(eez_trend, GROUP) %>% 
  summarize(n = n()) %>%
  mutate(n_perc = n/length(aphiaid)) %>% 
  ungroup() 
```

# Look at EEZs and their count of new/loss species

```{r}
eez_loss_gain <- eez_list_threatened %>% 
  group_by(GEONAME, status) %>% 
  summarize(n = n())
```

```{r}
eez_loss_gain_sf <- left_join(eez_loss_gain, eez)
```

```{r}
eez_loss_gain_only_sf <- eez_loss_gain_sf %>% 
  filter(status %in% c("entry into new", "exit from historical"))
```

```{r}
ggplot() +
  geom_sf(data = eez_loss_gain_only_sf, aes(geometry = geom, fill = n)) +
  scale_fill_viridis_c(limits = c(1, 65)) +
  theme_bw() +
  facet_wrap(~status) +
  labs(title = "Species count by EEZ (min = 1, max = 65)")
```

```{r}
ggplot() +
  geom_sf(data = eez_loss_gain_sf %>% 
            filter(status == "continued presence"),
          aes(geometry = geom, fill = n)) +
  scale_fill_viridis_c(limits = c(1, 400)) + 
  theme_bw() +
  labs(title = "Species count: continued presence")
```

```{r}
eez_c_f_sf <- eez_loss_gain_sf %>% 
  st_drop_geometry %>% 
  mutate(status = str_replace_all(status, " ", "_")) %>% 
  pivot_wider(names_from = status, values_from = n) %>% 
  replace_na(list(continued_presence = 0, exit_from_historical = 0,
                  entry_into_new = 0, no_presence = 0)) %>% 
  mutate(current_spp = continued_presence + exit_from_historical,
         future_spp = continued_presence + entry_into_new)
```

```{r}
eez_loss_gain_only <- eez_loss_gain_only_sf %>% 
  mutate(status = str_replace_all(status, " ", "_")) %>% 
  pivot_wider(names_from = status, values_from = n)
```

```{r}
ggplot() +
  geom_point(data = eez_c_f_sf, aes(x = current_spp, y = future_spp,
                                            color = Y_1), alpha = 0.75,
             position = position_jitterdodge()) +
  scale_color_viridis_c() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(color = "latitude",
       title = "Threatened current and future (RCP 8.5 2100) by EEZ",
       y = "future species count",
       x = "current species count") +
  theme_bw()
```

```{r}
# Can see that points above line are more yellow and purple (by poles)
ggplot() +
  geom_point(data = eez_c_f_sf, aes(x = exit_from_historical, y = entry_into_new,
                                            color = Y_1), alpha = 0.5,
             position = position_jitterdodge()) +
  scale_color_viridis_c() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(color = "latitude",
       title = "Threatened species loss and gain by EEZ",
       y = "species gain count",
       x = "species loss count") +
  theme_bw()
```


```{r}
# What about species change and SSI at EEZ level?
eez_c_f_ssi_sf <- eez_c_f_sf %>% 
  mutate(spp_change = future_spp - current_spp,
         ssi = (2 * continued_presence)/(current_spp + future_spp))
```

```{r}
ggplot() +
  geom_sf(data = eez_c_f_ssi_sf,
          aes(geometry = geom, fill = spp_change)) +
  #scale_fill_viridis_c() + 
  scale_color_distiller(palette = "BrBG", direction = 1) + 
  scale_fill_distiller(palette = "BrBG", direction = 1) +
  theme_bw() +
  labs(title = "Threatened species change")
```

```{r}
ggplot() +
  geom_sf(data = eez_c_f_ssi_sf,
          aes(geometry = geom, fill = ssi)) +
  scale_fill_viridis_c(direction = -1) + 
  theme_bw() +
  labs(title = "Threatened species SSI")
```

```{r}
# mostly the species are staying the same in each EEZ
summary(eez_c_f_ssi$ssi)
```

## Create above figures with vulnerable species seperate from endangered & critically endangered

```{r}
eez_list_threatened_iucn <- left_join(eez_list_threatened %>% 
                                        rename(AphiaID = aphiaid),
                                      meta_iucn_no_na_no_dupes)
```

```{r}
eez_loss_gain_vu <- eez_list_threatened_iucn %>% 
  filter(iucn_category == "Vulnerable") %>% 
  group_by(GEONAME, status) %>% 
  summarize(n = n())
```

```{r}
eez_loss_gain_en <- eez_list_threatened_iucn %>% 
  filter(iucn_category != "Vulnerable") %>% 
  group_by(GEONAME, status) %>% 
  summarize(n = n())
```

```{r}
eez_loss_gain_cr_en <- eez_list_threatened_iucn %>% 
  filter(iucn_category == "Critically Endangered") %>% 
  group_by(GEONAME, status) %>% 
  summarize(n = n())
```

```{r}
eez_loss_gain_en_sf <- left_join(eez_loss_gain_en, eez)
eez_loss_gain_vu_sf <- left_join(eez_loss_gain_vu, eez)
eez_loss_gain_cr_en_sf <- left_join(eez_loss_gain_cr_en, eez)
```

```{r}
eez_loss_gain_only_en_sf <- eez_loss_gain_en_sf %>% 
  filter(status %in% c("entry into new", "exit from historical"))

eez_loss_gain_only_vu_sf <- eez_loss_gain_vu_sf %>% 
  filter(status %in% c("entry into new", "exit from historical"))

eez_loss_gain_only_cr_en_sf <- eez_loss_gain_cr_en_sf %>% 
  filter(status %in% c("entry into new", "exit from historical"))
```

```{r}
ggplot() +
  geom_sf(data = eez_loss_gain_only_vu_sf, aes(geometry = geom, fill = n)) +
  scale_fill_viridis_c() +
  theme_bw() +
  facet_wrap(~status) +
  labs(title = "Vulnerable species count by EEZ")

ggplot() +
  geom_sf(data = eez_loss_gain_only_en_sf, aes(geometry = geom, fill = n)) +
  scale_fill_viridis_c() +
  theme_bw() +
  facet_wrap(~status) +
  labs(title = "Endangered+ species count by EEZ")

ggplot() +
  geom_sf(data = eez_loss_gain_only_cr_en_sf, aes(geometry = geom, fill = n)) +
  scale_fill_viridis_c() +
  theme_bw() +
  facet_wrap(~status) +
  labs(title = "Critcally Endangered species count by EEZ")
```
```{r}
eez_c_f_vu_sf <- eez_loss_gain_vu_sf %>% 
  st_drop_geometry %>% 
  mutate(status = str_replace_all(status, " ", "_")) %>% 
  pivot_wider(names_from = status, values_from = n) %>% 
  replace_na(list(continued_presence = 0, exit_from_historical = 0,
                  entry_into_new = 0, no_presence = 0)) %>% 
  mutate(current_spp = continued_presence + exit_from_historical,
         future_spp = continued_presence + entry_into_new)

eez_c_f_en_sf <- eez_loss_gain_en_sf %>% 
  st_drop_geometry %>% 
  mutate(status = str_replace_all(status, " ", "_")) %>% 
  pivot_wider(names_from = status, values_from = n) %>% 
  replace_na(list(continued_presence = 0, exit_from_historical = 0,
                  entry_into_new = 0, no_presence = 0)) %>% 
  mutate(current_spp = continued_presence + exit_from_historical,
         future_spp = continued_presence + entry_into_new)

eez_c_f_cr_en_sf <- eez_loss_gain_cr_en_sf %>% 
  st_drop_geometry %>% 
  mutate(status = str_replace_all(status, " ", "_")) %>% 
  pivot_wider(names_from = status, values_from = n) %>% 
  replace_na(list(continued_presence = 0, exit_from_historical = 0,
                  entry_into_new = 0, no_presence = 0)) %>% 
  mutate(current_spp = continued_presence + exit_from_historical,
         future_spp = continued_presence + entry_into_new)
```

```{r}
ggplot() +
  geom_point(data = eez_c_f_vu_sf, aes(x = current_spp, y = future_spp,
                                            color = Y_1), alpha = 0.75,
             position = position_jitterdodge()) +
  scale_color_viridis_c() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(color = "latitude",
       title = "Vulnerable current and future (RCP 8.5 2100) by EEZ",
       y = "future species count",
       x = "current species count") +
  theme_bw()
```
```{r}
ggplot() +
  geom_point(data = eez_c_f_en_sf, aes(x = current_spp, y = future_spp,
                                            color = Y_1), alpha = 0.75,
             position = position_jitterdodge()) +
  scale_color_viridis_c() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(color = "latitude",
       title = "Endangered+ current and future (RCP 8.5 2100) by EEZ",
       y = "future species count",
       x = "current species count") +
  theme_bw()

ggplot() +
  geom_point(data = eez_c_f_cr_en_sf, aes(x = current_spp, y = future_spp,
                                            color = Y_1), alpha = 0.75,
             position = position_jitterdodge()) +
  scale_color_viridis_c() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(color = "latitude",
       title = "Critically endangered current and future (RCP 8.5 2100) by EEZ",
       y = "future species count",
       x = "current species count") +
  theme_bw()
```

```{r}
# Can see that points above line are more yellow and purple (by poles)
ggplot() +
  geom_point(data = eez_c_f_vu_sf, aes(x = exit_from_historical, y = entry_into_new,
                                            color = Y_1), alpha = 0.5,
             position = position_jitterdodge()) +
  scale_color_viridis_c() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(color = "latitude",
       title = "Vulnerable species loss and gain by EEZ",
       y = "species gain count",
       x = "species loss count") +
  theme_bw()

# Can see that points above line are more yellow and purple (by poles)
ggplot() +
  geom_point(data = eez_c_f_en_sf, aes(x = exit_from_historical, y = entry_into_new,
                                            color = Y_1), alpha = 0.5,
             position = position_jitterdodge()) +
  scale_color_viridis_c() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(color = "latitude",
       title = "Endangered+ species loss and gain by EEZ",
       y = "species gain count",
       x = "species loss count") +
  theme_bw()

# Can see that points above line are more yellow and purple (by poles)
ggplot() +
  geom_point(data = eez_c_f_cr_en_sf, aes(x = exit_from_historical, y = entry_into_new,
                                            color = Y_1), alpha = 0.5,
             position = position_jitterdodge()) +
  scale_color_viridis_c() +
  scale_x_continuous(limits = c(-1,10)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(color = "latitude",
       title = "Critically endangered species loss and gain by EEZ",
       y = "species gain count",
       x = "species loss count") +
  theme_bw()
```

```{r}
# What about species change and SSI at EEZ level?
eez_c_f_vu_ssi_sf <- eez_c_f_vu_sf %>% 
  mutate(spp_change = future_spp - current_spp,
         ssi = (2 * continued_presence)/(current_spp + future_spp))

eez_c_f_en_ssi_sf <- eez_c_f_en_sf %>% 
  mutate(spp_change = future_spp - current_spp,
         ssi = (2 * continued_presence)/(current_spp + future_spp))

eez_c_f_cr_en_ssi_sf <- eez_c_f_cr_en_sf %>% 
  mutate(spp_change = future_spp - current_spp,
         ssi = (2 * continued_presence)/(current_spp + future_spp))
```

```{r}
ggplot() +
  geom_sf(data = eez_c_f_vu_ssi_sf,
          aes(geometry = geom, fill = spp_change)) +
  #scale_fill_viridis_c() + 
  scale_color_distiller(palette = "BrBG", direction = 1, limits = c(-40, 40)) + 
  scale_fill_distiller(palette = "BrBG", direction = 1, limits = c(-40, 40)) +
  theme_bw() +
  labs(title = "Vulnerable species change")

ggplot() +
  geom_sf(data = eez_c_f_en_ssi_sf,
          aes(geometry = geom, fill = spp_change)) +
  #scale_fill_viridis_c() + 
  scale_color_distiller(palette = "BrBG", direction = 1) + 
  scale_fill_distiller(palette = "BrBG", direction = 1) +
  theme_bw() +
  labs(title = "Endangered+ species change")

ggplot() +
  geom_sf(data = eez_c_f_cr_en_ssi_sf,
          aes(geometry = geom, fill = spp_change)) +
  #scale_fill_viridis_c() + 
  scale_color_distiller(palette = "BrBG", direction = 1, limits = c(-40, 40)) + 
  scale_fill_distiller(palette = "BrBG", direction = 1, limits = c(-40, 40)) +
  theme_bw() +
  labs(title = "Critically endangered species change")
```

## Which species are projected down to 0 EEZs? Do they have any range left?
All nine species don't have any pixels of suitable habitat in RCP8.5 2100

```{r}
eez_future_0 <- eez_count_threatened_all %>% 
  filter(RCP85_2100_eez_count == 0)

eez_future_0_aphiaid <- eez_future_0$aphiaid

```

```{r}
# No pixels of suitable habitat for Rhincodon typus (whale shark) under rcp8.5 2100 
# 190 EEZs to 0
aphiaid_105847 <- read_parquet("/home/shares/usgs-cap-rangeshifts/data_aquax/iucn/SDM/FINAL_EMSDM_EMMEAN_SP_105847.parquet")
```

```{r}
# No pixels of suitable habitat for Thunnus obesus (bigeye tuna) under rcp8.5 2100 
# 180 EEZs to 0
aphiaid_127028 <- read_parquet("/home/shares/usgs-cap-rangeshifts/data_aquax/iucn/SDM/FINAL_EMSDM_EMMEAN_SP_127028.parquet")
```

```{r}
# No pixels of suitable habitat for Scymnodon ichiharai (Japanese velet dogfish) under rcp8.5 2100 
# 13 EEZs to 0
aphiaid_299099 <- read_parquet("/home/shares/usgs-cap-rangeshifts/data_aquax/iucn/SDM/FINAL_EMSDM_EMMEAN_SP_299099.parquet")
```

```{r}
# No pixels of suitable habitat for Neoopisthopterus cubanus (Cuban longfin herring) under rcp8.5 2100 
# 6 EEZs to 0
aphiaid_281695 <- read_parquet("/home/shares/usgs-cap-rangeshifts/data_aquax/iucn/SDM/FINAL_EMSDM_EMMEAN_SP_281695.parquet")
```

```{r}
# No pixels of suitable habitat for Rhynchobatus cooki (roughnose wedgefish) under rcp8.5 2100 
# 6 EEZs to 0
aphiaid_1022441 <- read_parquet("/home/shares/usgs-cap-rangeshifts/data_aquax/iucn/SDM/FINAL_EMSDM_EMMEAN_SP_1022441.parquet")
```

```{r}
# No pixels of suitable habitat for Platyrhina hyugaensis (Hyuga fanray) under rcp8.5 2100 
# 4 EEZs to 0
aphiaid_712386 <- read_parquet("/home/shares/usgs-cap-rangeshifts/data_aquax/iucn/SDM/FINAL_EMSDM_EMMEAN_SP_712386.parquet")
```

```{r}
# No pixels of suitable habitat for Holohalaelurus favus (honeycomb Izak) under rcp8.5 2100 
# 3 EEZs to 0
aphiaid_390077 <- read_parquet("/home/shares/usgs-cap-rangeshifts/data_aquax/iucn/SDM/FINAL_EMSDM_EMMEAN_SP_390077.parquet")
```

```{r}
# No pixels of suitable habitat for Pseudocaranx chilensis (Juan Fernandez trevally) under rcp8.5 2100 
# 2 EEZs to 0
aphiaid_273281 <- read_parquet("/home/shares/usgs-cap-rangeshifts/data_aquax/iucn/SDM/FINAL_EMSDM_EMMEAN_SP_273281.parquet")
```

```{r}
# No pixels of suitable habitat for Anchoa choerostoma (Bermuda anchovy) under rcp8.5 2100 
# 1 EEZs to 0
aphiaid_275510 <- read_parquet("/home/shares/usgs-cap-rangeshifts/data_aquax/iucn/SDM/FINAL_EMSDM_EMMEAN_SP_275510.parquet")
```