---
title: "Transfer AquaX files from Dropbox to Aurora"
author: "Casey"
format: html
---

## Test using `rdrop2` package

```{r}
library(tidyverse)
# remotes::install_github('karthik/rdrop2')
library(rdrop2)

```

## Authentication

See https://github.com/karthik/rdrop2 for vignette
```{r}
### NOTE: drop_auth() seems to fail on Aurora, perhaps firewall?
### Ran this on local and saved token as token.rds and then uploaded to Aurora
### ... and then added to .gitignore

# token <- drop_auth()
# saveRDS(token, file = "token.rds")
token <- readRDS(file = here::here('token.rds'))
# Then in any drop_* function, pass `dtoken = token`
# Tokens are valid until revoked.
```

NOTE: difficult to use this on Aurora - use on local and then copy to Aurora (and put token.rds in .gitignore):

```
scp token.rds ohara@aurora.nceas.ucsb.edu:~/github/aquax_sample/token.rds
```

## Access Dropbox and files

```{r}
acct_info <- drop_acc(dtoken = token) %>% 
  data.frame()

dropbox_dirs <- drop_dir(path = '', dtoken = token) %>% 
    filter(.tag == "folder")

```


## Copy to Aurora

Verify "download" vs. "copy": looks like "download" should grab file and make local copy, while "copy" creates a duplicate ON DROPBOX.


```{r}
dropbox_paths <- paste0('CASEY_IUCN_', 1:3)

for(p in dropbox_paths) {
  
  ### identify all files in `p` path
  aquax_files <- drop_dir(path = p, dtoken = token)
  
  out_dir <- '/home/shares/usgs-cap-rangeshifts/data_aquax/iucn/'
  if(!dir.exists(out_dir)) dir.create(out_dir)
  
  ### loop over each file and copy to directory
  for(f in aquax_files$path_display) {
    ### f <- aquax_files$path_display[1]
    out_f <- file.path(out_dir, basename(f))
    if(!file.exists(out_f)) {
      drop_download(path = f, local_path = out_f,
                    dtoken = token)
    }
  }
}
```

At this stage, unzip files to `/home/shares/usgs-cap-rangeshifts/data_aquax/iucn/SDM`:

* Navigate to `/home/shares/usgs-cap-rangeshifts/data_aquax/iucn/` and unzip each .zip in turn.  The zips should each download the various .Rdata files into a folder called SDM (same for each of the three tranches).
