---
title: "R Dropbox access test"
author: "Casey"
format: html
editor: visual
---

## Test using `rdrop2` package

```{r}
library(tidyverse)
# remotes::install_github('karthik/rdrop2')
library(rdrop2)

### Sample data: .Rdata of spp traits?
sample_rdata_url <- 'https://www.dropbox.com/scl/fi/ol7dxw58mzyth9xovqjmn/spp_traits.RData?rlkey=5j504e0kwyd811an4752f38ay&st=lwy506pm&dl=0'
```

## Authentication

See https://github.com/karthik/rdrop2 for vignette
```{r}
### NOTE: drop_auth() seems to fail on Aurora, perhaps firewall?
### Ran this on local and saved token as token.rds and then uploaded to Aurora
### ... and then added to .gitignore

# token <- drop_auth()
# saveRDS(token, file = "token.rds")
token <- readRDS(file = here::here('token.rds'))
# Then in any drop_* function, pass `dtoken = token`
# Tokens are valid until revoked.
```


## Access Dropbox and files

```{r}
acct_info <- drop_acc(dtoken = token) %>% 
  data.frame()

dropbox_dirs <- drop_dir(path = '', dtoken = token) %>% 
    filter(.tag == "folder")

dropbox_top_fs <- drop_dir(dtoken = token) %>%
  filter(.tag == "file")
```


## Copy to Aurora

Verify "download" vs. "copy": looks like "download" should grab file and make local copy, while "copy" creates a duplicate ON DROPBOX.


```{r}
drop_download(path = "spp_traits.RData", local_path = here::here("spp_traits_copy.RData"),
              dtoken = token)
```

## Download from Dropbox Transfer (instead of account)

Try `download.file()` instead - modify the shared link, ensuring the URL ends with ?dl=1 or &dl=1 to force a direct download rather than a web preview.

```{r}
#| eval: false
#| 
drop1_url <- 'https://www.dropbox.com/transfer/AAAAAHm0Y1hqDlePZvioX23KzXvThKoCSTBVqOb5Rtv5lcE4nRgcLAA'

dest1 <- '/home/shares/usgs-cap-rangeshifts/data_aquax/aquax_iucn_drop1.zip'

download.file(url = drop1_url,
              # url = paste0(drop1_url, '?dl=1'),
              destfile = dest1, mode = 'wb')
```



