---
title: "Aquax - duckdb functions"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
library(arrow)
library(collapse)
```

This document improves upon 01_iucn_aquax_eez_overlap.qmd by correcting for errors in the merge of EEZ and AquaX data that was leaving stripes out of the dataset due to mismatch in X and Y. This document also uses functions that utilize duckdb's ability to filter the SDM data before it is read into R for efficiency improvements. 

This document uses the IUCN lookup and threat category data created in the chi repository. It connects IUCN status to AquaX data using the Aphia ID, which is a unique identifier for marine species in the World Register of Marine Species (WoRMS). 

```{r}
iucn_lookup <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/_data_raw/iucn_redlist_2026/iucn_worms_aphia_id_complete.csv")
```

```{r}
iucn_new_cat <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/_data_raw/iucn_redlist_2026/iucn_species_assessments.csv")
```

```{r}
# Load functions for processing AquaX data
source(here( "common_fxns.R"))
```

We'll use the functions to process the AquaX data for the count of threatened species entering and exiting EEZs.

## EEZ 

```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
eez_simple <- eez |>
  select(MRGID, GEONAME, geom)
```

```{r}
# Create a blank raster matching AquaX resolution (EEZs should be large enough to just do 0.05 degree raster)
eez_raster_blank <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05,
                  crs = "EPSG:4326")
```

```{r}
eez_raster_geoname <- rasterize(eez_simple, eez_raster_blank, field = "GEONAME")
```

```{r}
eez_raster_mrgid <- rasterize(eez_simple, eez_raster_blank, field = "MRGID")
```

```{r}
# Get GEONAMES of non-contested regions (will run all regions to include the most data)
eez_200nm <-eez %>% 
  filter(POL_TYPE == "200NM")

eez_200nm_list <- eez_200nm$GEONAME
```


# IUCN threatened species

```{r}
# Filter the iucn_lookup for species with an SDM
iucn_lookup_sdm <- iucn_lookup %>% filter(aquax == "SDM")
```

```{r}
# Filter iucn categories for threatened species (CR, EN, VU) and near threatened (NT) species
iucn_threatened <- iucn_new_cat %>% 
  filter(redlist_category %in% c("CR","EN", "VU", "NT"))
```

```{r}
# Merge in the iucn_lookup_sdm to get the AphiaID for the threatened species
aphiaid_threatened <- inner_join(iucn_lookup_sdm, iucn_threatened, by = c("iucn_assessment_id" = "assessment_id")) 
```

```{r}
# Get aphiaids for the threatened species
aphiaid_threat <- unique(aphiaid_threatened$aphia_id)
```

## Prep AquaX data
We'll be working with the 1,323 at-risk aphiaids

```{r}
# read in meta data on the aquax species
load("/home/shares/data-aquax/META_02122025.Rdata")
```

```{r}
meta_clean <- META %>% 
  mutate(worms_name = tolower(ScientificName)) %>% 
  select(!ScientificName) %>% 
  # simplify for now for easier viewing
  select(AphiaID, SPID, worms_name, GROUP, NR_CR, range_pixels, SDM)
```

```{r}
spp_fs <- list.files('/home/shares/data-aquax/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(META, by = c('aphiaid' = 'AphiaID'))
```

```{r}
# Filter for the AphiaID of the threatend species - get 1,047 threatened species and 276 non-threatened species with SDMs
df_threatened <- df %>% 
  filter(aphiaid %in% aphiaid_threat)
```

```{r}
test_f <- read_parquet("/home/shares/data-aquax/SDM/FINAL_EMSDM_EMMEAN_SP_1017389.parquet")
```

```{r}
set.seed(123)
test_spp_30 <- slice_sample(df_threatened, n = 30)
files_30 <- test_spp_30$f
aphiaid_30 <- test_spp_30$aphiaid
```

```{r}
# Have to use the EEZ raster with the numeric IDs (MRGID) for zonal stats
# Try // - took 2 min
system.time({
  eez_list_zonal_db <- parallel::mclapply(aphiaid_30, FUN = extract_eez_zonal_db, 
                                       df = test_spp_30, 
                                       s = "Current",
                                       eez_raster = eez_raster_mrgid, 
                                       mc.cores = 10)
})
```

```{r}
eez_list_all_zonal_db <- bind_rows(eez_list_zonal_db) 
```

## Test out full threatened sample

```{r}
files <- df_threatened$f
aphiaid <- df_threatened$aphiaid
```

```{r}
# Try // - took 1 min for 1323 species and 1 scenario
system.time({
  eez_list_threatened_current <- parallel::mclapply(aphiaid, FUN = extract_eez_zonal_db, 
                                            df = df_threatened,
                                            s = "Current",
                                            eez_raster = eez_raster_mrgid,
                                            mc.cores = 40)
})
```

```{r}
eez_list_threatened_current_bind <- bind_rows(eez_list_threatened_current) 
```

```{r}
# Get species pixel counts by EEZ using duckdb to filter first
extract_eez_zonal_db_test <- function(id, df, eez_raster, s) {
  spp_filter <- df %>% filter(aphiaid == id)
  
  if (nrow(spp_filter) == 0) {
    warning("No rows found for aphiaid: ", id)
    return(NULL)
  }
  
  # Create database connection
  con <- DBI::dbConnect(duckdb::duckdb())
  
  query_str <- paste0("SELECT ROUND(x, 3) AS x, ROUND(y, 3) AS y, ", paste(s, collapse = ", "), 
                      " FROM read_parquet('", spp_filter$f, "')",
                      " WHERE ", paste(s, collapse = ", "), " >= cutoff")
  
  # Run query and filter for scenario pixels above cutoff threshold
  
  spp_df <- DBI::dbGetQuery(con, query_str) 
  
  # Close connection
  duckdb::dbDisconnect(con, shutdown = TRUE)
  
  # Set option for if there are no pixels above the cutoff threshold
  if (nrow(spp_df) == 0) {
    warning("No pixels above cutoff threshold for aphiaid: ", id)
    # Still need to shut connection in this case
    duckdb::dbDisconnect(con, shutdown = TRUE)
    # Return NA to track loss species
    return(data.frame(MRGID = NA, n_pixel = NA, aphiaid = id, scenario = s))
  } else {
  # Create raster
  r_spp <- rast(spp_df, crs = "EPSG:4326") %>% 
    extend(ext(eez_raster))
  
  # zonal() counts the notNA pixels within each EEZ zone (don't need binary)
  result <- zonal(r_spp, eez_raster, fun = "notNA", na.rm = TRUE)
  
  # eez_raster zones are numeric IDs — join back to EEZ names
  result %>%
    #left_join(eez_zone_lookup, by = "GEONAME") %>%
    # add aphiaid to track each species
    mutate(aphiaid = id,
           scenario = s) %>% 
    rename(n_pixel = .data[[s]]) %>% 
    # Drop rows where there are 0  notNA pixels
    filter(n_pixel != 0)}
}
```



```{r}
# Try // - took 1 min for 1323 species and 1 scenario
system.time({
  eez_list_threatened_rcp85_2050 <- parallel::mclapply(aphiaid, FUN = extract_eez_zonal_db_test, 
                                            df = df_threatened,
                                            s = "RCP85_2050",
                                            eez_raster = eez_raster_mrgid,
                                            mc.cores = 40)
})
```
```{r}
# 1523828 created error for RCP85_2050 (250 in list)
test_f <- read_parquet("/home/shares/data-aquax/SDM/FINAL_EMSDM_EMMEAN_SP_1523828.parquet")
```

```{r}
eez_list_threatened_rcp85_2050_bind <- bind_rows(eez_list_threatened_rcp85_2050) 
```

```{r}
eez_rcp58_2050_na <- eez_list_threatened_rcp85_2050_bind %>% 
  filter(is.na(n_pixel))
```

