---
title: "AquaX and IUCN status"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
library(arrow)
```

This document uses the IUCN lookup and threat category data created in the chi repository. It connects IUCN status to AquaX data using the Aphia ID, which is a unique identifier for marine species in the World Register of Marine Species (WoRMS). 

```{r}
iucn_lookup <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/data/iucn_redlist/iucn_worms_aphia_id_complete.csv")
```

```{r}
iucn_new_cat <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/data/iucn_redlist/iucn_species_assessments.csv")
```

```{r}
# Load functions for processing AquaX data
source(here( "common_fxns.R"))
```

We'll use the functions to process the AquaX data for the count of threatened species entering and exiting EEZs.

## EEZ 

```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
eez_simple <- eez |>
  select(MRGID, GEONAME, geom)
```

```{r}
ggplot() +
  geom_sf(data = eez, aes()) +
  geom_sf(data = eez |> filter(GEONAME == "Russian Exclusive Economic Zone"), aes(), fill = "red") +
  theme_void() +
  theme(legend.position = "none")
```

```{r}
# Create a blank raster matching AquaX resolution (EEZs should be large enough to just do 0.05 degree raster)
eez_raster_blank <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05,
                  crs = "EPSG:4326")
```

```{r}
eez_raster_geoname <- rasterize(eez_simple, eez_raster_blank, field = "GEONAME")
```

```{r}
eez_raster_mrgid <- rasterize(eez_simple, eez_raster_blank, field = "MRGID")
```

# IUCN threatened species

```{r}
# Filter the iucn_lookup for species with an SDM
iucn_lookup_sdm <- iucn_lookup %>% filter(aquax == "SDM")
```

```{r}
# Filter iucn categories for threatened species (CR, EN, VU)
iucn_threatened <- iucn_new_cat %>% 
  filter(redlist_category %in% c("CR","EN", "VU"))
```

```{r}
# Merge in the iucn_lookup_sdm to get the AphiaID for the threatened species
aphiaid_threatened <- inner_join(iucn_lookup_sdm, iucn_threatened, by = c("iucn_assessment_id" = "assessment_id")) 
```

```{r}
# Get aphiaids for the threatened species
aphiaid_threat <- unique(aphiaid_threatened$aphia_id)
```

## Prep AquaX data
We'll be working with the 1,047 thretened aphiaids

```{r}
# read in meta data on the aquax sample of redlist species
load("/home/shares/data-aquax/iucn_request/meta.Rdata")
```

```{r}
meta_clean <- meta %>% 
  mutate(worms_name = tolower(ScientificName)) %>% 
  select(!ScientificName) %>% 
  # simplify for now for easier viewing
  select(AphiaID, SPID, worms_name, GROUP, NR_CR, range_pixels, SDM)
```

```{r}
spp_fs <- list.files('/home/shares/data-aquax/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(meta, by = c('aphiaid' = 'AphiaID'))
```

```{r}
# Filter for the AphiaID of the threatend species - get 987 species in AquaX data (even though the 1,047 species were in the iucn_lookup_sdm with AquaX SDMs)
df_threatened <- df %>% 
  filter(aphiaid %in% aphiaid_threat)
```

```{r}
test_f <- read_parquet("/home/shares/data-aquax/SDM/FINAL_EMSDM_EMMEAN_SP_1017389.parquet")
```

```{r}
set.seed(123)
test_spp_30 <- slice_sample(df_threatened, n = 30)
files_30 <- test_spp_30$f
aphiaid_30 <- test_spp_30$aphiaid
```

```{r}
test_r <- rast(test_f, crs = "EPSG:4326") %>%
    extend(ext(eez_raster_mrgid))
```

```{r}
# Test plot
plot(test_r_eq$RCP45_2050)
```

```{r}
r_binary45_2050 <- ifel(test_r[["RCP45_2050"]] >= mean(test_f$cutoff, na.rm = TRUE), 1, NA)
```

```{r}
# Have to use the EEZ raster with the numeric IDs (MRGID) for zonal stats
# Try // - took 2 min
system.time({
  eez_list_zonal <- parallel::mclapply(aphiaid_30, FUN = extract_eez_zonal, 
                                       df = test_spp_30, 
                                       eez_raster = eez_raster_mrgid, 
                                       mc.cores = 10)
})
```

```{r}
eez_list_all_zonal <- bind_rows(eez_list_zonal) 
```

```{r}
eez_ids <- eez %>% 
  st_drop_geometry() %>% 
  dplyr::select(MRGID, GEONAME)
```

```{r}
eez_list_geoname <- left_join(eez_list_all_zonal, eez_ids)
```


```{r}
eez_raster_df <- as.data.frame(eez_raster_mrgid_eq, xy = TRUE, cells = TRUE)
eez_raster_df_non_eq <- as.data.frame(eez_raster_mrgid, xy = TRUE, cells = TRUE)
```

## Test out full threatened sample

```{r}
files <- df_threatened$f
aphiaid <- df_threatened$aphiaid
```

```{r}
# Try // - took 21 min for 987 species and 4 scenarios
system.time({
  eez_list_threatened <- parallel::mclapply(aphiaid, FUN = extract_eez_zonal, 
                                            df = df_threatened,
                                            eez_raster = eez_raster_mrgid,
                                            mc.cores = 20)
})
```

```{r}
eez_list_threatened_bind <- bind_rows(eez_list_threatened) 
```

```{r}
# Merge in EEZ ids
eez_threatened_ids <- left_join(eez_list_threatened_bind, eez_ids)
```

```{r}
# Save data
write_csv(eez_threatened_ids, here("data/iucn_threatened_spp_eez_pxls_052126.csv"))
```

## Test out method for identifying the pixels entering/leaving the EEZs


```{r}
# Organize dataframe so that's it's easier to select a scenario and compare to current
eez_spp_pxls_long <- eez_list_geoname %>% 
  pivot_longer(cols = starts_with("n_RCP"),
               names_to = "scenario",
               values_to = "pxl_count") %>% 
  mutate(scenario = str_remove(scenario, "n_")) %>% 
  mutate(spp_change = case_when(is.na(n_current) &  pxl_count > 0 ~ 'entry into new',
                            n_current > 0  & is.na(pxl_count) ~ 'exit from historical',
                            is.na(n_current)  & is.na(pxl_count) == 0 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

```{r}
# Get counts of entry into new, exit from historical etc. by EEZ
eez_spp_entry <- eez_spp_pxls_long %>% 
  filter(spp_change == "entry into new")
```


```{r}
# Test a sample species that is moving into new EEZs
spp_test <- read_parquet("/home/shares/data-aquax/SDM/FINAL_EMSDM_EMMEAN_SP_126413.parquet")
```
```{r}
spp_test_r <- rast(spp_test, crs = "EPSG:4326") %>%
    extend(ext(eez_raster_mrgid))
```

```{r}
spp_test_r_eq <- project(spp_test_r$RCP45_2050, eq_earth, method = "near")
```
```{r}
# Get the MRGIDs for the future presence of species 
spp_test_eez <- eez_spp_entry |> 
  filter(aphiaid == 126413 & scenario == "RCP45_2050") |> 
  pull(unique(MRGID))
```

```{r}
# Create a mask of the EEZs that the species is predicted to be present in the future
test_eez_mask <- ifel(eez_raster_mrgid_eq %in% spp_test_eez, 1, NA)
```


```{r}
masked_spp <- terra::mask(spp_test_r_eq, test_eez_mask)
```

```{r}
masked_spp_test_df <- as.data.frame(masked_spp, xy = TRUE, cells = TRUE)
```

# Running into issues here
```{r}
# Get entry pixels for a species by EEZ
extract_eez_entry_pixels_df <- function(id, df, s, eez_raster, eez_raster_eq, eez_spp_entry, crs) {
  spp_filter <- df %>% filter(aphiaid == id)
  
  if (nrow(spp_filter) == 0) {
    warning("No rows found for aphiaid: ", id)
    return(NULL)
  }
  
  spp <- read_parquet(spp_filter$f) %>%
    mutate(aphiaid = id) %>% 
    dplyr::select(x, y, Current, all_of(s), cutoff)

  r_spp <- rast(spp, crs = "EPSG:4326") %>%
    extend(ext(eez_raster))

  r_spp_eq <- project(r_spp, crs, method = "near")
  
  cutoff <- mean(spp$cutoff, na.rm = TRUE)
  
  # Create binary layers: 1 if >= cutoff, NA otherwise
  r_binary <- ifel(r_spp_eq[[s]] >= cutoff, 1, NA)
  names(r_binary) <- s
  
  # Filter the df of entry into for this species & scenario
  entry_spp_df <- eez_spp_entry %>% 
    filter(aphiaid == id) %>% 
    filter(scenario == s)
  
  # Get the EEZs for which the species & scenario is entry
  spp_entry_eezs <- unique(entry_spp_df$MRGID)

  # Create a mask of the EEZs that the species is predicted to be present in the future
  eez_mask <- ifel(eez_raster_eq %in% spp_test_eez, 1, NA)

  # Apply mask
  masked_spp <- terra::mask(spp_test_r_eq, eez_mask)
  
  # Return stack of rasters
  raster_list <- list()
  raster_list[[aphiaid]] <- masked_spp
  return(raster_list)
}
```


```{r}
system.time({
  eez_list_pixels <- parallel::mclapply(aphiaid_30, FUN = extract_eez_entry_pixels_df, 
                                       s = "RCP45_2050",
                                       df = test_spp_30, 
                                       eez_raster = eez_raster_mrgid,
                                       eez_raster_eq = eez_raster_mrgid_eq,
                                       eez_spp_entry = eez_spp_entry,
                                       crs = eq_earth, 
                                       mc.cores = 10)
})
```