---
title: "AquaX and IUCN status"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
```

This document uses the IUCN lookup and threat category data created in the chi repository. It connects IUCN status to AquaX data using the Aphia ID, which is a unique identifier for marine species in the World Register of Marine Species (WoRMS). 

```{r}
iucn_lookup <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/data/iucn_redlist/iucn_worms_aphia_id_complete.csv")
```

```{r}
iucn_new_cat <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/data/iucn_redlist/iucn_species_assessments.csv")
```

```{r}
# Load functions for processing AquaX data
source(here( "common_fxns.R"))
```

We'll use the functions to process the AquaX data for the count of threatened species entering and exiting EEZs. We'll use an equal area projection

```{r}
# Save crs for equal area projection
eq_earth <- "+proj=eqearth"
```

## EEZ 

```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
eez_eq <- st_transform(eez, crs = st_crs(eq_earth))
```

```{r}
ggplot() +
  geom_sf(data = eez, aes()) +
  geom_sf(data = eez |> filter(GEONAME == "Russian Exclusive Economic Zone"), aes(), fill = "red") +
  theme_void() +
  theme(legend.position = "none")
```

```{r}
# Create a blank raster matching AquaX resolution (EEZs should be large enough to just do 0.05 degree raster)
eez_raster_blank <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05)
```

```{r}
eez_raster_geoname <- rasterize(eez, eez_raster_blank, field = "GEONAME")
```

```{r}
eez_raster_mrgid <- rasterize(eez, eez_raster_blank, field = "MRGID")
```


```{r}
# Transform to equal area projection
eez_raster_mrgid_eq <- project(eez_raster_mrgid, eq_earth)
```