---
title: "Aquax - duckdb functions"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
library(arrow)
library(collapse)
```

This document improves upon 01_iucn_aquax_eez_overlap.qmd by correcting for errors in the merge of EEZ and AquaX data that was leaving stripes out of the dataset due to mismatch in X and Y. This document also uses functions that utilize duckdb's ability to filter the SDM data before it is read into R for efficiency improvements. 

This document uses the IUCN lookup and threat category data created in the chi repository. It connects IUCN status to AquaX data using the Aphia ID, which is a unique identifier for marine species in the World Register of Marine Species (WoRMS). 

```{r}
iucn_lookup <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/ab9cd09931a4ce5c12d0bcd2984595d6522ed158/_data/iucn_redlist_2026/iucn_worms_aphia_id_complete.csv")
```

```{r}
iucn_new_cat <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/_data/iucn_redlist_2026/iucn_species_assessments.csv")
```

```{r}
# Load functions for processing AquaX data
source(here( "common_fxns.R"))
```

We'll use the functions to process the AquaX data for the count of threatened species entering and exiting EEZs.

## Species count in EEZs
Rounding issue doesn't impact the count of species in each eez so use iucn_threatened_spp_eez_pxls_061626.csv. This file was created using the 01_iucn_aquax_eez_overlap.qmd script and contains the count of pixels for each species in each EEZ for current and future scenarios.

```{r}
# Spp in EEZs
spp_eez_pxl_count <- read_csv(here("data/iucn_threatened_spp_eez_pxls_061626.csv"))
```

```{r}
# Organize dataframe so that's it's easier to select a scenario and compare to current
eez_spp_pxls_long <- spp_eez_pxl_count %>% 
  pivot_longer(cols = starts_with("n_RCP"),
               names_to = "scenario",
               values_to = "pxl_count") %>% 
  mutate(scenario = str_remove(scenario, "n_")) %>% 
  # Replace NA with 0, otherwise will be assigned to "continued presence"
  mutate(pxl_count = replace_na(pxl_count, 0),
         n_current = replace_na(n_current, 0)) %>%
  # Set 10 pixel threshold (at least ten pixels need in EEZ for species presence)
  mutate(spp_change = case_when(n_current < 10 &  pxl_count >= 10 ~ 'entry into new',
                            n_current >= 10  & pxl_count < 10 ~ 'exit from historical',
                            n_current < 10  & pxl_count < 10 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

## EEZ 

```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
eez_simple <- eez |>
  select(MRGID, GEONAME, geom)
```

```{r}
eez_ids <- eez %>% 
  st_drop_geometry() %>% 
  dplyr::select(MRGID, GEONAME)
```

```{r}
# Create a blank raster matching AquaX resolution (EEZs should be large enough to just do 0.05 degree raster)
eez_raster_blank <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05,
                  crs = "EPSG:4326")
```

```{r}
eez_raster_geoname <- rasterize(eez_simple, eez_raster_blank, field = "GEONAME")
```

```{r}
eez_raster_mrgid <- rasterize(eez_simple, eez_raster_blank, field = "MRGID")
```

```{r}
# Get GEONAMES of non-contested regions (will run all regions to include the most data)
eez_200nm <-eez %>% 
  filter(POL_TYPE == "200NM")

eez_200nm_list <- eez_200nm$MRGID
```


# IUCN threatened species

```{r}
# Filter the iucn_lookup for species with an SDM
iucn_lookup_sdm <- iucn_lookup %>% filter(aquax == "SDM")
```

```{r}
# Filter iucn categories for threatened species (CR, EN, VU) and near threatened (NT) species
iucn_threatened <- iucn_new_cat %>% 
  filter(redlist_category %in% c("CR","EN", "VU", "NT"))
```

```{r}
# Merge in the iucn_lookup_sdm to get the AphiaID for the threatened species
aphiaid_threatened <- inner_join(iucn_lookup_sdm, iucn_threatened, by = c("iucn_assessment_id" = "assessment_id")) 
```

```{r}
# Get aphiaids for the threatened species
aphiaid_threat <- unique(aphiaid_threatened$aphia_id)
```

## Prep AquaX data
We'll be working with the 1,323 at-risk aphiaids

```{r}
# read in meta data on the aquax species
load("/home/shares/data-aquax/META_02122025.Rdata")
```

```{r}
meta_clean <- META %>% 
  mutate(worms_name = tolower(ScientificName)) %>% 
  select(!ScientificName) %>% 
  # simplify for now for easier viewing
  select(AphiaID, SPID, worms_name, GROUP, NR_CR, range_pixels, SDM)
```

```{r}
spp_fs <- list.files('/home/shares/data-aquax/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(META, by = c('aphiaid' = 'AphiaID'))
```

```{r}
# Filter for the AphiaID of the threatend species - get 1,047 threatened species and 276 non-threatened species with SDMs
df_threatened <- df %>% 
  filter(aphiaid %in% aphiaid_threat)
```

```{r}
test_f <- read_parquet("/home/shares/data-aquax/SDM/FINAL_EMSDM_EMMEAN_SP_1017389.parquet")
```

## Get the pixels of entry and exit 
Update function to fix XY rounding error that created stripes in maps and also update function to only read in necessary data in duckdb for efficiency.

```{r}
# Get counts of entry into new, exit from historical etc. by EEZ
eez_spp_entry <- eez_spp_pxls_long %>% 
  filter(spp_change == "entry into new")
```

```{r}
# Create dataframe of EEZ raster
eez_raster_df <- as.data.frame(eez_raster_mrgid, xy = TRUE) %>% 
  # Round so that the X and Y match the AquaX data (which is rounded to 3 decimal places)
  mutate(x = round(x, 3),
         y = round(y, 3))
```

# Test sample

```{r}
set.seed(123)
test_spp_30 <- slice_sample(df_threatened, n = 100)
files_30 <- test_spp_30$f
aphiaid_30 <- test_spp_30$aphiaid
```

```{r}
system.time({
  eez_list_count_df <- parallel::mclapply(aphiaid_30, FUN = extract_eez_entry_pixels_df_db, 
                                       df = test_spp_30, 
                                       s = "RCP85_2050",
                                       eez_spp_entry = eez_spp_entry,
                                       eez_raster_df = eez_raster_df,
                                       mc.cores = 10)
})
```

```{r}
eez_count_df <- bind_rows(eez_list_count_df)
```

```{r}
# Sum by XY cell, using collapse
system.time({
eez_sum_df <- eez_count_df %>% 
  fgroup_by(MRGID, x, y) %>% 
  fcount()
})
```

```{r}
# Plot to see if stripes are gone
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8328)) +
  coord_sf(expand = FALSE) +
  geom_tile(data = eez_sum_df %>% filter(MRGID == 8328), aes(x = x, y = y, fill = N, color = N)) +
  #geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5)+
  theme_minimal()
```

```{r}
# Plot to see if stripes are gone
ggplot() +
  geom_sf(data = eez) +
  coord_sf(expand = FALSE) +
  geom_tile(data = eez_sum_df, aes(x = x, y = y, fill = N, color = N)) +
  #geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5)+
  theme_minimal()
```

# Run full set - species entries

```{r}
# Took 34 seconds over 30 cores
system.time({
  entry_rcp85_list <- parallel::mclapply(aphiaid_threat, FUN = extract_eez_entry_pixels_df_db, 
                                       df = df_threatened, 
                                       s = "RCP85_2050",
                                       eez_spp_entry = eez_spp_entry,
                                       eez_raster_df = eez_raster_df,
                                       mc.cores = 30)
})
```

```{r}
entry_rcp85_df <- bind_rows(entry_rcp85_list) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse
system.time({
entry_rcp85_sum_df <- entry_rcp85_df %>% 
  fgroup_by(MRGID, x, y) %>% 
  fcount() %>% 
  mutate(scenario = "RCP85_2050")
})
```

```{r}
# Plot to see if stripes are gone
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8328)) +
  coord_sf(expand = FALSE) +
  geom_tile(data = entry_rcp85_sum_df  %>% filter(MRGID == 8328), aes(x = x, y = y, fill = N, color = N)) +
  #geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5)+
  scale_fill_viridis_c(direction = -1) +
  scale_color_viridis_c(direction = -1) + 
  theme_minimal()
```

```{r}
# Plot to see if stripes are gone (all EEZ regions; including joint regime and contested regions)
ggplot() +
  geom_sf(data = eez) +
  coord_sf(expand = FALSE) +
  geom_tile(data = entry_rcp85_sum_df, aes(x = x, y = y, fill = N, color = N)) +
  scale_fill_viridis_c(limits = c(1,32), 
                      breaks = c(1, 2, 4, 8, 16, 32),
                      direction = -1,
                      trans = "log", # visually stretches the lower value data
                      guide = guide_colorbar(
                      barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                      barheight = unit(0.4, "cm"),
                      ticks = TRUE,
                      ticks.colour = "white",
                      ticks.linewidth = 0.8
  )) +
  scale_color_viridis_c(limits = c(1,32), 
                        breaks = c(1, 2, 4, 8, 16, 32),
                        direction = -1,
                        trans = "log", # visually stretches the lower value data
                        guide = "none")  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "species entry count",
       color = "species entry count",
       x = "",
       y = "") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt")) 
```

```{r}
# Took 34 seconds over 30 cores
system.time({
  entry_rcp45_2050_list <- parallel::mclapply(aphiaid_threat, FUN = extract_eez_entry_pixels_df_db, 
                                       df = df_threatened, 
                                       s = "RCP45_2050",
                                       eez_spp_entry = eez_spp_entry,
                                       eez_raster_df = eez_raster_df,
                                       mc.cores = 30)
})
```

```{r}
entry_rcp45_2050_df <- bind_rows(entry_rcp45_2050_list) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse
system.time({
entry_rcp45_2050_sum_df <- entry_rcp45_2050_df %>% 
  fgroup_by(MRGID, x, y) %>% 
  fcount() %>% 
  mutate(scenario = "RCP45_2050")
})
```

```{r}
# Took 34 seconds over 30 cores
system.time({
  entry_rcp45_2100_list <- parallel::mclapply(aphiaid_threat, FUN = extract_eez_entry_pixels_df_db, 
                                       df = df_threatened, 
                                       s = "RCP45_2100",
                                       eez_spp_entry = eez_spp_entry,
                                       eez_raster_df = eez_raster_df,
                                       mc.cores = 30)
})
```

```{r}
entry_rcp45_2100_df <- bind_rows(entry_rcp45_2100_list) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse
system.time({
entry_rcp45_2100_sum_df <- entry_rcp45_2100_df %>% 
  fgroup_by(MRGID, x, y) %>% 
  fcount() %>% 
  mutate(scenario = "RCP45_2100")
})
```

```{r}
# Took 34 seconds over 30 cores (if doesn't work, helps to clear memory)
system.time({
  entry_rcp85_2100_list <- parallel::mclapply(aphiaid_threat, FUN = extract_eez_entry_pixels_df_db, 
                                       df = df_threatened, 
                                       s = "RCP85_2100",
                                       eez_spp_entry = eez_spp_entry,
                                       eez_raster_df = eez_raster_df,
                                       mc.cores = 30)
})
```

```{r}
entry_rcp85_2100_df <- bind_rows(entry_rcp85_2100_list) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse
system.time({
entry_rcp85_2100_sum_df <- entry_rcp85_2100_df %>% 
  fgroup_by(MRGID, x, y) %>% 
  fcount() %>% 
  mutate(scenario = "RCP85_2100")
})
```

```{r}
# Bind the entry cells together, add in GEONAME for EEZs
eez_sum_threatened_scenarios <- rbind(entry_rcp45_2050_sum_df,
                                      entry_rcp45_2100_sum_df,
                                      entry_rcp85_sum_df,
                                      entry_rcp85_2100_sum_df) %>% 
  # Threshold for species presence is 10 pixels
  mutate(pxl_thres = "10") %>% 
  rename(n_entry_spp = N)

eez_sum_threatened_scenarios_mrgid <- left_join(eez_sum_threatened_scenarios, eez_ids) %>% 
  dplyr::select(x, y, GEONAME, MRGID, pxl_thres, scenario, n_entry_spp)
```

```{r}
# save data as rds
saveRDS(eez_sum_threatened_scenarios_mrgid, here("data/iucn_atrisk_10pxlthres_entry_pxls_072826.rds"))
```

## Species exits
```{r}
# Get counts of entry into new, exit from historical etc. by EEZ
eez_spp_exit <- eez_spp_pxls_long %>% 
  filter(spp_change == "exit from historical")
```

# Test subsample 

```{r}
set.seed(123)
test_spp_30 <- slice_sample(df_threatened, n = 100)
files_30 <- test_spp_30$f
aphiaid_30 <- test_spp_30$aphiaid
```

```{r}
system.time({
  eez_list_exit_count_df <- parallel::mclapply(aphiaid_30, FUN = extract_eez_exit_pixels_df_db, 
                                       df = test_spp_30, 
                                       s = "RCP85_2050",
                                       eez_spp_exit = eez_spp_exit,
                                       eez_raster_df = eez_raster_df,
                                       mc.cores = 10)
})
```

```{r}
eez_exit_count_df <- bind_rows(eez_list_exit_count_df) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse
system.time({
eez_exit_sum_df <- eez_exit_count_df %>% 
  fgroup_by(MRGID, x, y) %>% 
  fcount()
})
```

```{r}
# Plot to see if stripes are gone
ggplot() +
  geom_sf(data = eez) +
  coord_sf(expand = FALSE) +
  geom_tile(data = eez_exit_sum_df, aes(x = x, y = y, fill = N, color = N)) +
  #geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5)+
  theme_minimal()
```

## Run full sample for exits

```{r}
# Took 30 seconds over 30 cores
system.time({
  exit_rcp85_list <- parallel::mclapply(aphiaid_threat, FUN = extract_eez_exit_pixels_df_db, 
                                       df = df_threatened, 
                                       s = "RCP85_2050",
                                       eez_spp_exit = eez_spp_exit,
                                       eez_raster_df = eez_raster_df,
                                       mc.cores = 30)
})
```

```{r}
exit_rcp85_df <- bind_rows(exit_rcp85_list) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse
system.time({
exit_rcp85_sum_df <- exit_rcp85_df %>% 
  fgroup_by(MRGID, x, y) %>% 
  fcount() %>% 
  mutate(scenario = "RCP85_2050")
})
```

```{r}
exit_rcp85_sum_df_200nm <- exit_rcp85_sum_df %>% 
  filter(MRGID %in% eez_200nm_list)
```

```{r}
# Plot to see if stripes are gone (all EEZ regions; including joint regime and contested regions)
ggplot() +
  geom_sf(data = eez %>% filter(MRGID %in% eez_200nm_list)) +
  coord_sf(expand = FALSE) +
  geom_tile(data = exit_rcp85_sum_df_200nm, aes(x = x, y = y, fill = N, color = N)) +
  scale_fill_viridis_c(limits = c(1,40), 
                      breaks = c(1, 2, 4, 8, 16, 28, 40),
                      direction = -1,
                      trans = "log", # visually stretches the lower value data
                      guide = guide_colorbar(
                      barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                      barheight = unit(0.4, "cm"),
                      ticks = TRUE,
                      ticks.colour = "white",
                      ticks.linewidth = 0.8
  )) +
  scale_color_viridis_c(limits = c(1,40), 
                        breaks = c(1, 2, 4, 8, 16, 28, 40),
                        direction = -1,
                        trans = "log", # visually stretches the lower value data
                        guide = "none")  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "species exit count",
       color = "species exit count",
       x = "",
       y = "") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt")) 
```

```{r}
# Plot to see if stripes are gone
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8309)) +
  coord_sf(expand = FALSE) +
  geom_tile(data = exit_rcp85_sum_df_200nm %>% filter(MRGID == 8309), aes(x = x, y = y, fill = N, color = N)) +
  scale_fill_viridis_c(direction = -1) +
  scale_color_viridis_c(direction = -1) +
  #geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5)+
  theme_minimal()
```

```{r}
# Took 30 seconds over 30 cores
system.time({
  exit_rcp85_2100_list <- parallel::mclapply(aphiaid_threat, FUN = extract_eez_exit_pixels_df_db, 
                                       df = df_threatened, 
                                       s = "RCP85_2100",
                                       eez_spp_exit = eez_spp_exit,
                                       eez_raster_df = eez_raster_df,
                                       mc.cores = 30)
})
```

```{r}
exit_rcp85_2100_df <- bind_rows(exit_rcp85_2100_list) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse
system.time({
exit_rcp85_2100_sum_df <- exit_rcp85_2100_df %>% 
  fgroup_by(MRGID, x, y) %>% 
  fcount() %>% 
  mutate(scenario = "RCP85_2100")
})
```

```{r}
# Took 30 seconds over 30 cores
system.time({
  exit_rcp45_2050_list <- parallel::mclapply(aphiaid_threat, FUN = extract_eez_exit_pixels_df_db, 
                                       df = df_threatened, 
                                       s = "RCP45_2050",
                                       eez_spp_exit = eez_spp_exit,
                                       eez_raster_df = eez_raster_df,
                                       mc.cores = 30)
})
```

```{r}
exit_rcp45_2050_df <- bind_rows(exit_rcp45_2050_list) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse
system.time({
exit_rcp45_2050_sum_df <- exit_rcp45_2050_df %>% 
  fgroup_by(MRGID, x, y) %>% 
  fcount() %>% 
  mutate(scenario = "RCP45_2050")
})
```

```{r}
# Took 30 seconds over 30 cores
system.time({
  exit_rcp45_2100_list <- parallel::mclapply(aphiaid_threat, FUN = extract_eez_exit_pixels_df_db, 
                                       df = df_threatened, 
                                       s = "RCP45_2100",
                                       eez_spp_exit = eez_spp_exit,
                                       eez_raster_df = eez_raster_df,
                                       mc.cores = 30)
})
```

```{r}
exit_rcp45_2100_df <- bind_rows(exit_rcp45_2100_list) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse
system.time({
exit_rcp45_2100_sum_df <- exit_rcp45_2100_df %>% 
  fgroup_by(MRGID, x, y) %>% 
  fcount() %>% 
  mutate(scenario = "RCP45_2100")
})
```

```{r}
# Bind the exit cells together, add in GEONAME for EEZs
eez_sum_exit_scenarios <- rbind(exit_rcp45_2050_sum_df,
                                      exit_rcp45_2100_sum_df,
                                      exit_rcp85_sum_df,
                                      exit_rcp85_2100_sum_df) %>% 
  # Threshold for species presence is 10 pixels
  mutate(pxl_thres = "10") %>% 
  rename(n_exit_spp = N)

eez_sum_exit_scenarios_mrgid <- left_join(eez_sum_exit_scenarios, eez_ids) %>% 
  dplyr::select(x, y, GEONAME, MRGID, pxl_thres, scenario, n_exit_spp)
```

```{r}
# save data as rds
saveRDS(eez_sum_exit_scenarios_mrgid, here("data/iucn_atrisk_10pxlthres_exit_pxls_072926.rds"))
```

# Run counts of Least Concern species 

```{r}
# Filter iucn categories for least concern (LC) species
iucn_least_concern <- iucn_new_cat %>% 
  filter(redlist_category == "LC")
```

```{r}
# Merge in the iucn_lookup_sdm to get the AphiaID for the least concern species
aphiaid_least_concern <- inner_join(iucn_lookup_sdm, iucn_least_concern, by = c("iucn_assessment_id" = "assessment_id")) 
```

```{r}
# Get aphiaids for the least concern species
aphiaid_least_concern <- unique(aphiaid_least_concern$aphia_id)
```

```{r}
# Filter for the AphiaID of the least concern species - get 9,714 least concern species with SDMs
df_least_concern <- df %>% 
  filter(aphiaid %in% aphiaid_least_concern)
```

```{r}
test_f <- read_parquet("/home/shares/data-aquax/SDM/FINAL_EMSDM_EMMEAN_SP_1017389.parquet")
```

# Test subsample for least concern pixels
```{r}
set.seed(123)
test_spp_30 <- slice_sample(df_least_concern, n = 1000)
files_30 <- test_spp_30$f
aphiaid_30 <- test_spp_30$aphiaid
```


```{r}
system.time({
 least_concern_list_count_df <- parallel::mclapply(aphiaid_30, FUN = least_concern_count, 
                                       df = test_spp_30, 
                                       scenario = "Current",
                                       n_threads = 2,
                                       mc.cores = 40)
})
```

```{r}
least_concern_count_df <- bind_rows(least_concern_list_count_df) 
```

```{r}
# Sum by XY cell, using collapse
system.time({
lc_sum_df <- least_concern_count_df %>% 
  # drop binary = 0 (no pixels above threshold)
  fsubset(binary == 1) %>%
  # count the number of species in each XY cell
  fgroup_by(x, y) %>% 
  fcount()
})
```

```{r}
# dplyr 10x slower
system.time({
lc_sum_df_dplyr <- least_concern_count_df %>% 
  filter(binary == 1) %>%
  group_by(x, y) %>% 
  summarize(n_lc_spp = n()) %>% 
  ungroup()
})
```


```{r}
ggplot() +
  geom_sf(data = eez) +
  coord_sf(expand = FALSE) +
  geom_tile(data = lc_sum_df, aes(x = x, y = y,
                                               fill = N,
                                               color = N)) +
  scale_y_continuous(limits = c(-90, 90)) +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  scale_color_viridis_c(option = "plasma", direction = -1) +
  theme_minimal() + 
  labs(fill = "lc species count",
       color = "lc species count") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt"))
```

# Run full lc set

```{r}
# 5 minute for all data (-180 to 180)
system.time({
 lc_list_count_neg90_0_df <- parallel::mclapply(aphiaid_least_concern, FUN = least_concern_count_xmin_xmax, 
                                       df = df_least_concern, 
                                       scenario = "RCP85_2050",
                                       xmin = -180,
                                       xmax = 180,
                                       n_threads = 2,
                                       mc.cores = 40)
})
```

```{r}
lc_count_df <- bind_rows(lc_list_count_neg90_0_df) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse (2 min)
system.time({
lc_sum_df <- lc_count_df %>% 
  fgroup_by(x, y) %>% 
  fcount()
})
```

```{r}
lc_plot_count <- ggplot() +
  geom_sf(data = eez) +
  coord_sf(expand = FALSE) +
  geom_tile(data = lc_sum_df, aes(x = x, y = y,
                                               fill = N,
                                               color = N)) +
  scale_y_continuous(limits = c(-90, 90)) +
  scale_fill_viridis_c(limits = c(1,3000), 
                      breaks = c(1, 25, 50, 100, 500, 1500, 3000),
                      direction = -1,
                      trans = "log", # visually stretches the lower value data
                      guide = guide_colorbar(
                      barwidth  = unit(10, "cm"),   # wide bar for bottom placement
                      barheight = unit(0.4, "cm"),
                      ticks = TRUE,
                      ticks.colour = "white",
                      ticks.linewidth = 0.8)) +
  scale_color_viridis_c(limits = c(1,3000), 
                        breaks = c(1, 25, 50, 100, 500, 1500, 3000),
                        direction = -1,
                        trans = "log", # visually stretches the lower value data
                        guide = "none") +
  theme_minimal() + 
  labs(fill = "count",
       color = "count",
       title = "Least concern species count - RCP 8.5 2050") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt"))
```

```{r}
# Save plot of least concern species
ggsave(here("figures/iucn_lc_spp_rcp85_2050_080426.png"), lc_plot_count, width = 10, height = 5, dpi = 300)
```


```{r}
lc_sum_neg90_neg180 <- lc_sum_df %>% 
  filter(x < -90)

lc_sum_neg90_zero <- lc_sum_df %>% 
  filter(x >= -90 & x < 0)

lc_sum_zero_90 <- lc_sum_df %>% 
  filter(x >= 0 & x < 90)

lc_sum_90_180 <- lc_sum_df %>% 
  filter(x >= 90)
```


```{r}
saveRDS(lc_sum_neg90_neg180, here("data/iucn_least_concern/iucn_lc_rcp85_2050_neg180_to_neg90_080426.rds"))

saveRDS(lc_sum_neg90_zero, here("data/iucn_least_concern/iucn_lc_rcp85_2050_neg90_to_zero_080426.rds"))

saveRDS(lc_sum_zero_90, here("data/iucn_least_concern/iucn_lc_rcp85_2050_zero_to_90_080426.rds"))

saveRDS(lc_sum_90_180, here("data/iucn_least_concern/iucn_lc_rcp85_2050_90_to_180_080426.rds"))
```

Next, run current scenario least concern species

```{r}
# 6 minutes for Current data
system.time({
 lc_list_count_current_df <- parallel::mclapply(aphiaid_least_concern, FUN = least_concern_count, 
                                       df = df_least_concern, 
                                       scenario = "Current",
                                       n_threads = 2,
                                       mc.cores = 40)
})
```

```{r}
# Too many rows in list, need to do in sections and filter out the binary = 0 rows to reduce memory usage
lc_list_current_1 <- lc_list_count_current_df[1:4000]
lc_list_current_2 <- lc_list_count_current_df[4001:9741]
```

```{r}
lc_count_current_1_df <- data.table::rbindlist(lc_list_current_1) %>% 
  fsubset(binary == 1)

lc_count_current_2_df <- data.table::rbindlist(lc_list_current_2) %>% 
  fsubset(binary == 1)
```

```{r}
lc_count_current_1_df_neg <- lc_count_current_1_df %>% 
  fsubset(x < 0)

lc_count_current_1_df_pos <- lc_count_current_1_df %>% 
  fsubset(x >= 0)
```

```{r}
lc_count_current_2_df_neg <- lc_count_current_2_df %>% 
  fsubset(x < 0)

lc_count_current_2_df_pos <- lc_count_current_2_df %>% 
  fsubset(x >= 0)
```

```{r}
# recombine the two sets of dataframes in order to group and summarize
lc_count_current_df_neg <- data.table::rbindlist(list(lc_count_current_1_df_neg, lc_count_current_2_df_neg))
lc_count_current_df_pos <- data.table::rbindlist(list(lc_count_current_1_df_pos, lc_count_current_2_df_pos))
```

```{r}
# Sum by XY cell, using collapse (2 min)
system.time({
lc_sum_current_df_neg <- lc_count_current_df_neg %>% 
  # count the number of species in each XY cell
  fgroup_by(x, y) %>% 
  fcount()
})
```

```{r}
# Sum by XY cell, using collapse (2 min)
system.time({
lc_sum_current_df_pos <- lc_count_current_df_pos %>% 
  # count the number of species in each XY cell
  fgroup_by(x, y) %>% 
  fcount()
})
```

```{r}
lc_plot_current_count <- ggplot() +
  geom_sf(data = eez) +
  coord_sf(expand = FALSE) +
  geom_tile(data = lc_sum_current_df_neg, aes(x = x, y = y,
                                               fill = N,
                                               color = N)) +
  geom_tile(data = lc_sum_current_df_pos, aes(x = x, y = y,
                                               fill = N,
                                               color = N)) +                          
  scale_y_continuous(limits = c(-90, 90)) +
  scale_fill_viridis_c(limits = c(1,3200), 
                      breaks = c(1, 25, 50, 100, 500, 1500, 3200),
                      direction = -1,
                      trans = "log", # visually stretches the lower value data
                      guide = guide_colorbar(
                      barwidth  = unit(10, "cm"),   # wide bar for bottom placement
                      barheight = unit(0.4, "cm"),
                      ticks = TRUE,
                      ticks.colour = "white",
                      ticks.linewidth = 0.8)) +
  scale_color_viridis_c(limits = c(1,3200), 
                        breaks = c(1, 25, 50, 100, 500, 1500, 3200),
                        direction = -1,
                        trans = "log", # visually stretches the lower value data
                        guide = "none") +
  theme_minimal() + 
  labs(fill = "count",
       color = "count",
       title = "Least concern species count - Current") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt"))
```

```{r}
# Save plot of least concern species
ggsave(here("figures/iucn_lc_spp_current_080526.png"), lc_plot_current_count, width = 10, height = 5, dpi = 300)
```


```{r}
# Prepare data for saving as rds files
lc_sum_current_neg90_neg180 <- lc_sum_current_df_neg %>% 
  filter(x < -90)

lc_sum_current_neg90_zero <- lc_sum_current_df_neg %>% 
  filter(x >= -90 & x < 0)

lc_sum_current_zero_90 <- lc_sum_current_df_pos %>% 
  filter(x >= 0 & x < 90)

lc_sum_current_90_180 <- lc_sum_current_df_pos %>% 
  filter(x >= 90)
```

```{r}
saveRDS(lc_sum_current_neg90_neg180, here("data/iucn_least_concern/iucn_lc_current_neg180_to_neg90_080526.rds"))

saveRDS(lc_sum_current_neg90_zero, here("data/iucn_least_concern/iucn_lc_current_neg90_to_zero_080526.rds"))

saveRDS(lc_sum_current_zero_90, here("data/iucn_least_concern/iucn_lc_current_zero_to_90_080526.rds"))

saveRDS(lc_sum_current_90_180, here("data/iucn_least_concern/iucn_lc_current_90_to_180_080526.rds"))
```