---
title: "AquaX and IUCN status"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
library(arrow)
library(collapse)
```

This document uses the IUCN lookup and threat category data created in the chi repository. It connects IUCN status to AquaX data using the Aphia ID, which is a unique identifier for marine species in the World Register of Marine Species (WoRMS). 

```{r}
iucn_lookup <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/_data_raw/iucn_redlist_2026/iucn_worms_aphia_id_complete.csv")
```

```{r}
iucn_new_cat <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/_data_raw/iucn_redlist_2026/iucn_species_assessments.csv")
```

```{r}
# Load functions for processing AquaX data
source(here( "common_fxns.R"))
```

We'll use the functions to process the AquaX data for the count of threatened species entering and exiting EEZs.

## EEZ 

```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
eez_simple <- eez |>
  select(MRGID, GEONAME, geom)
```

```{r}
ggplot() +
  geom_sf(data = eez, aes()) +
  geom_sf(data = eez |> filter(GEONAME == "Russian Exclusive Economic Zone"), aes(), fill = "red") +
  theme_void() +
  theme(legend.position = "none")
```

```{r}
# Create a blank raster matching AquaX resolution (EEZs should be large enough to just do 0.05 degree raster)
eez_raster_blank <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05,
                  crs = "EPSG:4326")
```

```{r}
eez_raster_geoname <- rasterize(eez_simple, eez_raster_blank, field = "GEONAME")
```

```{r}
eez_raster_mrgid <- rasterize(eez_simple, eez_raster_blank, field = "MRGID")
```

# IUCN threatened species

```{r}
# Filter the iucn_lookup for species with an SDM
iucn_lookup_sdm <- iucn_lookup %>% filter(aquax == "SDM")
```

```{r}
# Filter iucn categories for threatened species (CR, EN, VU) and near threatened (NT) species
iucn_threatened <- iucn_new_cat %>% 
  filter(redlist_category %in% c("CR","EN", "VU", "NT"))
```

```{r}
# Merge in the iucn_lookup_sdm to get the AphiaID for the threatened species
aphiaid_threatened <- inner_join(iucn_lookup_sdm, iucn_threatened, by = c("iucn_assessment_id" = "assessment_id")) 
```

```{r}
# Get aphiaids for the threatened species
aphiaid_threat <- unique(aphiaid_threatened$aphia_id)
```

## Prep AquaX data
We'll be working with the 1,047 thretened aphiaids

```{r}
# read in meta data on the aquax species
load("/home/shares/data-aquax/META_02122025.Rdata")
```

```{r}
meta_clean <- META %>% 
  mutate(worms_name = tolower(ScientificName)) %>% 
  select(!ScientificName) %>% 
  # simplify for now for easier viewing
  select(AphiaID, SPID, worms_name, GROUP, NR_CR, range_pixels, SDM)
```

```{r}
spp_fs <- list.files('/home/shares/data-aquax/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(META, by = c('aphiaid' = 'AphiaID'))
```

```{r}
# Filter for the AphiaID of the threatend species - get 1,047 threatened species and 276 non-threatened species with SDMs
df_threatened <- df %>% 
  filter(aphiaid %in% aphiaid_threat)
```

```{r}
test_f <- read_parquet("/home/shares/data-aquax/SDM/FINAL_EMSDM_EMMEAN_SP_1017389.parquet")
```

```{r}
set.seed(123)
test_spp_30 <- slice_sample(df_threatened, n = 30)
files_30 <- test_spp_30$f
aphiaid_30 <- test_spp_30$aphiaid
```

```{r}
test_r <- rast(test_f, crs = "EPSG:4326") %>%
    extend(ext(eez_raster_mrgid))
```

```{r}
# Test plot
plot(test_r_eq$RCP45_2050)
```

```{r}
r_binary45_2050 <- ifel(test_r[["RCP45_2050"]] >= mean(test_f$cutoff, na.rm = TRUE), 1, NA)
```

```{r}
# Have to use the EEZ raster with the numeric IDs (MRGID) for zonal stats
# Try // - took 2 min
system.time({
  eez_list_zonal <- parallel::mclapply(aphiaid_30, FUN = extract_eez_zonal, 
                                       df = test_spp_30, 
                                       eez_raster = eez_raster_mrgid, 
                                       mc.cores = 10)
})
```

```{r}
eez_list_all_zonal <- bind_rows(eez_list_zonal) 
```

```{r}
eez_ids <- eez %>% 
  st_drop_geometry() %>% 
  dplyr::select(MRGID, GEONAME)
```

```{r}
eez_list_geoname <- left_join(eez_list_all_zonal, eez_ids)
```


```{r}
eez_raster_df <- as.data.frame(eez_raster_mrgid_eq, xy = TRUE, cells = TRUE)
eez_raster_df_non_eq <- as.data.frame(eez_raster_mrgid, xy = TRUE, cells = TRUE)
```

## Test out full threatened sample

```{r}
files <- df_threatened$f
aphiaid <- df_threatened$aphiaid
```

```{r}
# Try // - took 20 min for 1323 species and 4 scenarios (21 min for 987 species and 4 scenarios and 20 cores)
system.time({
  eez_list_threatened <- parallel::mclapply(aphiaid, FUN = extract_eez_zonal, 
                                            df = df_threatened,
                                            eez_raster = eez_raster_mrgid,
                                            mc.cores = 40)
})
```

```{r}
eez_list_threatened_bind <- bind_rows(eez_list_threatened) 
```

```{r}
# Merge in EEZ ids
eez_threatened_ids <- left_join(eez_list_threatened_bind, (eez_simple %>% st_drop_geometry()))
```

```{r}
# Save data
write_csv(eez_threatened_ids, here("data/iucn_threatened_spp_eez_pxls_061626.csv"))
```

# Identify pixels entering/exiting EEZs

```{r}
# Organize dataframe so that's it's easier to select a scenario and compare to current
eez_spp_pxls_long <- eez_threatened_ids %>% 
  pivot_longer(cols = starts_with("n_RCP"),
               names_to = "scenario",
               values_to = "pxl_count") %>% 
  mutate(scenario = str_remove(scenario, "n_")) %>% 
  # Replace NA with 0, otherwise will be assigned to "continued presence"
  replace_na(list(pxl_count = 0, n_current = 0)) %>% 
  # Set 10 pixel threshold (at least ten pixels need in EEZ for species presence)
  mutate(spp_change = case_when(n_current < 10 &  pxl_count >= 10 ~ 'entry into new',
                            n_current >= 10  & pxl_count < 10 ~ 'exit from historical',
                            n_current < 10  & pxl_count < 10 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

```{r}
# Get counts of entry into new, exit from historical etc. by EEZ
eez_spp_entry <- eez_spp_pxls_long %>% 
  filter(spp_change == "entry into new")
```

```{r}
# Create dataframe of EEZ raster
eez_raster_df <- as.data.frame(eez_raster_geoname, xy = TRUE)
```

```{r}
# Join to the test df
test_f_eez <- inner_join(test_f, eez_raster_df)

test_f_us <- test_f_eez %>% 
  filter(GEONAME == "United States Exclusive Economic Zone") %>% 
  #mutate(bi_RCP45_2050 = ifelse(RCP45_2050 >= cutoff, 1, 0))
  filter(RCP45_2050 >= cutoff) %>% 
  mutate(bi_RCP45_2050 = 1)
```

```{r}
test_r_us_r <- rast(test_f_us,  crs = "EPSG:4326") 
```

```{r}
system.time({
  eez_list_count_df <- parallel::mclapply(aphiaid_30, FUN = extract_eez_entry_pixels_df, 
                                       df = test_spp_30, 
                                       s = "RCP45_2050",
                                       eez_spp_entry = eez_spp_entry,
                                       eez_raster_df = eez_raster_df,
                                       mc.cores = 10)
})
```

```{r}
eez_count_df <- bind_rows(eez_list_count_df)
```

```{r}
# Sum by XY cell, create raster
eez_sum_df <- eez_count_df %>% 
  group_by(x, y, GEONAME) %>% 
  summarize(n_entry_spp = sum(binary, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
eez_sum_raster_30 <- rast(eez_sum_df, crs = "EPSG:4326")
```

```{r}
plot(eez_sum_raster_30$n_entry_spp)
```

```{r}
# Try // - took ~1 min for 1,323 species
system.time({
  eez_pixels_threatened <- parallel::mclapply(aphiaid, FUN = extract_eez_entry_pixels_df, 
                                            df = df_threatened,
                                            s = "RCP45_2050",
                                            eez_spp_entry = eez_spp_entry,
                                            eez_raster_df = eez_raster_df,
                                            mc.cores = 20)
})
```

```{r}
eez_count_threatened_df <- bind_rows(eez_pixels_threatened)
```

```{r}
# Sum by XY cell 
eez_sum_threatened_df <- eez_count_threatened_df  %>% 
  group_by(x, y, GEONAME) %>% 
  summarize(n_entry_spp = sum(binary, na.rm = TRUE)) %>% 
  mutate(scenario = "RCP45_2050") %>% 
  ungroup()
```

```{r}
# Try // - took ~1 min for 987 species
system.time({
  eez_pixels_threatened_45_2100 <- parallel::mclapply(aphiaid, FUN = extract_eez_entry_pixels_df, 
                                            df = df_threatened,
                                            s = "RCP45_2100",
                                            eez_spp_entry = eez_spp_entry,
                                            eez_raster_df = eez_raster_df,
                                            mc.cores = 20)
})
```

```{r}
eez_count_threatened_45_2100_df <- bind_rows(eez_pixels_threatened_45_2100)
```

```{r}
# Sum by XY cell 
eez_sum_threatened_45_2100_df <- eez_count_threatened_45_2100_df  %>% 
  group_by(x, y, GEONAME) %>% 
  summarize(n_entry_spp = sum(binary, na.rm = TRUE)) %>% 
  mutate(scenario = "RCP45_2100") %>% 
  ungroup()
```

```{r}
# Try // - took ~1 min for 1,323 species
system.time({
  eez_pixels_threatened_85_2050 <- parallel::mclapply(aphiaid, FUN = extract_eez_entry_pixels_df, 
                                            df = df_threatened,
                                            s = "RCP85_2050",
                                            eez_spp_entry = eez_spp_entry,
                                            eez_raster_df = eez_raster_df,
                                            mc.cores = 20)
})
```

```{r}
eez_count_threatened_85_2050_df <- bind_rows(eez_pixels_threatened_85_2050)
```

```{r}
# Sum by XY cell 
eez_sum_threatened_85_2050_df <- eez_count_threatened_85_2050_df  %>% 
  group_by(x, y, GEONAME) %>% 
  summarize(n_entry_spp = sum(binary, na.rm = TRUE)) %>% 
  mutate(scenario = "RCP85_2050") %>% 
  ungroup()
```

```{r}
# Try // - took ~1 min for 1,323 species
system.time({
  eez_pixels_threatened_85_2100 <- parallel::mclapply(aphiaid, FUN = extract_eez_entry_pixels_df, 
                                            df = df_threatened,
                                            s = "RCP85_2100",
                                            eez_spp_entry = eez_spp_entry,
                                            eez_raster_df = eez_raster_df,
                                            mc.cores = 20)
})
```

```{r}
eez_count_threatened_85_2100_df <- bind_rows(eez_pixels_threatened_85_2100)
```

```{r}
# Sum by XY cell 
eez_sum_threatened_85_2100_df <- eez_count_threatened_85_2100_df  %>% 
  group_by(x, y, GEONAME) %>% 
  summarize(n_entry_spp = sum(binary, na.rm = TRUE)) %>% 
  mutate(scenario = "RCP85_2100") %>% 
  ungroup()
```

```{r}
# Bind the entry cells together, add in MRGID for EEZs
eez_sum_threatened_scenarios <- rbind(eez_sum_threatened_df,
                                      eez_sum_threatened_45_2100_df,
                                      eez_sum_threatened_85_2050_df,
                                      eez_sum_threatened_85_2100_df) %>% 
  # Threshold for species presence is 10 pixels
  mutate(pxl_thres = "10")

eez_sum_threatened_scenarios_mrgid <- left_join(eez_sum_threatened_scenarios, eez_ids) %>% 
  dplyr::select(x, y, GEONAME, MRGID, pxl_thres, scenario, n_entry_spp)
```

```{r}
# save data as rds
saveRDS(eez_sum_threatened_scenarios_mrgid, here("data/iucn_threatened_10pxlthres_entry_pxls_061726.rds"))
```

## Extract exit pixels

```{r}
eez_spp_exit <- eez_spp_pxls_long %>% 
  filter(spp_change == "exit from historical")
```

```{r}
# Try // - took ~1 min for 1323 species
system.time({
  eez_pixels_exit_45_2050 <- parallel::mclapply(aphiaid, 
                                                      FUN = extract_eez_exit_pixels_df, 
                                                      df = df_threatened,
                                                      s = "RCP45_2050",
                                                      eez_spp_exit = eez_spp_exit,
                                                      eez_raster_df = eez_raster_df,
                                                      mc.cores = 20)
})
```

```{r}
eez_pixels_exit_45_2050_df <- bind_rows(eez_pixels_exit_45_2050)
```

```{r}
# Sum by XY cell 
eez_sum_exit_45_2050_df <- eez_pixels_exit_45_2050_df  %>% 
  group_by(x, y, GEONAME) %>% 
  summarize(n_exit_spp = sum(binary_loss, na.rm = TRUE)) %>% 
  mutate(scenario = "RCP45_2050") %>% 
  ungroup()
```

```{r}
# Try // - took ~1 min for 906 species
system.time({
  eez_pixels_exit_45_2100 <- parallel::mclapply(aphiaid, 
                                                      FUN = extract_eez_exit_pixels_df, 
                                                      df = df_threatened,
                                                      s = "RCP45_2100",
                                                      eez_spp_exit = eez_spp_exit,
                                                      eez_raster_df = eez_raster_df,
                                                      mc.cores = 20)
})
```

```{r}
eez_pixels_exit_45_2100_df <- bind_rows(eez_pixels_exit_45_2100)
```

```{r}
# Sum by XY cell 
eez_sum_exit_45_2100_df <- eez_pixels_exit_45_2100_df  %>% 
  group_by(x, y, GEONAME) %>% 
  summarize(n_exit_spp = sum(binary_loss, na.rm = TRUE)) %>% 
  mutate(scenario = "RCP45_2100") %>% 
  ungroup()
```

```{r}
# Try // - took ~1 min for 1323 species
system.time({
  eez_pixels_exit_85_2050 <- parallel::mclapply(aphiaid, 
                                                      FUN = extract_eez_exit_pixels_df, 
                                                      df = df_threatened,
                                                      s = "RCP85_2050",
                                                      eez_spp_exit = eez_spp_exit,
                                                      eez_raster_df = eez_raster_df,
                                                      mc.cores = 20)
})
```

```{r}
eez_pixels_exit_85_2050_df <- bind_rows(eez_pixels_exit_85_2050)
```

```{r}
# Sum by XY cell 
eez_sum_exit_85_2050_df <- eez_pixels_exit_85_2050_df  %>% 
  group_by(x, y, GEONAME) %>% 
  summarize(n_exit_spp = sum(binary_loss, na.rm = TRUE)) %>% 
  mutate(scenario = "RCP85_2050") %>% 
  ungroup()
```

```{r}
# Try // - took ~1 min for 906 species
system.time({
  eez_pixels_exit_85_2100 <- parallel::mclapply(aphiaid, 
                                                      FUN = extract_eez_exit_pixels_df, 
                                                      df = df_threatened,
                                                      s = "RCP85_2100",
                                                      eez_spp_exit = eez_spp_exit,
                                                      eez_raster_df = eez_raster_df,
                                                      mc.cores = 20)
})
```

```{r}
eez_pixels_exit_85_2100_df <- bind_rows(eez_pixels_exit_85_2100)
```

```{r}
# Sum by XY cell 
eez_sum_exit_85_2100_df <- eez_pixels_exit_85_2100_df  %>% 
  group_by(x, y, GEONAME) %>% 
  summarize(n_exit_spp = sum(binary_loss, na.rm = TRUE)) %>% 
  mutate(scenario = "RCP85_2100") %>% 
  ungroup()
```

```{r}
# Bind the exit cells together, add in MRGID for EEZs
eez_sum_exit_scenarios <- rbind(eez_sum_exit_45_2050_df,
                                      eez_sum_exit_45_2100_df,
                                      eez_sum_exit_85_2050_df,
                                      eez_sum_exit_85_2100_df) %>% 
  # Threshold for species presence is 10 pixels
  mutate(pxl_thres = "10")

eez_sum_exit_scenarios_mrgid <- left_join(eez_sum_exit_scenarios, eez_ids) %>% 
  dplyr::select(x, y, GEONAME, MRGID, pxl_thres, scenario, n_exit_spp)
```

```{r}
# save data as rds
saveRDS(eez_sum_exit_scenarios_mrgid, here("data/iucn_threatened_10pxlthres_exit_pxls_061726.rds"))
```

# Identify continued presence pixels

```{r}
eez_spp_cp <- eez_spp_pxls_long %>% 
  filter(spp_change == "continued presence")
```

```{r}
# Try // - took ~1 min for 1323 species
system.time({
  eez_pixels_cp_85_2050 <- parallel::mclapply(aphiaid, 
                                                      FUN = extract_eez_cp_pixels_df, 
                                                      df = df_threatened,
                                                      s = "RCP85_2050",
                                                      eez_spp_cp = eez_spp_cp,
                                                      eez_raster_df = eez_raster_df,
                                                      mc.cores = 20)
})
```

```{r}
eez_pixels_cp_85_2050_df <- bind_rows(eez_pixels_cp_85_2050)
```

```{r}
# Sum by XY cell 
eez_sum_cp_85_2050_df <- eez_pixels_cp_85_2050_df  %>% 
  group_by(x, y, GEONAME) %>% 
  summarize(n_cp_spp = sum(binary_loss, na.rm = TRUE)) %>% 
  mutate(scenario = "RCP85_2050",
         pxl_thres = "10") %>% 
  ungroup()
```

```{r}
eez_sum_cp_rcp85_2050_mrgid <- left_join(eez_sum_cp_85_2050_df, eez_ids) %>% 
  dplyr::select(x, y, GEONAME, MRGID, pxl_thres, scenario, n_cp_spp)
```

```{r}
# Save RDS for RCP 85 2050 for continued presence pixels
saveRDS(eez_sum_cp_rcp85_2050_mrgid, here("data/iucn_threatened_10pxlthres_rcp85_2050_cp_pxls_061826.rds"))
```

## Test out cp graphs 
```{r}
 ggplot() +
  geom_sf(data = eez) +
  coord_sf(expand = FALSE) +
  geom_tile(data = eez_sum_cp_85_2050_df, aes(x = x, y = y,
                                               fill = n_cp_spp,
                                               color = n_cp_spp)) +
  scale_fill_viridis_c(limits = c(1,300), direction = -1) +
  scale_color_viridis_c(limits = c(1,300), direction = -1)  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "species continued presence count",
       color = "species continued presence count") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt"))
```
# Merge in exit and entry species (RCP 85, 2050)

```{r}
eez_sum_cp_exit_85_2050 <- full_join(eez_sum_cp_85_2050_df, eez_sum_exit_85_2050_df)
eez_sum_all_85_2050 <- full_join(eez_sum_cp_exit_85_2050, eez_sum_threatened_85_2050_df) %>% 
  # Replace NA with 0
  #replace_na(list(n_cp_spp = 0, n_exit_spp = 0, n_entry_spp = 0)) %>% 
  mutate(n_current_spp = sum(n_cp_spp, n_exit_spp, na.rm = TRUE),
         n_future_spp = sum(n_cp_spp, n_entry_spp, na.rm = TRUE),
         # percent of current species that exit
         perc_exit_spp = 100 * n_exit_spp/(n_cp_spp + n_exit_spp),
         # percent of future species that entered
         perc_entry_spp = 100 * n_entry_spp/(n_cp_spp + n_entry_spp))
```

```{r}
 ggplot() +
  geom_sf(data = eez) +
  coord_sf(expand = FALSE) +
  geom_tile(data = eez_sum_all_85_2050, aes(x = x, y = y,
                                               fill = n_cp_spp,
                                               color = n_cp_spp)) +
  scale_fill_viridis_c(limits = c(1,300), direction = -1) +
  scale_color_viridis_c(limits = c(1,300), direction = -1)  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "species count",
       color = "species count") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt"))
```

```{r}
# Need to take a closer look at these and what they're saying 
ggplot() +
  geom_sf(data = eez) +
  coord_sf(expand = FALSE) +
  geom_tile(data = (eez_sum_all_85_2050 %>% filter(!is.na(perc_exit_spp))), aes(x = x, y = y,
                                               fill = perc_exit_spp,
                                               color = perc_exit_spp)) +
  scale_fill_viridis_c(limits = c(1,100), direction = -1) +
  scale_color_viridis_c(limits = c(1,100), direction = -1)  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "percent species lost",
       color = "percent species lost") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt"))
```

```{r}
 ggplot() +
  geom_sf(data = eez) +
  coord_sf(expand = FALSE) +
  geom_tile(data = (eez_sum_all_85_2050 %>% filter(!is.na(perc_entry_spp))), aes(x = x, y = y,
                                               fill = perc_entry_spp,
                                               color = perc_entry_spp)) +
  scale_fill_viridis_c(limits = c(1,100), direction = -1) +
  scale_color_viridis_c(limits = c(1,100), direction = -1)  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "percent species gain",
       color = "percent species gain") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt"))
```

# Run counts of Least Concern species 

```{r}
# Filter iucn categories for least concern (LC) species
iucn_least_concern <- iucn_new_cat %>% 
  filter(redlist_category == "LC")
```

```{r}
# Merge in the iucn_lookup_sdm to get the AphiaID for the least concern species
aphiaid_least_concern <- inner_join(iucn_lookup_sdm, iucn_least_concern, by = c("iucn_assessment_id" = "assessment_id")) 
```

```{r}
# Get aphiaids for the least concern species
aphiaid_least_concern <- unique(aphiaid_least_concern$aphia_id)
```

```{r}
# Filter for the AphiaID of the least concern species - get 9,714 least concern species with SDMs
df_least_concern <- df %>% 
  filter(aphiaid %in% aphiaid_least_concern)
```

```{r}
test_f <- read_parquet("/home/shares/data-aquax/SDM/FINAL_EMSDM_EMMEAN_SP_1017389.parquet")
```

# Test subsample for least concern pixels
```{r}
set.seed(123)
test_spp_30 <- slice_sample(df_least_concern, n = 1000)
files_30 <- test_spp_30$f
aphiaid_30 <- test_spp_30$aphiaid
```


```{r}
system.time({
 least_concern_list_count_df <- parallel::mclapply(aphiaid_30, FUN = least_concern_count, 
                                       df = test_spp_30, 
                                       scenario = "RCP85_2050",
                                       xmin = -180,
                                       xmax = -90,
                                       mc.cores = 40)
})
```

```{r}
least_concern_count_df <- bind_rows(least_concern_list_count_df)
```

```{r}
# Sum by XY cell, using collapse
system.time({
lc_sum_df <- least_concern_count_df %>% 
  fgroup_by(x, y) %>% 
  fcount()
})
```

```{r}
# dplyr 100x slower
system.time({
lc_sum_df_dplyr <- least_concern_count_df %>% 
  group_by(x, y) %>% 
  summarize(n_lc_spp = n()) %>% 
  ungroup()
})
```


```{r}
ggplot() +
  geom_sf(data = eez) +
  coord_sf(expand = FALSE) +
  geom_tile(data = lc_sum_df, aes(x = x, y = y,
                                               fill = N,
                                               color = N)) +
  #scale_fill_viridis_c(limits = c(1,300), direction = -1) +
  #scale_color_viridis_c(limits = c(1,300), direction = -1)  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "lc species count",
       color = "lc species count") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt"))
```

# Run full lc set

```{r}
# 2 minutes for first stripe (-180 to -90)
system.time({
 lc_list_count_90_180_df <- parallel::mclapply(aphiaid_least_concern, FUN = least_concern_count, 
                                       df = df_least_concern, 
                                       scenario = "RCP85_2050",
                                       xmin = 90,
                                       xmax = 180,
                                       mc.cores = 40)
})
```

```{r}
lc_count_90_180_df <- bind_rows(lc_list_count_90_180_df)
```

```{r}
# Sum by XY cell, using collapse (20 seconds)
system.time({
lc_sum_90_180_df <- lc_count_90_180_df %>% 
  fgroup_by(x, y) %>% 
  fcount()
})
```

```{r}
plot_lc_85_2050 <- ggplot() +
  geom_tile(data = lc_sum_df, aes(x = x, y = y,
                                               fill = N,
                                               color = N)) +
  geom_tile(data = lc_sum_neg90_zero_df, aes(x = x, y = y,
                                               fill = N,
                                               color = N)) +
    geom_tile(data = lc_sum_zero_90_df, aes(x = x, y = y,
                                               fill = N,
                                               color = N)) +
    geom_tile(data = lc_sum_90_180_df, aes(x = x, y = y,
                                               fill = N,
                                               color = N)) +
  scale_fill_viridis_c(limits = c(1,2100), 
                      breaks = c(1, 50, 100, 500, 1000, 2100),
                      direction = -1,
                      trans = "log", # visually stretches the lower value data
                      guide = guide_colorbar(
                      barwidth  = unit(6, "cm"),   # wide bar for bottom placement
                      barheight = unit(0.4, "cm"),
                      ticks = TRUE,
                      ticks.colour = "white",
                      ticks.linewidth = 0.8
  )) +
  scale_color_viridis_c(limits = c(1,2100), 
                        breaks = c(1, 50, 100, 500, 1000, 2100),
                        direction = -1,
                        trans = "log", # visually stretches the lower value data
                        guide = "none")  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "lc species count",
       color = "lc species count",
       title = "Least concern species count - RCP 8.5 2050") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt"))
```

```{r}
# Save plot of least concern species
ggsave(here("figures/iucn_lc_spp_rcp85_2050_071026.png"), plot_lc_85_2050, width = 10, height = 5, dpi = 300)
```


```{r}
# Save first stripe of data (-90 to -180, RCP85_2050)
saveRDS(lc_sum_df, here("data/iucn_least_concern/iucn_lc_rcp85_2050_neg180_to_neg90_070926.rds"))
```

```{r}
# Split next stripe into 2 b/c rds file is too large
lc_sum_neg90_neg45_df <- lc_sum_neg90_zero_df %>% 
  filter(x < -45)

lc_sum_neg45_zero_df <- lc_sum_neg90_zero_df %>% 
  filter(x >= -45)
```


```{r}
# Save next stripe of data (-90 to -45 and -45 to 0, RCP85_2050)
saveRDS(lc_sum_neg90_neg45_df, here("data/iucn_least_concern/iucn_lc_rcp85_2050_neg90_to_neg45_070926.rds"))
saveRDS(lc_sum_neg45_zero_df, here("data/iucn_least_concern/iucn_lc_rcp85_2050_neg45_to_0_070926.rds"))
```

```{r}
# Save next stripe of data (0 to 90, RCP85_2050)
saveRDS(lc_sum_zero_90_df, here("data/iucn_least_concern/iucn_lc_rcp85_2050_zero_to_90_070926.rds"))
```

```{r}
# Split next stripe into 3 b/c rds file is too large
lc_sum_90_120_df <- lc_sum_90_180_df %>% 
  filter(x < 120)

lc_sum_120_150_df <- lc_sum_90_180_df %>% 
  filter(x >= 120 & x < 150)

lc_sum_150_180_df <- lc_sum_90_180_df %>% 
  filter(x >= 150)

```

```{r}
# Save next stripe of data (-90 to -45 and -45 to 0, RCP85_2050)
saveRDS(lc_sum_90_120_df, here("data/iucn_least_concern/iucn_lc_rcp85_2050_90_to_120_070926.rds"))
saveRDS(lc_sum_120_150_df, here("data/iucn_least_concern/iucn_lc_rcp85_2050_120_to_150_070926.rds"))
saveRDS(lc_sum_150_180_df, here("data/iucn_least_concern/iucn_lc_rcp85_2050_150_to_180_070926.rds"))
```

# Run threatened species to get percent threatened

# Run full lc set

```{r}
# Fewer species so run full globe
system.time({
 threat_list_count_df <- parallel::mclapply(aphiaid_threat, FUN = least_concern_count, 
                                       df = df_threatened, 
                                       scenario = "RCP85_2050",
                                       xmin = -180,
                                       xmax = 180,
                                       mc.cores = 40)
})
```

```{r}
threat_count_df <- bind_rows(threat_list_count_df)
```

```{r}
# Sum by XY cell, using collapse (20 seconds)
system.time({
threat_sum_df <- threat_count_df %>% 
  fgroup_by(x, y) %>% 
  fcount()
})
```

```{r}
plot_threat_85_2050 <- ggplot() +
  geom_tile(data = threat_sum_df, aes(x = x, y = y,
                                               fill = N,
                                               color = N)) +
  scale_fill_viridis_c(limits = c(1,302), 
                      breaks = c(1, 10, 25, 50, 150, 302),
                      direction = -1,
                      trans = "log", # visually stretches the lower value data
                      guide = guide_colorbar(
                      barwidth  = unit(6, "cm"),   # wide bar for bottom placement
                      barheight = unit(0.4, "cm"),
                      ticks = TRUE,
                      ticks.colour = "white",
                      ticks.linewidth = 0.8
  )) +
  scale_color_viridis_c(limits = c(1,302), 
                        breaks = c(1, 10, 20, 50, 150, 302),
                        direction = -1,
                        trans = "log", # visually stretches the lower value data
                        guide = "none")  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "at-risk species count",
       color = "at-risk species count",
       title = "At-risk species count - RCP 8.5 2050") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt"))
```

```{r}
# Save plot of least concern species
ggsave(here("figures/iucn_atrisk_spp_rcp85_2050_071026.png"), plot_threat_85_2050, width = 10, height = 5, dpi = 300)
```