---
title: "AquaX and IUCN status"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
library(arrow)
```

This document uses the IUCN lookup and threat category data created in the chi repository. It connects IUCN status to AquaX data using the Aphia ID, which is a unique identifier for marine species in the World Register of Marine Species (WoRMS). 

```{r}
iucn_lookup <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/data/iucn_redlist/iucn_worms_aphia_id_complete.csv")
```

```{r}
iucn_new_cat <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/data/iucn_redlist/iucn_species_assessments.csv")
```

```{r}
# Load functions for processing AquaX data
source(here( "common_fxns.R"))
```

We'll use the functions to process the AquaX data for the count of threatened species entering and exiting EEZs. We'll use an equal area projection

```{r}
# Save crs for equal area projection
eq_earth <- "+proj=eqearth"
```

## EEZ 

```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
eez_eq <- st_transform(eez, crs = st_crs(eq_earth))
```

```{r}
ggplot() +
  geom_sf(data = eez_eq, aes()) +
  geom_sf(data = eez_eq |> filter(GEONAME == "Russian Exclusive Economic Zone"), aes(), fill = "red") +
  theme_void() +
  theme(legend.position = "none")
```

```{r}
# Create a blank raster matching AquaX resolution (EEZs should be large enough to just do 0.05 degree raster)
eez_raster_blank <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05)
```

```{r}
eez_raster_geoname <- rasterize(eez, eez_raster_blank, field = "GEONAME")
```

```{r}
eez_raster_mrgid <- rasterize(eez, eez_raster_blank, field = "MRGID")
```


```{r}
# Transform to equal area projection
eez_raster_mrgid_eq <- project(eez_raster_mrgid, eq_earth)
```

# IUCN threatened species

```{r}
iucn_threatened <- iucn_new_cat %>% 
  filter(redlist_category %in% c("CR","EN", "VU"))
```

```{r}
# Merge in the iucn_lookup to get the AphiaID for the threatened species
aphiaid_threatened <- left_join(iucn_threatened, iucn_lookup, by = c("assessment_id" = "iucn_assessment_id")) 
```

```{r}
# what are the NA values - they're the IUCN names w/o an AphiaID match in the lookup table
aphiaid_threatened_na <- aphiaid_threatened %>% filter(is.na(aphia_id))
```

```{r}
# Drop the NA values
aphiaid_threatened_nona <- aphiaid_threatened %>% filter(!is.na(aphia_id))
```

```{r}
# Get aphiaids for the threatened species
aphiaid_threat <- unique(aphiaid_threatened_nona$aphia_id)
```

## Prep AquaX data
We'll be working with the 1,485 thretened aphiaids

```{r}
# read in meta data on the aquax sample of redlist species
load("/home/shares/data-aquax/iucn_request/meta.Rdata")
```

```{r}
meta_clean <- meta %>% 
  mutate(worms_name = tolower(ScientificName)) %>% 
  select(!ScientificName) %>% 
  # simplify for now for easier viewing
  select(AphiaID, SPID, worms_name, GROUP, NR_CR, range_pixels, SDM)
```

```{r}
spp_fs <- list.files('/home/shares/data-aquax/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(meta, by = c('aphiaid' = 'AphiaID'))
```

```{r}
# Filter for the AphiaID of the threatend species
# For example didn't get 1451203 (file doesn't exist)
df_threatened <- df %>% 
  filter(aphiaid %in% aphiaid_threat)
```

```{r}
test_f <- read_parquet("/home/shares/data-aquax/SDM/FINAL_EMSDM_EMMEAN_SP_1017389.parquet")
```

```{r}
set.seed(123)
test_spp_30 <- slice_sample(df_threatened, n = 30)
files_30 <- test_spp_30$f
aphiaid_30 <- test_spp_30$aphiaid
```

```{r}
test_r <- rast(test_f, crs = "EPSG:4326") %>%
    extend(ext(eez_raster_mrgid))
```

```{r}
test_r_eq <- project(test_r, eq_earth)
```

```{r}
# Test plot
plot(test_r_eq$RCP45_2050)
```

```{r}
r_binary45_2050 <- ifel(test_r[["RCP45_2050"]] >= mean(test_f$cutoff, na.rm = TRUE), 1, NA)
```

```{r}
# Have to use the EEZ raster with the numeric IDs (MRGID) for zonal stats
# Try // - took 2 min
system.time({
  eez_list_zonal <- parallel::mclapply(aphiaid_30, FUN = extract_eez_zonal, 
                                       df = test_spp_30, 
                                       eez_raster = eez_raster_mrgid, 
                                       mc.cores = 10)
})
```