---
title: "Aquax - 1 pxl threshold"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
library(arrow)
library(collapse)
```

This document is a supplement to 01_iucn_aquax_eez_overlap.qmd and runs the species entry and exit data for a threshold of 1 pixel instead of 10 pixels.

This document uses the IUCN lookup and threat category data created in the chi repository. It connects IUCN status to AquaX data using the Aphia ID, which is a unique identifier for marine species in the World Register of Marine Species (WoRMS). 

```{r}
iucn_lookup <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/ab9cd09931a4ce5c12d0bcd2984595d6522ed158/_data/iucn_redlist_2026/iucn_worms_aphia_id_complete.csv")
```

```{r}
iucn_new_cat <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/_data/iucn_redlist_2026/iucn_species_assessments.csv")
```

```{r}
# Load functions for processing AquaX data
source(here( "common_fxns.R"))
```

We'll use the functions to process the AquaX data for the count of threatened species entering and exiting EEZs.

## Species count in EEZs
Rounding issue doesn't impact the count of species in each eez so use iucn_threatened_spp_eez_pxls_061626.csv. This file was created using the 01_iucn_aquax_eez_overlap.qmd script and contains the count of pixels for each species in each EEZ for current and future scenarios.

```{r}
# Spp in EEZs
spp_eez_pxl_count <- read_csv(here("data/iucn_threatened_spp_eez_pxls_061626.csv"))
```

```{r}
# Organize dataframe so that's it's easier to select a scenario and compare to current with a threshold of 1 pixel
eez_spp_pxls_long <- spp_eez_pxl_count %>% 
  pivot_longer(cols = starts_with("n_RCP"),
               names_to = "scenario",
               values_to = "pxl_count") %>% 
  mutate(scenario = str_remove(scenario, "n_")) %>% 
  # Replace NA with 0, otherwise will be assigned to "continued presence"
  mutate(pxl_count = replace_na(pxl_count, 0),
         n_current = replace_na(n_current, 0)) %>%
  # Set 1 pixel threshold (at least one pixel needs to be in EEZ for species presence)
  mutate(spp_change = case_when(n_current < 1 &  pxl_count >= 1 ~ 'entry into new',
                            n_current >= 1  & pxl_count < 1 ~ 'exit from historical',
                            n_current < 1  & pxl_count < 1 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

## EEZ 

```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
eez_simple <- eez |>
  select(MRGID, GEONAME, geom)
```

```{r}
eez_ids <- eez %>% 
  st_drop_geometry() %>% 
  dplyr::select(MRGID, GEONAME)
```

```{r}
# Create a blank raster matching AquaX resolution (EEZs should be large enough to just do 0.05 degree raster)
eez_raster_blank <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05,
                  crs = "EPSG:4326")
```

```{r}
eez_raster_geoname <- rasterize(eez_simple, eez_raster_blank, field = "GEONAME")
```

```{r}
eez_raster_mrgid <- rasterize(eez_simple, eez_raster_blank, field = "MRGID")
```

```{r}
# Get GEONAMES of non-contested regions (will run all regions to include the most data)
eez_200nm <-eez %>% 
  filter(POL_TYPE == "200NM")

eez_200nm_list <- eez_200nm$MRGID
```


# IUCN threatened species

```{r}
# Filter the iucn_lookup for species with an SDM
iucn_lookup_sdm <- iucn_lookup %>% filter(aquax == "SDM")
```

```{r}
# Filter iucn categories for threatened species (CR, EN, VU) and near threatened (NT) species
iucn_threatened <- iucn_new_cat %>% 
  filter(redlist_category %in% c("CR","EN", "VU", "NT"))
```

```{r}
# Merge in the iucn_lookup_sdm to get the AphiaID for the threatened species
aphiaid_threatened <- inner_join(iucn_lookup_sdm, iucn_threatened, by = c("iucn_assessment_id" = "assessment_id")) 
```

```{r}
# Get aphiaids for the threatened species
aphiaid_threat <- unique(aphiaid_threatened$aphia_id)
```

## Prep AquaX data
We'll be working with the 1,323 at-risk aphiaids

```{r}
# read in meta data on the aquax species
load("/home/shares/data-aquax/META_02122025.Rdata")
```

```{r}
meta_clean <- META %>% 
  mutate(worms_name = tolower(ScientificName)) %>% 
  select(!ScientificName) %>% 
  # simplify for now for easier viewing
  select(AphiaID, SPID, worms_name, GROUP, NR_CR, range_pixels, SDM)
```

```{r}
spp_fs <- list.files('/home/shares/data-aquax/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(META, by = c('aphiaid' = 'AphiaID'))
```

```{r}
# Filter for the AphiaID of the threatend species - get 1,047 threatened species and 276 non-threatened species with SDMs
df_threatened <- df %>% 
  filter(aphiaid %in% aphiaid_threat)
```

```{r}
test_f <- read_parquet("/home/shares/data-aquax/SDM/FINAL_EMSDM_EMMEAN_SP_1017389.parquet")
```

## Get the pixels of entry and exit 
Update function to fix XY rounding error that created stripes in maps and also update function to only read in necessary data in duckdb for efficiency.

```{r}
# Get counts of entry into new, exit from historical etc. by EEZ
eez_spp_entry <- eez_spp_pxls_long %>% 
  filter(spp_change == "entry into new")
```

```{r}
# Create dataframe of EEZ raster
eez_raster_df <- as.data.frame(eez_raster_mrgid, xy = TRUE) %>% 
  # Round so that the X and Y match the AquaX data (which is rounded to 3 decimal places)
  mutate(x = round(x, 3),
         y = round(y, 3))
```

# Run full set - species entries

```{r}
# Took 34 seconds over 30 cores
system.time({
  entry_rcp85_list <- parallel::mclapply(aphiaid_threat, FUN = extract_eez_entry_pixels_df_db, 
                                       df = df_threatened, 
                                       s = "RCP85_2050",
                                       eez_spp_entry = eez_spp_entry,
                                       eez_raster_df = eez_raster_df,
                                       mc.cores = 30)
})
```

```{r}
entry_rcp85_df <- bind_rows(entry_rcp85_list) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse
system.time({
entry_rcp85_sum_df <- entry_rcp85_df %>% 
  fgroup_by(MRGID, x, y) %>% 
  fcount() %>% 
  mutate(scenario = "RCP85_2050")
})
```

```{r}
# Plot to see if stripes are gone
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8328)) +
  coord_sf(expand = FALSE) +
  geom_tile(data = entry_rcp85_sum_df  %>% filter(MRGID == 8328), aes(x = x, y = y, fill = N, color = N)) +
  #geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5)+
  scale_fill_viridis_c(direction = -1) +
  scale_color_viridis_c(direction = -1) + 
  theme_minimal()
```

```{r}
# Plot to see if stripes are gone (all EEZ regions; including joint regime and contested regions)
ggplot() +
  geom_sf(data = eez) +
  coord_sf(expand = FALSE) +
  geom_tile(data = entry_rcp85_sum_df, aes(x = x, y = y, fill = N, color = N)) +
  scale_fill_viridis_c(limits = c(1,32), 
                      breaks = c(1, 2, 4, 8, 16, 32),
                      direction = -1,
                      trans = "log", # visually stretches the lower value data
                      guide = guide_colorbar(
                      barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                      barheight = unit(0.4, "cm"),
                      ticks = TRUE,
                      ticks.colour = "white",
                      ticks.linewidth = 0.8
  )) +
  scale_color_viridis_c(limits = c(1,32), 
                        breaks = c(1, 2, 4, 8, 16, 32),
                        direction = -1,
                        trans = "log", # visually stretches the lower value data
                        guide = "none")  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "species entry count",
       color = "species entry count",
       x = "",
       y = "") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt")) 
```

```{r}
# Took 34 seconds over 30 cores
system.time({
  entry_rcp45_2050_list <- parallel::mclapply(aphiaid_threat, FUN = extract_eez_entry_pixels_df_db, 
                                       df = df_threatened, 
                                       s = "RCP45_2050",
                                       eez_spp_entry = eez_spp_entry,
                                       eez_raster_df = eez_raster_df,
                                       mc.cores = 30)
})
```

```{r}
entry_rcp45_2050_df <- bind_rows(entry_rcp45_2050_list) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse
system.time({
entry_rcp45_2050_sum_df <- entry_rcp45_2050_df %>% 
  fgroup_by(MRGID, x, y) %>% 
  fcount() %>% 
  mutate(scenario = "RCP45_2050")
})
```

```{r}
# Took 34 seconds over 30 cores
system.time({
  entry_rcp45_2100_list <- parallel::mclapply(aphiaid_threat, FUN = extract_eez_entry_pixels_df_db, 
                                       df = df_threatened, 
                                       s = "RCP45_2100",
                                       eez_spp_entry = eez_spp_entry,
                                       eez_raster_df = eez_raster_df,
                                       mc.cores = 30)
})
```

```{r}
entry_rcp45_2100_df <- bind_rows(entry_rcp45_2100_list) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse
system.time({
entry_rcp45_2100_sum_df <- entry_rcp45_2100_df %>% 
  fgroup_by(MRGID, x, y) %>% 
  fcount() %>% 
  mutate(scenario = "RCP45_2100")
})
```

