---
title: "IUCN AquaX Range Area"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
library(arrow)
```

This document will process the range area for threatened species in the AquaX dataset. This will give us a sense for how many species are experiencing range contractions and expansions. 

I'll also test out using duckdb to only read in the scenario of interest.

## Prep data

```{r}
iucn_lookup <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/_data_raw/iucn_redlist_2026/iucn_worms_aphia_id_complete.csv")
```

```{r}
iucn_new_cat <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/_data_raw/iucn_redlist_2026/iucn_species_assessments.csv")
```

```{r}
# Load functions for processing AquaX data
source(here( "common_fxns.R"))
```

## Prep threatened species data

```{r}
# Filter the iucn_lookup for species with an SDM
iucn_lookup_sdm <- iucn_lookup %>% filter(aquax == "SDM")
```

```{r}
# Filter iucn categories for threatened species (CR, EN, VU) and near-threatened (NT)
iucn_threatened <- iucn_new_cat %>% 
  filter(redlist_category %in% c("CR","EN", "VU", "NT"))
```

```{r}
# Merge in the iucn_lookup_sdm to get the AphiaID for the threatened species
aphiaid_threatened <- inner_join(iucn_lookup_sdm, iucn_threatened, by = c("iucn_assessment_id" = "assessment_id")) 
```

```{r}
# Get aphiaids for the threatened species
aphiaid_threat <- unique(aphiaid_threatened$aphia_id)
```


## Prep SDM data for area calculations

```{r}
# read in meta data on the aquax sample of redlist species
load("/home/shares/data-aquax/META_02122025.Rdata")
```

```{r}
meta_clean <- META %>% 
  mutate(worms_name = tolower(ScientificName)) %>% 
  select(!ScientificName) %>% 
  # simplify for now for easier viewing
  select(AphiaID, SPID, worms_name, GROUP, NR_CR, range_pixels, SDM)
```

```{r}
spp_fs <- list.files('/home/shares/data-aquax/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(META, by = c('aphiaid' = 'AphiaID'))
```

```{r}
# Filter for the AphiaID of the threatend species 
df_threatened <- df %>% 
  filter(aphiaid %in% aphiaid_threat)
```

## Run sample of species through area calculations
```{r}
set.seed(123)
test_spp_30 <- slice_sample(df, n = 100)
files_30 <- test_spp_30$f
aphiaid_30 <- test_spp_30$aphiaid
```

```{r}
system.time({
  eez_list_area <- parallel::mclapply(aphiaid_30, FUN = calculate_area, 
                                       df = test_spp_30, 
                                       scenario = "RCP85_2100",
                                       unit = "km", 
                                       mc.cores = 20)
})
```

```{r}
system.time({
  eez_list_area_current <- parallel::mclapply(aphiaid_30, FUN = calculate_area, 
                                       df = test_spp_30, 
                                       scenario = "Current",
                                       unit = "km", 
                                       mc.cores = 20)
})
```

```{r}
area_df <- bind_rows(c(eez_list_area_current, eez_list_area)) %>% 
    pivot_wider(id_cols = aphiaid, names_from = s, values_from = area, names_glue = "{s}_area_km2") %>% 
    mutate(area_change = case_when(
        Current_area_km2 > RCP85_2100_area_km2 & RCP85_2100_area_km2 != 0 ~ "contraction",
        Current_area_km2 < RCP85_2100_area_km2 ~ "expansion",
        Current_area_km2 > RCP85_2100_area_km2 & RCP85_2100_area_km2 == 0 ~ "spp range loss",
        TRUE ~ "no change"
    ))
```

## Run full set of threatened species through area calculations

Each scenario is run separately and takes about 30 seconds each to run.

```{r}
files <- df_threatened$f
aphiaid <- df_threatened$aphiaid
```

```{r}
# 45 seconds
system.time({
  area_current <- parallel::mclapply(aphiaid, FUN = calculate_area, 
                                       df = df_threatened, 
                                       scenario = "Current",
                                       unit = "km", 
                                       mc.cores = 20)
})
```

```{r}
# 30 seconds
system.time({
  area_rcp45_2050 <- parallel::mclapply(aphiaid, FUN = calculate_area, 
                                       df = df_threatened, 
                                       scenario = "RCP45_2050",
                                       unit = "km", 
                                       mc.cores = 20)
})
```

```{r}
# 30 seconds
system.time({
  area_rcp85_2050 <- parallel::mclapply(aphiaid, FUN = calculate_area, 
                                       df = df_threatened, 
                                       scenario = "RCP85_2050",
                                       unit = "km", 
                                       mc.cores = 20)
})
```

```{r}
# 30 seconds
system.time({
  area_rcp45_2100 <- parallel::mclapply(aphiaid, FUN = calculate_area, 
                                       df = df_threatened, 
                                       scenario = "RCP45_2100",
                                       unit = "km", 
                                       mc.cores = 20)
})
```

```{r}
# 30 seconds
system.time({
  area_rcp85_2100 <- parallel::mclapply(aphiaid, FUN = calculate_area, 
                                       df = df_threatened, 
                                       scenario = "RCP85_2100",
                                       unit = "km", 
                                       mc.cores = 20)
})
```

## Format data

```{r}
area_45_2050_df <- bind_rows(c(area_current, area_rcp45_2050)) %>% 
    pivot_wider(id_cols = aphiaid, names_from = s, values_from = area, names_glue = "{s}_area_km2") %>% 
    mutate(area_change = case_when(
        Current_area_km2 > RCP45_2050_area_km2 & RCP45_2050_area_km2 != 0 ~ "contraction",
        Current_area_km2 < RCP45_2050_area_km2 ~ "expansion",
        Current_area_km2 > RCP45_2050_area_km2 & RCP45_2050_area_km2 == 0 ~ "spp range loss",
        TRUE ~ "no change"
    ))
```

```{r}
area_85_2050_df <- bind_rows(c(area_current, area_rcp85_2050)) %>% 
    pivot_wider(id_cols = aphiaid, names_from = s, values_from = area, names_glue = "{s}_area_km2") %>% 
    mutate(area_change = case_when(
        Current_area_km2 > RCP85_2050_area_km2 & RCP85_2050_area_km2 != 0 ~ "contraction",
        Current_area_km2 < RCP85_2050_area_km2 ~ "expansion",
        Current_area_km2 > RCP85_2050_area_km2 & RCP85_2050_area_km2 == 0 ~ "spp range loss",
        TRUE ~ "no change"
    ))
```

```{r}
area_45_2100_df <- bind_rows(c(area_current, area_rcp45_2100)) %>% 
    pivot_wider(id_cols = aphiaid, names_from = s, values_from = area, names_glue = "{s}_area_km2") %>% 
    mutate(area_change = case_when(
        Current_area_km2 > RCP45_2100_area_km2 & RCP45_2100_area_km2 != 0 ~ "contraction",
        Current_area_km2 < RCP45_2100_area_km2 ~ "expansion",
        Current_area_km2 > RCP45_2100_area_km2 & RCP45_2100_area_km2 == 0 ~ "spp range loss",
        TRUE ~ "no change"
    ))
```

```{r}
area_85_2100_df <- bind_rows(c(area_current, area_rcp85_2100)) %>% 
    pivot_wider(id_cols = aphiaid, names_from = s, values_from = area, names_glue = "{s}_area_km2") %>% 
    mutate(area_change = case_when(
        Current_area_km2 > RCP85_2100_area_km2 & RCP85_2100_area_km2 != 0 ~ "contraction",
        Current_area_km2 < RCP85_2100_area_km2 ~ "expansion",
        Current_area_km2 > RCP85_2100_area_km2 & RCP85_2100_area_km2 == 0 ~ "spp range loss",
        TRUE ~ "no change"
    ))
```

## Save data in a different format

```{r}
area_current_df <- bind_rows(area_current) %>% 
  dplyr::select(aphiaid, area) %>% 
  rename(current_area_km2 = area)

area_scenarios_df <- bind_rows(c(area_rcp45_2050, area_rcp45_2100, area_rcp85_2050, area_rcp85_2100)) %>% 
  dplyr::select(aphiaid, s, area) %>%
  rename(scenario = s, scenario_area_km2 = area)
```

```{r}
# Merge current and scenario area dataframes
area_all_df <- left_join(area_current_df, area_scenarios_df) %>% 
  mutate(area_change = case_when(
    current_area_km2 > scenario_area_km2 & scenario_area_km2 != 0 ~ "contraction",
    current_area_km2 < scenario_area_km2 ~ "expansion",
    current_area_km2 > scenario_area_km2 & scenario_area_km2 == 0 ~ "spp range loss",
    TRUE ~ "no change"
  ))
```

```{r}
# Save data
write_csv(area_all_df, here("data/iucn_threatened_spp_range_area_061826.csv"))
```

## Summary table

```{r}
area_summary <- area_all_df %>% 
  group_by(scenario, area_change) %>% 
  summarise(n_spp = n(),
  perc_spp = n()/length(unique(area_all_df$aphiaid))* 100)
```

## Summary graph

```{r}
ggplot(area_summary, aes(x = scenario, y = perc_spp, fill = area_change)) +
  geom_col(position = "stack") +
  labs(x = "Scenario", y = "Percentage of species", fill = "Change",
  title = "Threatened species range area change (n = 1,323)") +
  scale_fill_viridis_d() + 
  theme_minimal()
```