---
title: "Plots of Threatened Species Ranges & EEZs"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
library(arrow)
library(patchwork)
library(cowplot)
```

This document creates plots for the data calculated in 01_iucn_aquax_eez_overlap.qmd and 02_iucn_aquax_rangearea.qmd.

## Read in threatened species data

```{r}
# Spp in EEZs
spp_eez_pxl_count <- read_csv(here("data/iucn_threatened_spp_eez_pxls_061626.csv"))
```

```{r}
# Spp EEZ entry/exit pixels
spp_entry_pxls <- read_rds(here("data/iucn_threatened_10pxlthres_entry_pxls_061726.rds"))
spp_exit_pxls <- read_rds(here("data/iucn_threatened_10pxlthres_exit_pxls_061726.rds"))
```

```{r}
# Species range area
area_df <- read_csv(here("data/iucn_threatened_spp_range_area_061826.csv"))
```

## Read in EEZ shapefile
```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')

eez_simple <- eez %>% 
  dplyr::select(MRGID, GEONAME)
```

```{r}
# Create a blank raster matching AquaX resolution (EEZs should be large enough to just do 0.05 degree raster)
eez_raster_blank <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05)
```

```{r}
eez_raster_geoname <- rasterize(eez, eez_raster_blank, field = "GEONAME")
```

```{r}
eez_joint <- eez %>% 
  filter(POL_TYPE == "Joint regime")
```

```{r}
ggplot() +
  geom_sf(data = eez_no_overlap) +
  geom_sf(data = (eez_no_overlap %>% filter(MRGID== 8487)), fill = "blue", alpha = 0.5) +
  geom_sf(data = eez_joint, color = "red", alpha = 0.5) 
```

## Identify overlapping claim and joint regime EEZs

```{r}
overlap_claim <- eez %>% 
  filter(POL_TYPE != "200NM")

overlap_claim_list <- unique(overlap_claim$MRGID)
```

```{r}
# eez no overlapping claim
eez_no_overlap <- eez %>% 
  filter(!MRGID %in% overlap_claim_list)

eez_no_overlap_df <- eez_no_overlap %>% 
  st_drop_geometry()
```

## Plot species pixels

```{r}
spp_entry_pxls_85_2050 <- spp_entry_pxls %>% 
  filter(scenario == "RCP85_2050") %>% 
  filter(!MRGID %in% overlap_claim_list)

spp_exit_pxls_85_2050 <- spp_exit_pxls %>% 
  filter(scenario == "RCP85_2050") %>% 
  filter(!MRGID %in% overlap_claim_list)
```

```{r}
# Plot EEZ entry pixels
plot_entry_10_map <- ggplot() +
  geom_sf(data = eez_no_overlap) +
  coord_sf(expand = FALSE) +
  geom_tile(data = spp_entry_pxls_85_2050, aes(x = x, y = y,
                                               fill = n_entry_spp,
                                               color = n_entry_spp)) +
  scale_fill_viridis_c(limits = c(1,28), 
                      breaks = c(1, 2, 4, 8, 16, 28),
                      direction = -1,
                      trans = "log", # visually stretches the lower value data
                      guide = guide_colorbar(
                      barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                      barheight = unit(0.4, "cm"),
                      ticks = TRUE,
                      ticks.colour = "white",
                      ticks.linewidth = 0.8
  )) +
  scale_color_viridis_c(limits = c(1,28), 
                        breaks = c(1, 2, 4, 8, 16, 28),
                        direction = -1,
                        trans = "log", # visually stretches the lower value data
                        guide = "none")  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "species entry count",
       color = "species entry count",
       x = "",
       y = "") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt")) 
```

```{r}
plot_entry_10_density <- ggplot(spp_entry_pxls_85_2050, aes(x = y)) +
  geom_density(fill = "forestgreen", alpha = 0.5) +
  coord_flip() +
  scale_x_continuous(limits = c(-90, 90)) +
  theme_minimal() +
  labs(x = "",
       y = "") +
  theme(plot.margin = margin(0, 0, 0, 0))
```

```{r}
entry_combined_plot <- plot_entry_10_map + plot_entry_10_density + 
  plot_annotation(title = "Species entry, RCP 8.5 2050") +
  plot_layout(widths = c(4, 1),
              axis_titles = "collect",
              guides = "collect") &
  theme(legend.position = "bottom") 
```

```{r}
# Plot EEZ exit pixels
plot_exit_10_map <- ggplot() +
  geom_sf(data = eez_no_overlap) +
  coord_sf(expand = FALSE) +
  geom_tile(data = spp_exit_pxls_85_2050, aes(x = x, y = y,
                                              fill = n_exit_spp,
                                              color = n_exit_spp)) +
  scale_fill_viridis_c(limits = c(1,28), 
                       breaks = c(1, 2, 4, 8, 16, 28),
                       direction = -1, 
                       trans = "log", # visually stretches the lower value data
                       guide = guide_colorbar(
                       barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                       barheight = unit(0.4, "cm"),
                       ticks = TRUE,
                       ticks.colour = "white",
                       ticks.linewidth = 0.8)) +
  scale_color_viridis_c(limits = c(1,28), 
                        breaks = c(1, 2, 4, 8, 16, 28), 
                        direction = -1,
                        trans = "log", # visually stretches the lower value data
                        guide = "none")  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "species exit count",
       color = "species exit count",
       x = "",
       y = "") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt"))
```

```{r}
plot_exit_10_density <- ggplot(spp_exit_pxls_85_2050, aes(x = y)) +
  geom_density(fill = "darkmagenta", alpha = 0.5) +
  coord_flip() +
  scale_x_continuous(limits = c(-90, 90)) +
  theme_minimal() +
  labs(x = "",
       y = "") +
  theme(plot.margin = margin(0, 0, 0, 0))
```

```{r}
exit_combined_plot <- plot_exit_10_map + plot_exit_10_density + 
  plot_annotation(title = "Species exit, RCP 8.5 2050") +
  plot_layout(widths = c(4, 1),
              axis_titles = "collect",
              guides = "collect") &
  theme(legend.position = "bottom") 
```

```{r}
# Save entry and exit maps
ggsave(here("figures/iucn_threatened_spp_entry_10pxl_map_062526.png"), entry_combined_plot, width = 10, height = 5, dpi = 300)
ggsave(here("figures/iucn_threatened_spp_exit_10pxl_map_062526.png"), exit_combined_plot, width = 10, height = 5, dpi = 300)
```

## Plot some areas with high numbers of entries/exits
```{r}
ggplot() +
  geom_sf(data = eez_no_overlap %>% filter(MRGID %in% c("8328"))) +
  geom_tile(data = spp_entry_pxls_85_2050 %>% 
            filter(MRGID %in% c("8328")), 
            aes(x = x, y = y,
            fill = n_entry_spp,
            color = n_entry_spp)) +
  scale_fill_viridis_c(limits = c(1,32), 
                       breaks = c(1, 2, 4, 8, 16, 32),
                       direction = -1, 
                       trans = "log", # visually stretches the lower value data
                       guide = guide_colorbar(
                       barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                       barheight = unit(0.4, "cm"),
                       ticks = TRUE,
                       ticks.colour = "white",
                       ticks.linewidth = 0.8)) +
  scale_color_viridis_c(limits = c(1,32), 
                        breaks = c(1, 2, 4, 8, 16, 32),
                        direction = -1,
                        trans = "log", # visually stretches the lower value data
                        guide = "none") +
  theme_minimal() +
  labs(title = "EEZs with most threatened species entries under RCP8.5 2050",
       subtitle = "Top EEZs: North Korea") +
  theme(legend.position = "bottom")
```


```{r}
ggplot() +
  geom_sf(data = eez_no_overlap %>% 
  filter(stringr::str_detect(GEONAME, "Australia"))) +
  geom_tile(data = spp_exit_pxls_85_2050 %>% 
            filter(stringr::str_detect(GEONAME, "Australia")), 
            aes(x = x, y = y,
            fill = n_exit_spp,
            color = n_exit_spp)) +
  scale_fill_viridis_c(limits = c(1,32), 
                       breaks = c(1, 2, 4, 8, 16, 32),
                       direction = -1, 
                       trans = "log", # visually stretches the lower value data
                       guide = guide_colorbar(
                       barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                       barheight = unit(0.4, "cm"),
                       ticks = TRUE,
                       ticks.colour = "white",
                       ticks.linewidth = 0.8)) +
  scale_color_viridis_c(limits = c(1,32), 
                        breaks = c(1, 2, 4, 8, 16, 32),
                        direction = -1,
                        trans = "log",
                        guide = "none") +
  theme_minimal() +
  theme(legend.position = "bottom")+
  labs(title = "EEZs with most threatened species exits under RCP8.5 2050",
       subtitle = "Austrailian EEZ and Joint Regime Area")
```

```{r}
ggplot() +
  geom_sf(data = eez_no_overlap %>% 
  filter(stringr::str_detect(GEONAME, "United States"))) +
  geom_tile(data = spp_entry_pxls_85_2050 %>% 
            filter(stringr::str_detect(GEONAME, "United States")), 
            aes(x = x, y = y,
            fill = n_entry_spp,
            color = n_entry_spp)) +
  scale_fill_viridis_c(limits = c(1,32), 
                       breaks = c(1, 2, 4, 8, 16, 32),
                       direction = -1, 
                       trans = "log", # visually stretches the lower value data
                       guide = guide_colorbar(
                       barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                       barheight = unit(0.4, "cm"),
                       ticks = TRUE,
                       ticks.colour = "white",
                       ticks.linewidth = 0.8)) +
  scale_color_viridis_c(limits = c(1,32), 
                        breaks = c(1, 2, 4, 8, 16, 32),
                        direction = -1,
                        trans = "log",
                        guide = "none") +
  theme_minimal() +
  theme(legend.position = "bottom")+
  labs(title = "EEZs with most threatened species entries under RCP8.5 2050",
       subtitle = "United States EEZ and Joint Regime Area")
```

## Spatial percent change in threatened species by pixel
```{r}
# Read in RDS for RCP 85 2050 for continued presence pixels
spp_cp_pxls <- readRDS(here("data/iucn_threatened_10pxlthres_rcp85_2050_cp_pxls_061826.rds"))
```

```{r}
eez_sum_cp_exit_85_2050 <- full_join(spp_cp_pxls, (spp_exit_pxls %>% filter(scenario == "RCP85_2050")))
eez_sum_all_85_2050 <- full_join(eez_sum_cp_exit_85_2050, (spp_entry_pxls %>% filter(scenario == "RCP85_2050"))) %>% 
  # Replace NA with 0
  #replace_na(list(n_cp_spp = 0, n_exit_spp = 0, n_entry_spp = 0)) %>% 
  # Drop overlapping claim EEZs
  filter(!MRGID %in% overlap_claim_list) %>%
  mutate(n_current_spp = sum(n_cp_spp, n_exit_spp, na.rm = TRUE),
         n_future_spp = sum(n_cp_spp, n_entry_spp, na.rm = TRUE),
         # percent of current species that exit
         perc_exit_spp = 100 * n_exit_spp/(n_cp_spp + n_exit_spp),
         # percent of future species that entered
         perc_entry_spp = 100 * n_entry_spp/(n_cp_spp + n_entry_spp))
```

```{r}
plot_exit_percent <- ggplot() +
  geom_sf(data = eez_no_overlap) +
  coord_sf(expand = FALSE) +
  geom_tile(data = (eez_sum_all_85_2050 %>% filter(!is.na(perc_exit_spp))), aes(x = x, y = y,
                                               fill = perc_exit_spp,
                                               color = perc_exit_spp)) +
  scale_fill_viridis_c(limits = c(1,100), 
                       breaks = c(1, 25, 50, 75, 100),
                       direction = -1,
                       guide = guide_colorbar(
                       barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                       barheight = unit(0.4, "cm"),
                       ticks = TRUE,
                       ticks.colour = "white",
                       ticks.linewidth = 0.8)) +
  scale_color_viridis_c(limits = c(1,100), direction = -1,
                        breaks = c(1, 25, 50, 75, 100),
                        guide = "none")  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "percent species lost",
       color = "percent species lost",
       title = "Species exits by percent, RCP 8.5 2050") +
  theme(legend.position = "bottom")
```

```{r}
# Save plot
ggsave(here("figures/iucn_threatened_spp_exit_perc_10pxl_map_062326.png"), plot_exit_percent, width = 7, height = 5, dpi = 300)
```

## Other plots 

```{r}
# How many EEZs are each spp present in for current and future?
eez_spp_counts <- spp_eez_pxl_count %>%  
  # Drop overlapping claim EEZs
  filter(!MRGID %in% overlap_claim_list) %>%
  group_by(aphiaid) %>% 
  summarize(current_eez_count = sum(n_current >= 10, na.rm = TRUE),
            RCP45_2050_eez_count = sum(n_RCP45_2050 >= 10, na.rm = TRUE),
            RCP45_2100_eez_count = sum(n_RCP45_2100  >= 10, na.rm = TRUE),
            RCP85_2050_eez_count = sum(n_RCP85_2050  >= 10, na.rm = TRUE),
            RCP85_2100_eez_count = sum(n_RCP85_2100  >= 10, na.rm = TRUE)) %>% 
  ungroup() 
```

```{r}
ggplot() +
  geom_point(data = eez_spp_counts, aes(x = current_eez_count, y = RCP85_2050_eez_count),
             alpha = 0.5, color = "maroon") +
  #geom_point(data = eez_spp_counts, aes(x = current_eez_count, y = RCP85_2100_eez_count),
             #alpha = 0.5, color = "forestgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  labs(title = "IUCN Threatened Species EEZ count (RCP 8.5 2050)",
       y = "RCP 85 future EEZ count")
```


```{r}
# Organize dataframe so that's it's easier to select a scenario and compare to current
eez_spp_pxls_long <- spp_eez_pxl_count %>% 
  # Drop overlapping claim EEZs
  filter(!MRGID %in% overlap_claim_list) %>%
  pivot_longer(cols = starts_with("n_RCP"),
               names_to = "scenario",
               values_to = "pxl_count") %>% 
  mutate(scenario = str_remove(scenario, "n_")) %>% 
  # Replace NA with 0, otherwise will be assigned to "continued presence"
  replace_na(list(pxl_count = 0, n_current = 0)) %>% 
  mutate(spp_change = case_when(n_current < 10 &  pxl_count >= 10 ~ 'entry into new',
                            n_current >= 10  & pxl_count < 10 ~ 'exit from historical',
                            n_current < 10  & pxl_count < 10 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

```{r}
# Get counts of entry into new, exit from historical etc. by EEZ
eez_spp_change <- eez_spp_pxls_long %>% 
  group_by(GEONAME, scenario, spp_change) %>% 
  summarize(spp_change_count = n()) %>% 
  ungroup()
```

```{r}
eez_c_f <- eez_spp_change %>% 
  mutate(spp_change = str_replace_all(spp_change, " ", "_")) %>% 
  pivot_wider(names_from = spp_change, values_from = spp_change_count) %>% 
  replace_na(list(continued_presence = 0, exit_from_historical = 0,
                  entry_into_new = 0, no_presence = 0)) %>% 
  mutate(current_spp = continued_presence + exit_from_historical,
         future_spp = continued_presence + entry_into_new,
         perc_exit = 100 * exit_from_historical/current_spp,
         perc_enter = 100 * entry_into_new/future_spp,
         spp_change = future_spp - current_spp,
         ssi = (2 * continued_presence)/(current_spp + future_spp)) 
```

```{r}
# EEZ centroids
eez_simple_valid <- st_make_valid(eez_simple)
eez_centroid <- st_centroid(eez_simple_valid)  %>%
  cbind(st_coordinates(.)) %>%
  st_drop_geometry()
```

```{r}
# Merge in centroids
eez_c_f_y <- left_join(eez_c_f, eez_centroid)
```


```{r}
plot_spp_count <- ggplot() +
  geom_point(data = eez_c_f_y %>% filter(scenario == "RCP85_2050"),
   aes(fill = Y, x = current_spp, y = future_spp),
             alpha = 0.75, shape= 21, color = "gray50", size = 1.5, 
             position = position_jitterdodge()) +
  #scale_color_viridis_c() +
  #scale_color_distiller(palette = "PRGn", direction = 1, limits = c(-80, 80)) + 
  scale_fill_distiller(palette = "PRGn", direction = 1, limits = c(-80, 80),
                      guide = guide_colorbar(
                       barwidth  = unit(0.4, "cm"), 
                       barheight = unit(5, "cm"))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(fill = "latitude",
       title = "Current and future spp by EEZ, RCP 8.5 2050",
       y = "future species count",
       x = "current species count") +
  theme_bw() 
```

```{r}
ggplot() +
  geom_point(data = eez_c_f_y %>% filter(scenario == "RCP85_2050"),
             aes(color = Y, x = exit_from_historical, y = entry_into_new), fill = "gray30", alpha = 0.8, 
             position = position_jitterdodge()) +
  #scale_color_viridis_c() +
  scale_color_distiller(palette = "PRGn", direction = 1, limits = c(-80, 80)) + 
  scale_fill_distiller(palette = "PRGn", direction = 1, limits = c(-80, 80)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(#color = "latitude",
       title = "Threatened species loss and gain by EEZ",
       y = "species gain count",
       x = "species loss count",
       color = "latitude") +
  theme_bw() +
  facet_wrap(~scenario)

ggplot() +
  geom_point(data = eez_c_f_y %>% filter(scenario == "RCP85_2050"),
             aes(fill = Y, x = exit_from_historical, y = entry_into_new), alpha = 0.8, shape= 21, color = "gray50", size = 2,
             position = position_jitterdodge()) +
  #scale_color_distiller(palette = "PRGn", direction = 1, limits = c(-80, 80)) + 
  scale_fill_distiller(palette = "PRGn", direction = 1, limits = c(-80, 80),
                      guide = guide_colorbar(
                       barwidth  = unit(0.4, "cm"), 
                       barheight = unit(5, "cm"))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(title = "Threatened species count by EEZ region, RCP 8.5 2050",
       y = "species gain",
       x = "species loss",
       fill = "latitude") +
  scale_y_continuous(limits = c(0, 50)) +
  scale_x_continuous(limits = c(0, 50)) +
  theme_bw() 
```

```{r}
# The four rows removed in plotting don't have current or future species
plot_spp_perc <- ggplot() +
  geom_point(data = eez_c_f_y %>% filter(scenario == "RCP85_2050"),
             aes(fill = Y, x = perc_exit, y = perc_enter), alpha = 0.8, shape= 21, color = "gray50", size = 2) +
  #scale_color_distiller(palette = "PRGn", direction = 1, limits = c(-80, 80)) + 
  scale_fill_distiller(palette = "PRGn", direction = 1, limits = c(-80, 80),
                      guide = guide_colorbar(
                       barwidth  = unit(0.4, "cm"), 
                       barheight = unit(5, "cm"))) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(title = "Threatened species percent change by EEZ region, RCP 8.5 2050",
       y = "species gain %",
       x = "species loss %",
       fill = "latitude") +
  scale_y_continuous(limits = c(0, 35)) +
  scale_x_continuous(limits = c(0, 35)) +
  theme_bw() 
```

```{r}
# Add sf 
eez_c_f_sf <- left_join(eez_c_f, eez_simple)
```

```{r}
# Save plots
ggsave(here("figures/iucn_threatened_spp_count_plot_062326.png"), plot_spp_count, width = 7, height = 5, dpi = 300)
ggsave(here("figures/iucn_threatened_spp_perc_plot_062326.png"), plot_spp_perc, width = 7, height = 5, dpi = 300)
```

# Spatial plots of threatened species change by EEZ
```{r}
# Threatened species change
ggplot() +
  geom_sf(data = eez_c_f_sf, aes(geometry = geom, fill = spp_change)) +
  scale_color_distiller(palette = "BrBG", direction = 1, limits = c(-75, 75)) + 
  scale_fill_distiller(palette = "BrBG", direction = 1, limits = c(-75, 75)) +
  theme_bw() +
  labs(title = "Threatened species change by EEZ") +
  facet_wrap(~scenario)
```

```{r}
# Threatened species percentage
ggplot() +
  geom_sf(data = eez_c_f_sf, aes(geometry = geom, fill = perc_exit)) +
  theme_bw() +
  labs(title = "Threatened species change by EEZ") +
  facet_wrap(~scenario)
```

## Biplot
Code created with assistance from Claude to categorize the raster entry and exit pixels into a bivariate color scheme using the following categories to create a 3x3 grid:
* 0 species entry and 0 species exit
* 0 species entry and 1 species exit
* 1 species entry and 0 species exit
* 1 species entry and 1 species exit
* 2+ species entry and 2+ species exit

```{r}
# Merge the entry and exit dataframes so that there are the same number of pixels in each raster
spp_entry_exit_85_2050 <- full_join(spp_entry_pxls_85_2050, spp_exit_pxls_85_2050) %>% 
  # Replace NA with 0
  replace_na(list(n_entry_spp = 0, n_exit_spp = 0))
```

```{r}
# Create a new column that categorizes the number of species entry and exit into 3 categories: 0, 1, and 2+
spp_entry_exit_cat_85_2050 <- spp_entry_exit_85_2050 %>% 
  # Assign numeric_category
  mutate(entry_cat = case_when(n_entry_spp == 0 ~ 1,
                               n_entry_spp == 1 ~ 2,
                               n_entry_spp >= 2 ~ 3),
         exit_cat = case_when(n_exit_spp == 0 ~ 1,
                              n_exit_spp == 1 ~ 2,
                              n_exit_spp >= 2 ~ 3)) %>% 
  # Create a new column that combines the entry and exit categories into a single category
  mutate(entry_exit_cat = (exit_cat - 1) * 3 + entry_cat)
```

```{r}
# Define numeric categories for the bivariate color scheme
bivar_levels <- data.frame(
  entry_exit_cat    = 1:9,
  label = c(
    "En0_Ex0", "En1_Ex0", "En2_Ex0",   # exits = 0
    "En0_Ex1", "En1_Ex1", "En2_Ex1",   # exits = 1
    "En0_Ex2", "En1_Ex2", "En2_Ex2"    # exits = 2+
  )
)
spp_entry_exit_cat_85_2050 <- left_join(spp_entry_exit_cat_85_2050, bivar_levels)
```

```{r}
# Color palette
# Entries increase left → right (blue axis)
# Exits increase bottom → top (pink/purple axis)
bivar_colors <- c(
  "En0_Ex0" = "#e8e8e8",  # entries=0,   exits=0
  "En1_Ex0" = "#ace4e4",  # entries=1, exits=0
  "En2_Ex0" = "#5ac8c8",  # entries=2+,  exits=0
  "En0_Ex1" = "#dfb0d6",  # entries=0,   exits=1
  "En1_Ex1" = "#a5add3",  # entries=1, exits=1
  "En2_Ex1" = "#5698b9",  # entries=2+,  exits=1
  "En0_Ex2" = "#be64ac",  # entries=0,   exits=2+
  "En1_Ex2" = "#8c62aa",  # entries=1, exits=2+
  "En2_Ex2" = "#3b4994"   # entries=2+,  exits=2+
)
```

```{r}
biplot <- ggplot() +
  geom_sf(data = eez_no_overlap, fill = "#e8e8e8") +
  coord_sf(expand = FALSE) +
  geom_tile(data = spp_entry_exit_cat_85_2050, aes(x = x, y = y,
                                              fill = label,
                                              color = label)) +
  scale_fill_manual(values = bivar_colors, na.value = "white") +
  scale_color_manual(values = bivar_colors, na.value = "white") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Species exits vs entries, RCP 8.5 2050",
       x = "",
       y = "")
```

```{r}
# Create legend for bivariate color scheme
legend_df <- expand.grid(entries = 1:3, exits = 1:3)
legend_df$label <- paste0("En", legend_df$entries - 1, "_Ex", legend_df$exits - 1)

legend_biplot <- ggplot(legend_df, aes(x = entries, y = exits, fill = label)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_manual(values = bivar_colors) +
  scale_x_continuous(breaks = 1:3, labels = c("0", "1", "2+")) +
  scale_y_continuous(breaks = 1:3, labels = c("0", "1", "2+")) +
  labs(x = "Entries →", y = "Exits →") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text       = element_text(size = 8),
    axis.title      = element_text(size = 9)
  )
```

```{r}
# Combine plot
bivariate_map <- ggdraw() +
  draw_plot(biplot,    0,    0,    1,    1) +
  draw_plot(legend_biplot, 0.75, 0.75, 0.2, 0.25)
```

```{r}
# Save entry and exit maps
ggsave(here("figures/iucn_threatened_spp_entry_exit_10pxl_map_062426.png"), bivariate_map, width = 7, height = 5, dpi = 300)
```

# Add density plot to bivariate map

```{r}
biplot_density <-ggplot() +
  geom_density(data = spp_exit_pxls_85_2050, aes(x = y),fill = "#be64ac", color = "#be64ac", alpha = 0.5, size = 1) +
  geom_density(data = spp_entry_pxls_85_2050, aes(x = y),fill = "#5698b9", color = "#5698b9", alpha = 0.5, size = 1) +
  coord_flip() +
  scale_x_continuous(limits = c(-90, 90)) +
  theme_minimal() +
  labs(x = "",
       y = "") +
  theme(plot.margin = margin(0, 0, 0, 0))
```

```{r}
biplot_2 <- ggplot() +
  geom_sf(data = eez_no_overlap, fill = "#e8e8e8") +
  coord_sf(expand = FALSE) +
  geom_tile(data = spp_entry_exit_cat_85_2050, aes(x = x, y = y,
                                              fill = label,
                                              color = label)) +
  scale_fill_manual(values = bivar_colors, na.value = "white") +
  scale_color_manual(values = bivar_colors, na.value = "white") +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt")) +
  labs(title = "",
       x = "",
       y = "") 
```

```{r}
biplot_and_density <- biplot_2 + biplot_density + 
  plot_annotation(title = "Species exit vs entry, RCP 8.5 2050") +
  plot_layout(widths = c(4, 1),
              axis_titles = "collect",
              guides = "collect") &
  theme(legend.position = "none") 
```

```{r}
biplot_all <- ggdraw() +
  draw_plot(biplot_and_density,    0,    0,    1,    1) +
  draw_plot(legend_biplot, 0.6, 0.8, 0.2, 0.2)
```

```{r}
# Save entry and exit maps
ggsave(here("figures/iucn_threatened_spp_entry_exit_density_10pxl_map_062426.png"), biplot_all, width = 10, height = 5, dpi = 300)
```

## Test biplot

```{r}
# Merge the entry and exit dataframes
entry_exit_spp_count_pxl <- full_join(spp_entry_pxls, spp_exit_pxls) 
```

```{r}
entry_exit_spp_sqrt <- entry_exit_spp_count_pxl %>% 
    filter(scenario == "RCP85_2100") %>%
  replace_na(list(n_entry_spp = 0)) %>% 
  replace_na(list(n_exit_spp = 0)) %>% 
  mutate(n_entry_spp_sqrt = sqrt(n_entry_spp),
         n_exit_spp_sqrt = sqrt(n_exit_spp))
```


```{r}
library(biscale)
library(cowplot)
```

```{r}
biplot_data <- bi_class(entry_exit_spp_sqrt, x = n_entry_spp_sqrt, y = n_exit_spp_sqrt,
                        style = "equal", dim = 3)
```

```{r}
biplot_85_2100 <- ggplot() + 
  geom_tile(data = biplot_data, mapping = aes(fill = bi_class,
                                              color = bi_class,
                                              x = x, y = y), show.legend = FALSE) +
  bi_scale_fill(pal = "DkBlue2", dim = 3, rotate_pal = FALSE) +
  bi_scale_color(pal = "DkBlue2", dim = 3, rotate_pal = FALSE) +
  geom_sf(data = eez, fill = NA) +
  labs(x = "",
       y = "",
       subtitle = "RCP 8.5 2100 spp change") +
  bi_theme()
```

```{r}
legend <- bi_legend(pal = "DkBlue2",
                    dim = 3,
                    xlab = "More new spp ",
                    ylab = "More loss spp",
                    size = 8)
```

```{r}
final_plot_85_2100 <- ggdraw() +
  draw_plot(biplot_85_2100, 0, 0, 1, 1) +
  draw_plot(legend, .8, .8, 0.2, 0.2)
```

```{r}
final_plot_85_2100
```

## Plots by taxonomic group

```{r}
iucn_lookup <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/_data_raw/iucn_redlist_2026/iucn_worms_aphia_id_complete.csv")
```

```{r}
iucn_cat <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/_data_raw/iucn_redlist_2026/iucn_species_assessments.csv")
```

```{r}
# Add aphiaid to iucn_cat
iucn_cat_aphiaid <- left_join(iucn_cat, (iucn_lookup %>% dplyr::select(iucn_assessment_id, aphia_id)),
 by = c("assessment_id" = "iucn_assessment_id")) %>% 
  rename(aphiaid = aphia_id) 
```

```{r}
# Merge in iucn_cat to get group for each species
spp_eez_pxl_count_group <- left_join(spp_eez_pxl_count, iucn_cat_aphiaid)
```

```{r}
# Get count of number of EEZ by group and count of IUCN cat by group (create figure similar to bar chart)
group_eez_count <- spp_eez_pxl_count_group %>% 
  # Filter for 
  filter(!MRGID %in% overlap_claim_list) %>%
  group_by(group, aphiaid, redlist_category) %>% 
  summarize(current_eez_count = sum(n_current >= 10, na.rm = TRUE),
            RCP45_2050_eez_count = sum(n_RCP45_2050 >= 10, na.rm = TRUE),
            RCP45_2100_eez_count = sum(n_RCP45_2100 >= 10, na.rm = TRUE),
            RCP85_2050_eez_count = sum(n_RCP85_2050 >= 10, na.rm = TRUE),
            RCP85_2100_eez_count = sum(n_RCP85_2100 >= 10, na.rm = TRUE)) %>% 
  ungroup()

group_iucncat_count <- group_eez_count %>% 
  group_by(group, redlist_category) %>% 
  summarize(redlist_count = n()) %>% 
  ungroup()
```


```{r}
ggplot() +
  geom_density(data = group_eez_count, aes(x = current_eez_count, fill = group), alpha = 0.5) +
  labs(title = "Distribution of current EEZ counts by taxonomic group",
       x = "current EEZ count") +
  facet_wrap(~group) +
  theme_bw()
```

```{r}
ggplot() +
  geom_density(data = group_eez_count, aes(x = RCP85_2100_eez_count, fill = group), alpha = 0.5) +
  labs(title = "Distribution of RCP85_2100 EEZ counts by taxonomic group",
       x = "RCP85_2100 EEZ count") +
  facet_wrap(~group) +
  theme_bw()
```


```{r}
# Assign EEZ count thresholds based on plotted density (drop not comprehensively assessed)
group_eez_count_threshold <- group_eez_count %>% 
  filter(group != "not comprehensively assessed") %>%
  mutate(current_eez_count_cat = case_when(current_eez_count == 1 ~ "1",
                                          current_eez_count == 2 ~ "2",
                                           current_eez_count > 2 & current_eez_count <= 5 ~ "3-5",
                                           current_eez_count > 5 & current_eez_count <= 15 ~ "6-15",
                                           current_eez_count > 15 & current_eez_count <= 25 ~ "16-25",
                                           current_eez_count > 25 & current_eez_count <= 50 ~ "26-50",
                                           current_eez_count > 50 & current_eez_count <= 100 ~ "51-100",
                                           current_eez_count > 100 ~ ">100",
                                           TRUE ~ "0"),
         RCP85_2050_eez_count_cat = case_when(RCP85_2050_eez_count == 1 ~ "1",
                                            RCP85_2050_eez_count == 2 ~ "2",
                                           RCP85_2050_eez_count > 2 & RCP85_2050_eez_count <= 5 ~ "3-5",
                                           RCP85_2050_eez_count > 5 & RCP85_2050_eez_count <= 15 ~ "6-15",
                                           RCP85_2050_eez_count > 15 & RCP85_2050_eez_count <= 25 ~ "16-25",
                                           RCP85_2050_eez_count > 25 & RCP85_2050_eez_count <= 50 ~ "26-50",
                                           RCP85_2050_eez_count > 50 & RCP85_2050_eez_count <= 100 ~ "51-100",
                                           RCP85_2050_eez_count > 100 ~ ">100",
                                           TRUE ~ "0")) %>%

  mutate(general_group = case_when(group %in% c("hagfishes", "sturgeons", "tunas_and_billfishes", "tarpons_and_ladyfishes")~ "pelagic_fishes",
                                   group %in% c("angelfishes", "blennies", "butterfly_fishes", "groupers", "pufferfishes", "seabreams_porgies_picarels", "surgeonfishes", "wrasses_and_parrotfishes") ~ "reef_fishes",
                                   group %in% c("marine_turtles", "seasnakes") ~ "marine_turtles_and_seasnakes",
                                   TRUE ~ group))
```

```{r}
cat_summary <- group_eez_count_threshold %>% 
  group_by(group, current_eez_count_cat) %>% 
  summarise(n_cat = n()) %>% 
  ungroup()

# Get total number in each group
group_count <- group_eez_count_threshold %>% 
  group_by(group) %>% 
  summarise(n_group = n()) %>% 
  ungroup()

# Merge in group count to get percentage
cat_summary_perc <- left_join(cat_summary, group_count) %>% 
  mutate(perc_cat = n_cat/n_group * 100)
```

```{r}
group_eez_count_threshold$current_eez_count_cat <- factor(group_eez_count_threshold$current_eez_count_cat, 
levels = c("0", "1", "2", "3-5", "6-15", "16-25", "26-50", "51-100", ">100"))

group_eez_count_threshold$RCP85_2050_eez_count_cat <- factor(group_eez_count_threshold$RCP85_2050_eez_count_cat, 
levels = c("0", "1", "2", "3-5", "6-15", "16-25", "26-50", "51-100", ">100"))
```

```{r}
library(forcats)
# Plot a bar graph by group with fill as the EEZ count category
ggplot() +
  geom_bar(data = group_eez_count_threshold, aes(y = fct_rev(general_group), fill = fct_rev(current_eez_count_cat)), position = "fill") +
  scale_fill_viridis_d() +
  labs(title = "Current EEZ count category by IUCN group",
       y = "",
       x = "proportion",
       fill = "EEZ count") +
  theme_bw()
```

```{r}
# Plot a bar graph by group with fill as the EEZ count category
ggplot() +
  geom_bar(data = group_eez_count_threshold, aes(y = fct_rev(general_group), fill = fct_rev(RCP85_2050_eez_count_cat)), position = "fill") +
  scale_fill_viridis_d() +
  labs(title = "RCP 8.5 2050 EEZ count category by IUCN group",
       y = "",
       x = "proportion",
       fill = "EEZ count") +
  theme_bw()
```

## Set up dataframe so that it's easier to plot current vs future side by side for each group 
```{r}
eez_count_group_simple <- group_eez_count_threshold %>% 
  dplyr::select(general_group, current_eez_count_cat, RCP85_2050_eez_count_cat) %>% 
  pivot_longer(cols = c(current_eez_count_cat, RCP85_2050_eez_count_cat), names_to = "scenario", values_to = "eez_count_cat") %>% 
  mutate(scenario = str_replace(scenario, "current_eez_count_cat", "current"),
         scenario = str_replace(scenario, "RCP85_2050_eez_count_cat", "RCP85_2050"))
```

```{r}
# Plot a bar graph by group with fill as the EEZ count category
taxon_bar_plot <- ggplot() +
  geom_bar(data = eez_count_group_simple, aes(y = fct_rev(general_group), fill = fct_rev(eez_count_cat)), position = "fill") +
  scale_fill_viridis_d(option = "turbo") +
  labs(title = "",
       y = "",
       x = "proportion",
       fill = "EEZ count") +
  theme_bw() + 
  theme(legend.position = "bottom") +
  facet_wrap(~scenario)
```

## Create jitter plot of the difference btwn RCP85_2050 and current EEZ count by group 

```{r}
group_eez_count_diff <- group_eez_count_threshold %>% 
  mutate(eez_count_diff = RCP85_2050_eez_count - current_eez_count) %>% 
  dplyr::select(group, general_group, aphiaid, current_eez_count,RCP85_2050_eez_count, eez_count_diff)
```

```{r}
taxon_jitter_plot_only <- ggplot() +
  geom_jitter(data = group_eez_count_diff, aes(x = eez_count_diff, y = fct_rev(general_group), alpha = 0.6),
   color = "gray10", position = position_jitterdodge(dodge.width = 0, jitter.width = .6),) +
  scale_y_discrete() + 
  labs(title = "EEZ subregion counts by IUCN comprehensive group",
       x = "EEZ subregion count change\n(RCP8.5 2050 - current)",
       y = "") +
  theme_bw() +
  theme(legend.position = "none",
        plot.margin = margin(0, 0, 0, 0))
```

```{r}
taxon_jitter_plot <- ggplot() +
  geom_jitter(data = group_eez_count_diff, aes(x = eez_count_diff, y = fct_rev(general_group), alpha = 0.6),
   color = "gray10", position = position_jitterdodge(dodge.width = 0, jitter.width = .75),) +
  scale_color_viridis_d(option = "viridis", direction = -1) +
  labs(title = "",
       x = "EEZ count change\n(RCP85 2050 - current)",
       y = "") +
  theme_bw() +
  theme(legend.position = "none",
        plot.margin = margin(0, 0, 0, 0),
        axis.text.y = element_blank())
```

```{r}
taxon_combined <- taxon_bar_plot + taxon_jitter_plot + 
  plot_annotation(title = "EEZ counts by IUCN comprehensive group") +
  plot_layout(widths = c(6, 3))  
```

```{r}
# Save only the jitter plot
ggsave(here("figures/iucn_threatened_subregion_taxon_jitterplot_070126.png"), taxon_jitter_plot_only, width = 7, height = 5, dpi = 300)
```

```{r}
# Save plots
ggsave(here("figures/iucn_threatened_taxon_plot_062326.png"), taxon_combined, width = 10, height = 5, dpi = 300)
```

## IUCN threats
For those species experiencing range shifts, what are the most common IUCN threats? Are there particular threats that are more common for species experiencing range loss vs gain?
https://www.iucnredlist.org/resources/threat-classification-scheme

According to the IUCN, threats marked with timing as “ongoing”, severity as “rapid” or “very rapid” declines and scopes of “majority” or “whole” are typically considered major threats to a species. IUCN stopped using scores in 2022 due to lack of sceintific method.

```{r}
library(iucnredlist) ### remotes::install_github("IUCN-UK/iucnredlist")
```

```{r}
api_key <- Sys.getenv("IUCN_API_KEY") ### usethis::edit_r_environ() to add key to R environment file
api <- init_api(api_key)
```

```{r}
iucn_cat_threat <- iucn_cat %>% 
  filter(redlist_category %in% c("CR", "EN", "VU"))
```

```{r}
# Get IUCN ID data 
iucn_id_data <- iucn_cat_threat$assessment_id
```

```{r}
# still need to run this through assesment_data_many
iucn_id_data <- assessment_data_many(api, iucn_id_data[1:5], wait_time = 0.5)
```

```{r}
# run element_names() to see what categories can be extracted
iucn_threats <- extract_element(iucn_id_data, "threats")
```

Threat information is pretty complicated, so maybe will focus on a subset of species for now. Perhaps exiting a certain EEZ. 

```{r}
# Singapore / Malaysia & Malaysia EEZs has the most threatened species exits under RCP8.5 2050 - get list of species exiting this EEZ
spp_exit_eez_top <- spp_eez_pxl_count_group %>% 
  filter(MRGID %in% c("64440", "8485")) %>%
  filter(is.na(n_RCP85_2050) & n_current > 0)
```

```{r}
# Get IUCN IDs for these species
iucn_ids_exit_eez_top <- unique(spp_exit_eez_top$assessment_id)
```

```{r}
# still need to run this through assesment_data_many
iucn_id_data_eez_top <- assessment_data_many(api, iucn_ids_exit_eez_top, wait_time = 0.5)
```

```{r}
# run element_names() to see what categories can be extracted
iucn_threats_eez_top <- extract_element(iucn_id_data_eez_top, "threats") %>% 
  mutate(assessment_id = as.numeric(assessment_id))
```

```{r}
# Merge in the threat data to the spp_exit_eez_top dataframe to see if there are particular threats that are more common for species exiting this EEZ
spp_exit_eez_top_threats <- inner_join(spp_exit_eez_top, iucn_threats_eez_top, by = c("assessment_id" = "assessment_id"))
```


```{r}
spp_exit_eez_top_threats_summary <- spp_exit_eez_top_threats %>% 
  group_by(MRGID, GEONAME, code, description) %>% 
  summarize(n_spp = n()) %>% 
  ungroup() %>% 
  arrange(desc(n_spp))

```

```{r}
# Try filtering for only severe threats (scope of majority or whole, severity of rapid or very rapid, timing of ongoing)
# 7 of the 75 species exiting the Singapore/Malaysia EEZ under RCP8.5 2050 are experiencing severe threats according to IUCN criteria
# Threats are all connected to fishing and harvest, along with shipping lanes
spp_exit_eez_top_severe_threats_summary <- spp_exit_eez_top_threats %>% 
  filter(scope.y %in% c("Majority (50-90%)", "Whole (>90%)") &
         severity %in% c("Rapid Declines", "Very Rapid Declines") &
         timing == "Ongoing") %>%
  group_by(MRGID, GEONAME, code, description) %>% 
  summarize(n_spp = n()) %>% 
  ungroup() %>% 
  arrange(desc(n_spp))

```


```{r}
# What about all the threatened species leaving EEZs under RCP8.5 2050
spp_exit_eez <- spp_eez_pxl_count_group %>% 
  filter(is.na(n_RCP85_2050) & n_current > 0)
```

```{r}
# Get IUCN IDs for these species
iucn_ids_exit_eez <- unique(spp_exit_eez$assessment_id)
```

```{r}
# still need to run this through assesment_data_many
iucn_id_data_eez <- assessment_data_many(api, iucn_ids_exit_eez, wait_time = 0.5)
```

```{r}
# run element_names() to see what categories can be extracted
iucn_threats_eez <- extract_element(iucn_id_data_eez, "threats") %>% 
  mutate(assessment_id = as.numeric(assessment_id))
```

```{r}
# Merge in the threat data to the spp_exit_eez_top dataframe to see if there are particular threats that are more common for species exiting this EEZ
spp_exit_eez_threats <- inner_join(spp_exit_eez, iucn_threats_eez, by = c("assessment_id" = "assessment_id"))
```

```{r}
# Filter for the severe threats (33 threatened species experiencing severe threats)
spp_exit_eez_severe_threats <- spp_exit_eez_threats %>% 
  filter(scope.y %in% c("Majority (50-90%)", "Whole (>90%)") &
         severity %in% c("Rapid Declines", "Very Rapid Declines") &
         timing == "Ongoing")  

spp_exit_eez_severe_threats_summary <- spp_exit_eez_severe_threats %>% 
  group_by(MRGID, GEONAME, code, description) %>% 
  summarize(n_spp = n()) %>% 
  ungroup() %>% 
  arrange(desc(n_spp))
```


```{r}
# IDs for severe threat species that are leaving EEZS under RCP8.5 2050
iucn_ids_exit_eez_severe_threats <- unique(spp_exit_eez_severe_threats$aphiaid)
```

## Plots with combined Sovereign EEZs

```{r}
# Organize species EEZ count data so that EEZs sections are combined
# This will count up the species that are entering and exiting sovereign EEZs - alternatively we could still count species entries into subparts of EEZs
spp_sov_eez_pxl_count <- left_join(spp_eez_pxl_count, eez_no_overlap_df) %>% 
  # Filter for only EEZs (no joint regime)
  filter(POL_TYPE == "200NM") %>%
  # Group by sovereign EEZ and aphiaid
  group_by(SOVEREIGN1, MRGID_SOV1, aphiaid) %>% 
  summarize(n_current = sum(n_current, na.rm = TRUE),
            n_RCP45_2050 = sum(n_RCP45_2050, na.rm = TRUE),
            n_RCP45_2100 = sum(n_RCP45_2100, na.rm = TRUE),
            n_RCP85_2050 = sum(n_RCP85_2050, na.rm = TRUE),
            n_RCP85_2100 = sum(n_RCP85_2100, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
spp_sov_eez_spp_count <- spp_sov_eez_pxl_count %>% 
  group_by(aphiaid) %>% 
  summarize(n_current = sum(n_current > 0, na.rm = TRUE),
            n_RCP45_2050 = sum(n_RCP45_2050 > 0, na.rm = TRUE),
            n_RCP45_2100 = sum(n_RCP45_2100 > 0, na.rm = TRUE),
            n_RCP85_2050 = sum(n_RCP85_2050 > 0, na.rm = TRUE),
            n_RCP85_2100 = sum(n_RCP85_2100 > 0, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
ggplot() +
  geom_point(data = spp_sov_eez_spp_count, aes(x = n_current, y = n_RCP85_2050),
             alpha = 0.5, color = "maroon") +
  #geom_point(data = eez_spp_counts, aes(x = current_eez_count, y = RCP85_2100_eez_count),
             #alpha = 0.5, color = "forestgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  labs(title = "IUCN Threatened Species Sovereign EEZ count (RCP 8.5 2050)",
       y = "RCP 85 future EEZ count",
       x = "Current EEZ count") 
```


```{r}
# Organize dataframe so that's it's easier to select a scenario and compare to current
sov_eez_spp_pxls_long <- spp_sov_eez_pxl_count %>% 
  pivot_longer(cols = starts_with("n_RCP"),
               names_to = "scenario",
               values_to = "pxl_count") %>% 
  mutate(scenario = str_remove(scenario, "n_")) %>% 
  # Replace NA with 0, otherwise will be assigned to "continued presence"
  replace_na(list(pxl_count = 0, n_current = 0)) %>% 
  mutate(spp_change = case_when(n_current < 10 &  pxl_count >= 10 ~ 'entry into new',
                            n_current >= 10  & pxl_count < 10 ~ 'exit from historical',
                            n_current < 10  & pxl_count < 10 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

```{r}
# Get counts of entry into new, exit from historical etc. by EEZ
sov_eez_spp_change <- sov_eez_spp_pxls_long %>% 
  group_by(SOVEREIGN1, MRGID_SOV1, scenario, spp_change) %>% 
  summarize(spp_change_count = n()) %>% 
  ungroup()
```

Setting up the counts of species entries and exits by sovereign EEZ loses the nuance of species entering/exiting sub-parts of sovereign EEZs. For example, the US only has 10 species exits when you account for all components of the EEZ. However, when you treat the components of the US EEZ, separately, there are 46 species exits for the Howard & Baker Islands subpart. Interestingly, when the data is done this way, Singapore is in the top three for species exits and entries.

Could also create the taxonomic plot by the combined sovereign EEZs.

```{r}
# Merge in iucn_cat to get group for each species
spp_sov_eez_pxl_count_group <- left_join(spp_sov_eez_pxl_count, iucn_cat_aphiaid)

group_sov_eez_count <- spp_sov_eez_pxl_count_group %>% 
  group_by(group, aphiaid, redlist_category) %>% 
  summarize(current_eez_count = sum(n_current >= 10, na.rm = TRUE),
            RCP45_2050_eez_count = sum(n_RCP45_2050 >= 10, na.rm = TRUE),
            RCP45_2100_eez_count = sum(n_RCP45_2100 >= 10, na.rm = TRUE),
            RCP85_2050_eez_count = sum(n_RCP85_2050 >= 10, na.rm = TRUE),
            RCP85_2100_eez_count = sum(n_RCP85_2100 >= 10, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
# Assign EEZ count thresholds based on plotted density (drop not comprehensively assessed)
group_sov_eez_count_threshold <- group_sov_eez_count %>% 
  filter(group != "not comprehensively assessed") %>%
  mutate(current_eez_count_cat = case_when(current_eez_count == 1 ~ "1",
                                          current_eez_count == 2 ~ "2",
                                           current_eez_count > 2 & current_eez_count <= 5 ~ "3-5",
                                           current_eez_count > 5 & current_eez_count <= 15 ~ "6-15",
                                           current_eez_count > 15 & current_eez_count <= 25 ~ "16-25",
                                           current_eez_count > 25 & current_eez_count <= 50 ~ "26-50",
                                           current_eez_count > 50 & current_eez_count <= 100 ~ "51-100",
                                           current_eez_count > 100 ~ ">100",
                                           TRUE ~ "0"),
         RCP85_2050_eez_count_cat = case_when(RCP85_2050_eez_count == 1 ~ "1",
                                            RCP85_2050_eez_count == 2 ~ "2",
                                           RCP85_2050_eez_count > 2 & RCP85_2050_eez_count <= 5 ~ "3-5",
                                           RCP85_2050_eez_count > 5 & RCP85_2050_eez_count <= 15 ~ "6-15",
                                           RCP85_2050_eez_count > 15 & RCP85_2050_eez_count <= 25 ~ "16-25",
                                           RCP85_2050_eez_count > 25 & RCP85_2050_eez_count <= 50 ~ "26-50",
                                           RCP85_2050_eez_count > 50 & RCP85_2050_eez_count <= 100 ~ "51-100",
                                           RCP85_2050_eez_count > 100 ~ ">100",
                                           TRUE ~ "0")) %>%
  mutate(general_group = case_when(group %in% c("hagfishes", "sturgeons", "tunas_and_billfishes", "tarpons_and_ladyfishes")~ "pelagic_fishes",
                                   group %in% c("angelfishes", "blennies", "butterfly_fishes", "groupers", "pufferfishes", "seabreams_porgies_picarels", "surgeonfishes", "wrasses_and_parrotfishes") ~ "reef_fishes",
                                   group %in% c("marine_turtles", "seasnakes") ~ "marine_turtles_and_seasnakes",
                                   TRUE ~ group))
```

```{r}
group_sov_eez_count_threshold$current_eez_count_cat <- factor(group_sov_eez_count_threshold$current_eez_count_cat, 
levels = c("0", "1", "2", "3-5", "6-15", "16-25", "26-50", "51-100", ">100"))

group_sov_eez_count_threshold$RCP85_2050_eez_count_cat <- factor(group_sov_eez_count_threshold$RCP85_2050_eez_count_cat, 
levels = c("0", "1", "2", "3-5", "6-15", "16-25", "26-50", "51-100", ">100"))
```

```{r}
sov_eez_count_group_simple <- group_sov_eez_count_threshold %>% 
  dplyr::select(general_group, current_eez_count_cat, RCP85_2050_eez_count_cat) %>% 
  pivot_longer(cols = c(current_eez_count_cat, RCP85_2050_eez_count_cat), names_to = "scenario", values_to = "eez_count_cat") %>% 
  mutate(scenario = str_replace(scenario, "current_eez_count_cat", "current"),
         scenario = str_replace(scenario, "RCP85_2050_eez_count_cat", "RCP85_2050"))
```

```{r}
# Plot a bar graph by group with fill as the EEZ count category
sov_taxon_bar_plot <- ggplot() +
  geom_bar(data = sov_eez_count_group_simple, aes(y = fct_rev(general_group), fill = fct_rev(eez_count_cat)), position = "fill") +
  scale_fill_viridis_d(option = "turbo") +
  labs(title = "",
       y = "",
       x = "proportion",
       fill = "Sovereign count") +
  theme_bw() + 
  theme(legend.position = "bottom") +
  facet_wrap(~scenario)
```
```{r}
group_sov_eez_count_diff <- group_sov_eez_count_threshold %>% 
  mutate(eez_count_diff = RCP85_2050_eez_count - current_eez_count) %>% 
  dplyr::select(group, general_group, aphiaid, current_eez_count,RCP85_2050_eez_count, eez_count_diff)
```

```{r}
taxon_sov_jitter_plot_only <- ggplot() +
  geom_jitter(data = group_sov_eez_count_diff, aes(x = eez_count_diff, y = fct_rev(general_group), alpha = 0.6),
   color = "gray10", position = position_jitterdodge(dodge.width = 0, jitter.width = .6),) +
  scale_y_discrete() + 
  labs(title = "Sovereign counts by IUCN comprehensive group",
       x = "Sovereign count change\n(RCP8.5 2050 - current)",
       y = "") +
  theme_bw() +
  theme(legend.position = "none",
        plot.margin = margin(0, 0, 0, 0))
```

```{r}
# Save only the jitter plot
ggsave(here("figures/iucn_threatened_taxon_sovereign_jitterplot_070126.png"), taxon_sov_jitter_plot_only, width = 7, height = 5, dpi = 300)
```


Instead, use the original species entry/exit data but then get species entries and exits by sovereign EEZ. This will count species that are entering/exiting sub-parts of sovereign EEZs as entries/exits for the sovereign EEZ.

```{r}
# Get counts of entry into new, exit from historical etc. by EEZ
eez_spp_change_sov_counts <- left_join(eez_spp_pxls_long, eez_no_overlap_df) %>% 
  filter(POL_TYPE == "200NM") %>%
  group_by(SOVEREIGN1, MRGID_SOV1, scenario, spp_change) %>% 
  summarize(spp_change_count = n()) %>% 
  ungroup()
```

Now, France, North Korea, United Kingdom, Russia, and United Kingdom are the top five for species entry. France is also at the top of the list for species exits with 227 species exits (for France, across their EEZ subparts there are 227 instances of species entirely leaving EEZ regions - pretty interesting). In comparision, there are 11 species exits for France if you dissolve the subparts of their EEZ. 