---
title: "Plots of Threatened Species Ranges & EEZs"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
library(arrow)
library(patchwork)
library(cowplot)
```

This document creates updated plots useing the updated entry and exit data from 01_iucn_aquax_eez_overlap_db.qmd

## Read in at-risk species data

```{r}
# Spp in EEZs
spp_eez_pxl_count <- read_csv(here("data/iucn_threatened_spp_eez_pxls_061626.csv"))
```

```{r}
# Spp EEZ entry/exit pixels
spp_entry_pxls <- read_rds(here("data/iucn_atrisk_10pxlthres_entry_pxls_072826.rds"))
spp_exit_pxls <- read_rds(here("data/iucn_atrisk_10pxlthres_exit_pxls_072926.rds"))
```

```{r}
# Species range area
area_df <- read_csv(here("data/iucn_threatened_spp_range_area_061826.csv"))
```

## Read in EEZs

```{r}
eez <- st_read('/Users/abowd/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
# Get list of EEZ IDs that are not overlapping claims (i.e., POL_TYPE = 200NM)
eez_200nm_list <- eez %>% 
  filter(POL_TYPE == "200NM") %>% 
  pull(MRGID)
```

```{r}
eez_simple <- eez %>% 
  dplyr::select(MRGID, GEONAME) %>% 
  st_drop_geometry() %>% 
  filter(MRGID %in% eez_200nm_list)

eez_simple_sf <- eez %>% 
  dplyr::select(MRGID, GEONAME) %>% 
  filter(MRGID %in% eez_200nm_list)
```

## Plot species pixels
Only include 200 nm EEZs (no overlapping claims or joint regimes)

```{r}
spp_entry_pxls_85_2050 <- spp_entry_pxls %>% 
  filter(scenario == "RCP85_2050") %>% 
  filter(MRGID %in% eez_200nm_list)

spp_exit_pxls_85_2050 <- spp_exit_pxls %>% 
  filter(scenario == "RCP85_2050") %>% 
  filter(MRGID %in% eez_200nm_list)
```

```{r}
# Plot EEZ entry pixels
plot_entry_10_map <- ggplot() +
  geom_sf(data = eez_simple_sf) +
  coord_sf(expand = FALSE) +
  geom_tile(data = spp_entry_pxls_85_2050, aes(x = x, y = y,
                                               fill = n_entry_spp,
                                               color = n_entry_spp)) +
  scale_fill_viridis_c(limits = c(1, 40), 
                      breaks = c(1, 2, 4, 8, 16, 28, 40),
                      direction = -1,
                      trans = "log", # visually stretches the lower value data
                      guide = guide_colorbar(
                      barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                      barheight = unit(0.4, "cm"),
                      ticks = TRUE,
                      ticks.colour = "white",
                      ticks.linewidth = 0.8
  )) +
  scale_color_viridis_c(limits = c(1,40), 
                        breaks = c(1, 2, 4, 8, 16, 28, 40),
                        direction = -1,
                        trans = "log", # visually stretches the lower value data
                        guide = "none")  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "species entry count",
       color = "species entry count",
       x = "",
       y = "") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt")) 
```

```{r}
plot_entry_10_density <- ggplot(spp_entry_pxls_85_2050, aes(x = y)) +
  geom_density(fill = "forestgreen", alpha = 0.5) +
  coord_flip() +
  scale_x_continuous(limits = c(-90, 90)) +
  theme_minimal() +
  labs(x = "",
       y = "") +
  theme(plot.margin = margin(0, 0, 0, 0))
```

```{r}
entry_combined_plot <- plot_entry_10_map + plot_entry_10_density + 
  plot_annotation(title = "Species entry, RCP 8.5 2050") +
  plot_layout(widths = c(4, 1),
              axis_titles = "collect",
              guides = "collect") &
  theme(legend.position = "bottom") 
```

```{r}
# Plot EEZ exit pixels
plot_exit_10_map <- ggplot() +
  geom_sf(data = eez_simple_sf) +
  coord_sf(expand = FALSE) +
  geom_tile(data = spp_exit_pxls_85_2050, aes(x = x, y = y,
                                              fill = n_exit_spp,
                                              color = n_exit_spp)) +
  scale_fill_viridis_c(limits = c(1,40), 
                       breaks = c(1, 2, 4, 8, 16, 28, 40),
                       direction = -1, 
                       trans = "log", # visually stretches the lower value data
                       guide = guide_colorbar(
                       barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                       barheight = unit(0.4, "cm"),
                       ticks = TRUE,
                       ticks.colour = "white",
                       ticks.linewidth = 0.8)) +
  scale_color_viridis_c(limits = c(1,40), 
                        breaks = c(1, 2, 4, 8, 16, 32, 40), 
                        direction = -1,
                        trans = "log", # visually stretches the lower value data
                        guide = "none")  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "species exit count",
       color = "species exit count",
       x = "",
       y = "") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt"))
```

```{r}
plot_exit_10_density <- ggplot(spp_exit_pxls_85_2050, aes(x = y)) +
  geom_density(fill = "darkmagenta", alpha = 0.5) +
  coord_flip() +
  scale_x_continuous(limits = c(-90, 90)) +
  theme_minimal() +
  labs(x = "",
       y = "") +
  theme(plot.margin = margin(0, 0, 0, 0))
```

```{r}
exit_combined_plot <- plot_exit_10_map + plot_exit_10_density + 
  plot_annotation(title = "Species exit, RCP 8.5 2050") +
  plot_layout(widths = c(4, 1),
              axis_titles = "collect",
              guides = "collect") &
  theme(legend.position = "bottom") 
```

```{r}
# Save entry and exit maps
ggsave(here("figures/iucn_atrisk_spp_entry_10pxl_map_072926.png"), entry_combined_plot, width = 10, height = 5, dpi = 300)
ggsave(here("figures/iucn_atrisk_spp_exit_10pxl_map_072926.png"), exit_combined_plot, width = 10, height = 5, dpi = 300)
```

## Bivariate plot

Code created with assistance from Claude to categorize the raster entry and exit pixels into a bivariate color scheme using the following categories to create a 3x3 grid:
* 0 species entry and 0 species exit
* 0 species entry and 1 species exit
* 1 species entry and 0 species exit
* 1 species entry and 1 species exit
* 2+ species entry and 2+ species exit

```{r}
# Merge the entry and exit dataframes so that there are the same number of pixels in each raster
spp_entry_exit_85_2050 <- full_join(spp_entry_pxls_85_2050, spp_exit_pxls_85_2050) %>% 
  # Replace NA with 0
  replace_na(list(n_entry_spp = 0, n_exit_spp = 0))
```

```{r}
# Create a new column that categorizes the number of species entry and exit into 3 categories: 0, 1, and 2+
spp_entry_exit_cat_85_2050 <- spp_entry_exit_85_2050 %>% 
  # Assign numeric_category
  mutate(entry_cat = case_when(n_entry_spp == 0 ~ 1,
                               n_entry_spp == 1 ~ 2,
                               n_entry_spp >= 2 ~ 3),
         exit_cat = case_when(n_exit_spp == 0 ~ 1,
                              n_exit_spp == 1 ~ 2,
                              n_exit_spp >= 2 ~ 3)) %>% 
  # Create a new column that combines the entry and exit categories into a single category
  mutate(entry_exit_cat = (exit_cat - 1) * 3 + entry_cat)
```

```{r}
# Define numeric categories for the bivariate color scheme
bivar_levels <- data.frame(
  entry_exit_cat    = 1:9,
  label = c(
    "En0_Ex0", "En1_Ex0", "En2_Ex0",   # exits = 0
    "En0_Ex1", "En1_Ex1", "En2_Ex1",   # exits = 1
    "En0_Ex2", "En1_Ex2", "En2_Ex2"    # exits = 2+
  )
)
spp_entry_exit_cat_85_2050 <- left_join(spp_entry_exit_cat_85_2050, bivar_levels)
```

```{r}
# Color palette
# Entries increase left → right (blue axis)
# Exits increase bottom → top (pink/purple axis)
bivar_colors <- c(
  "En0_Ex0" = "#e8e8e8",  # entries=0,   exits=0
  "En1_Ex0" = "#ace4e4",  # entries=1, exits=0
  "En2_Ex0" = "#5ac8c8",  # entries=2+,  exits=0
  "En0_Ex1" = "#dfb0d6",  # entries=0,   exits=1
  "En1_Ex1" = "#a5add3",  # entries=1, exits=1
  "En2_Ex1" = "#5698b9",  # entries=2+,  exits=1
  "En0_Ex2" = "#be64ac",  # entries=0,   exits=2+
  "En1_Ex2" = "#8c62aa",  # entries=1, exits=2+
  "En2_Ex2" = "#3b4994"   # entries=2+,  exits=2+
)
```

```{r}
biplot <- ggplot() +
  geom_sf(data = eez_simple_sf, fill = "#e8e8e8") +
  coord_sf(expand = FALSE) +
  geom_tile(data = spp_entry_exit_cat_85_2050, aes(x = x, y = y,
                                              fill = label,
                                              color = label)) +
  scale_fill_manual(values = bivar_colors, na.value = "white") +
  scale_color_manual(values = bivar_colors, na.value = "white") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Species exits vs entries, RCP 8.5 2050",
       x = "",
       y = "")
```

```{r}
# Create legend for bivariate color scheme
legend_df <- expand.grid(entries = 1:3, exits = 1:3)
legend_df$label <- paste0("En", legend_df$entries - 1, "_Ex", legend_df$exits - 1)

legend_biplot <- ggplot(legend_df, aes(x = entries, y = exits, fill = label)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_manual(values = bivar_colors) +
  scale_x_continuous(breaks = 1:3, labels = c("0", "1", "2+")) +
  scale_y_continuous(breaks = 1:3, labels = c("0", "1", "2+")) +
  labs(x = "Entries →", y = "Exits →") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text       = element_text(size = 8),
    axis.title      = element_text(size = 9)
  )
```

```{r}
# Combine plot
bivariate_map <- ggdraw() +
  draw_plot(biplot,    0,    0,    1,    1) +
  draw_plot(legend_biplot, 0.75, 0.75, 0.2, 0.25)
```

# Add density plot to bivariate map

```{r}
biplot_density <-ggplot() +
  geom_density(data = spp_exit_pxls_85_2050, aes(x = y),fill = "#be64ac", color = "#be64ac", alpha = 0.5, size = 1) +
  geom_density(data = spp_entry_pxls_85_2050, aes(x = y),fill = "#5698b9", color = "#5698b9", alpha = 0.5, size = 1) +
  coord_flip() +
  scale_x_continuous(limits = c(-90, 90)) +
  theme_minimal() +
  labs(x = "",
       y = "") +
  theme(plot.margin = margin(0, 0, 0, 0))
```

```{r}
biplot_2 <- ggplot() +
  geom_sf(data = eez_simple_sf, fill = "#e8e8e8") +
  coord_sf(expand = FALSE) +
  geom_tile(data = spp_entry_exit_cat_85_2050, aes(x = x, y = y,
                                              fill = label,
                                              color = label)) +
  scale_fill_manual(values = bivar_colors, na.value = "white") +
  scale_color_manual(values = bivar_colors, na.value = "white") +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt")) +
  labs(title = "",
       x = "",
       y = "") 
```

```{r}
biplot_and_density <- biplot_2 + biplot_density + 
  plot_annotation(title = "Species exit vs entry, RCP 8.5 2050") +
  plot_layout(widths = c(4, 1),
              axis_titles = "collect",
              guides = "collect") &
  theme(legend.position = "none") 
```

```{r}
biplot_all <- ggdraw() +
  draw_plot(biplot_and_density,    0,    0,    1,    1) +
  draw_plot(legend_biplot, 0.6, 0.8, 0.2, 0.2)
```

```{r}
# Save entry and exit maps
ggsave(here("figures/iucn_atrisk_spp_entry_exit_density_10pxl_map_073026.png"), biplot_all, width = 10, height = 5, dpi = 300)
```

# Top species entry / exit by EEZ

```{r}
# Organize dataframe so that's it's easier to select a scenario and compare to current
eez_spp_pxls_long <- spp_eez_pxl_count %>% 
  # Drop overlapping claim EEZs
  filter(MRGID %in% eez_200nm_list) %>%
  pivot_longer(cols = starts_with("n_RCP"),
               names_to = "scenario",
               values_to = "pxl_count") %>% 
  mutate(scenario = str_remove(scenario, "n_")) %>% 
  # Replace NA with 0, otherwise will be assigned to "continued presence"
  replace_na(list(pxl_count = 0, n_current = 0)) %>% 
  mutate(spp_change = case_when(n_current < 10 &  pxl_count >= 10 ~ 'entry into new',
                            n_current >= 10  & pxl_count < 10 ~ 'exit from historical',
                            n_current < 10  & pxl_count < 10 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

```{r}
# Get counts of entry into new, exit from historical etc. by sovereign EEZ
eez_spp_change_sov_counts <- left_join(eez_spp_pxls_long, eez %>% filter(MRGID %in% eez_200nm_list)) %>% 
  group_by(SOVEREIGN1, MRGID_SOV1, scenario, spp_change) %>% 
  summarize(spp_change_count = n()) %>% 
  ungroup()
```

Now, France, North Korea, United Kingdom, Russia, and United Kingdom are the top five for species entry. France is also at the top of the list for species exits with 227 species exits (for France, across their EEZ subparts there are 227 instances of species entirely leaving EEZ regions - pretty interesting). In comparision, there are 11 species exits for France if you dissolve the subparts of their EEZ. 

```{r}
# Get counts of entry into new, exit from historical etc. by  EEZ region
eez_spp_change_counts <- left_join(eez_spp_pxls_long, eez %>% filter(MRGID %in% eez_200nm_list)) %>% 
  group_by(MRGID, GEONAME, scenario, spp_change) %>% 
  summarize(spp_change_count = n()) %>% 
  ungroup()
```

```{r}
eez_c_f <- eez_spp_change_counts  %>% 
  mutate(spp_change = str_replace_all(spp_change, " ", "_")) %>% 
  pivot_wider(names_from = spp_change, values_from = spp_change_count) %>% 
  replace_na(list(continued_presence = 0, exit_from_historical = 0,
                  entry_into_new = 0, no_presence = 0)) %>% 
  mutate(current_spp = continued_presence + exit_from_historical,
         future_spp = continued_presence + entry_into_new,
         perc_exit = 100 * exit_from_historical/current_spp,
         perc_enter = 100 * entry_into_new/current_spp,
         spp_change = future_spp - current_spp,
         ssi = (2 * continued_presence)/(current_spp + future_spp)) 
```

```{r}
# Save spsecies change data by EEZ
write_csv(eez_c_f, here("data/iucn_atrisk_eez_spp_change_all_scenarios.csv"))
```

```{r}
eez_sov_c_f <- eez_spp_change_sov_counts  %>% 
  mutate(spp_change = str_replace_all(spp_change, " ", "_")) %>% 
  pivot_wider(names_from = spp_change, values_from = spp_change_count) %>% 
  replace_na(list(continued_presence = 0, exit_from_historical = 0,
                  entry_into_new = 0, no_presence = 0)) %>% 
  mutate(current_spp = continued_presence + exit_from_historical,
         future_spp = continued_presence + entry_into_new,
         perc_exit = 100 * exit_from_historical/current_spp,
         perc_enter = 100 * entry_into_new/current_spp,
         spp_change = future_spp - current_spp,
         ssi = (2 * continued_presence)/(current_spp + future_spp)) 
```

# Entry and exit species percent by current IUCN listed species

```{r}
# Read in current global least concern species 
lc_sum_current_neg90_neg180 <- readRDS(here("data/iucn_least_concern/iucn_lc_current_neg180_to_neg90_080526.rds"))

lc_sum_current_neg90_zero <- readRDS(here("data/iucn_least_concern/iucn_lc_current_neg90_to_zero_080526.rds"))

lc_sum_current_zero_90 <- readRDS(here("data/iucn_least_concern/iucn_lc_current_zero_to_90_080526.rds"))

lc_sum_current_90_180 <- readRDS(here("data/iucn_least_concern/iucn_lc_current_90_to_180_080526.rds"))
```

```{r}
# Read in current global threatened species 
threat_sum_current_neg90_neg180 <- readRDS(here("data/iucn_atrisk_count_current_neg90_neg180_080526.rds"))

threat_sum_current_neg90_zero <- readRDS(here("data/iucn_atrisk_count_current_neg90_zero_080526.rds"))

threat_sum_current_zero_90 <- readRDS(here("data/iucn_atrisk_count_current_zero_90_080526.rds"))

threat_sum_current_90_180 <- readRDS(here("data/iucn_atrisk_count_current_90_180_080526.rds"))
```

```{r}
current_lc_threat_sum_df <- full_join(bind_rows(lc_sum_current_neg90_neg180,
                                           lc_sum_current_neg90_zero,
                                           lc_sum_current_zero_90,
                                           lc_sum_current_90_180),
                              bind_rows(threat_sum_current_neg90_neg180,
                                    threat_sum_current_neg90_zero,
                                    threat_sum_current_zero_90,
                                    threat_sum_current_90_180))
```

```{r}
# Join with species entry for RCP 8.5 2050 with current least concern and at-risk species
# Maintaining a consistent denominator to calculate the percent of species entering or exiting EEZs relative to the current number of species in each EEZ
current_lc_threat_count_entry <- left_join(spp_entry_pxls_85_2050,
                                           current_lc_threat_sum_df)
```

```{r}
n_entry_perc_allspp <- current_lc_threat_count_entry %>% 
  # replace na with 0 
  replace_na(0) %>% 
  mutate(n_entry_perc = round(100 * n_entry_spp/(N + N_atrisk), 2))
```

```{r}
# Plot percent entry pixels as percentage of current species present
plot_perc_entry_current <- ggplot() +
  geom_sf(data = eez_simple_sf) +
  coord_sf(expand = FALSE) +
  geom_tile(data = n_entry_perc_allspp, aes(x = x, y = y,
                                               fill = n_entry_perc,
                                               color = n_entry_perc)) +
  scale_fill_viridis_c(limits = c(0.03,100),
                      trans = "log",
                      option = "plasma",
                      breaks = c(0.03, 1, 5, 10, 25, 50, 100),
                      labels = c(0.03, 1, 5, 10, 25, 50, 100),
                      direction = -1,
                      guide = guide_colorbar(
                      barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                      barheight = unit(0.4, "cm"),
                      ticks = TRUE,
                      ticks.colour = "white",
                      ticks.linewidth = 0.8
  )) +
  scale_color_viridis_c(limits = c(0.03,100), 
                        trans = "log",
                        option = "plasma",
                        breaks = c(0.03, 1, 5, 10, 25, 50, 100),
                        labels = c(0.03, 1, 5, 10, 25, 50, 100),
                        direction = -1,
                        guide = "none")  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "species entry perc",
       color = "species entry perc",
       title = "Percent of current IUCN species that are new at-risk species by EEZ region",
       subtitle = "RCP 8.5 2050",
       x = "",
       y = "") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt")) 
```

```{r}
# Join with species exit for RCP 8.5 2050 with current least concern and at-risk species
# Maintaining a consistent denominator to calculate the percent of species entering or exiting EEZs relative to the current number of species in each EEZ
current_lc_threat_count_exit <- left_join(spp_exit_pxls_85_2050,
                                           current_lc_threat_sum_df)
```

```{r}
n_exit_perc_allspp <- current_lc_threat_count_exit %>% 
  # replace na with 0 
  replace_na(0) %>% 
  mutate(n_exit_perc = round(100 * n_exit_spp/(N + N_atrisk), 2))
```

```{r}
# Plot percent exit pixels as percentage of current species present
plot_perc_exit_current <- ggplot() +
  geom_sf(data = eez_simple_sf) +
  coord_sf(expand = FALSE) +
  geom_tile(data = n_exit_perc_allspp, aes(x = x, y = y,
                                               fill = n_exit_perc,
                                               color = n_exit_perc)) +
  scale_fill_viridis_c(limits = c(0.03,100),
                      trans = "log",
                      option = "plasma",
                      breaks = c(0.03, 1, 5, 10, 25, 50, 100),
                      labels = c(0.03, 1, 5, 10, 25, 50, 100),
                      direction = -1,
                      guide = guide_colorbar(
                      barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                      barheight = unit(0.4, "cm"),
                      ticks = TRUE,
                      ticks.colour = "white",
                      ticks.linewidth = 0.8
  )) +
  scale_color_viridis_c(limits = c(0.03,100), 
                        trans = "log",
                        option = "plasma",
                        breaks = c(0.03, 1, 5, 10, 25, 50,100),
                        labels = c(0.03, 1, 5, 10, 25, 50, 100),
                        direction = -1,
                        guide = "none")  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "species exit perc",
       color = "species exit perc",
       title = "Percent of current IUCN species that are at-risk species loss by EEZ region",
       subtitle = "RCP 8.5 2050",
       x = "",
       y = "") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt")) 
```

```{r}
# Save plot of percent entry and percent exits of current IUCN listed species
ggsave(here("figures/iucn_atrisk_perc_exit_spp_rcp85_2050_080626.png"), plot_perc_exit_current, width = 10, height = 5, dpi = 300)

ggsave(here("figures/iucn_atrisk_perc_entry_spp_rcp85_2050_080626.png"), plot_perc_entry_current, width = 10, height = 5, dpi = 300)
```