---
title: "Plots of Threatened Species Ranges & EEZs"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
library(arrow)
library(patchwork)
library(cowplot)
```

This document creates updated plots useing the updated entry and exit data from 01_iucn_aquax_eez_overlap_db.qmd

## Read in at-risk species data

```{r}
# Spp in EEZs
spp_eez_pxl_count <- read_csv(here("data/iucn_threatened_spp_eez_pxls_061626.csv"))
```

```{r}
# Spp EEZ entry/exit pixels
spp_entry_pxls <- read_rds(here("data/iucn_atrisk_10pxlthres_entry_pxls_072826.rds"))
spp_exit_pxls <- read_rds(here("data/iucn_atrisk_10pxlthres_exit_pxls_072926.rds"))
```

```{r}
# Species range area
area_df <- read_csv(here("data/iucn_threatened_spp_range_area_061826.csv"))
```

## Read in EEZs

```{r}
eez <- st_read('/Users/abowd/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
# Get list of EEZ IDs that are not overlapping claims (i.e., POL_TYPE = 200NM)
eez_200nm_list <- eez %>% 
  filter(POL_TYPE == "200NM") %>% 
  pull(MRGID)
```

```{r}
eez_simple <- eez %>% 
  dplyr::select(MRGID, GEONAME) %>% 
  st_drop_geometry() %>% 
  filter(MRGID %in% eez_200nm_list)

eez_simple_sf <- eez %>% 
  dplyr::select(MRGID, GEONAME) %>% 
  filter(MRGID %in% eez_200nm_list)
```

## Plot species pixels
Only include 200 nm EEZs (no overlapping claims or joint regimes)

```{r}
spp_entry_pxls_85_2050 <- spp_entry_pxls %>% 
  filter(scenario == "RCP85_2050") %>% 
  filter(MRGID %in% eez_200nm_list)

spp_exit_pxls_85_2050 <- spp_exit_pxls %>% 
  filter(scenario == "RCP85_2050") %>% 
  filter(MRGID %in% eez_200nm_list)
```

```{r}
# Plot EEZ entry pixels
plot_entry_10_map <- ggplot() +
  geom_sf(data = eez_simple_sf) +
  coord_sf(expand = FALSE) +
  geom_tile(data = spp_entry_pxls_85_2050, aes(x = x, y = y,
                                               fill = n_entry_spp,
                                               color = n_entry_spp)) +
  scale_fill_viridis_c(limits = c(1, 40), 
                      breaks = c(1, 2, 4, 8, 16, 28, 40),
                      direction = -1,
                      trans = "log", # visually stretches the lower value data
                      guide = guide_colorbar(
                      barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                      barheight = unit(0.4, "cm"),
                      ticks = TRUE,
                      ticks.colour = "white",
                      ticks.linewidth = 0.8
  )) +
  scale_color_viridis_c(limits = c(1,40), 
                        breaks = c(1, 2, 4, 8, 16, 28, 40),
                        direction = -1,
                        trans = "log", # visually stretches the lower value data
                        guide = "none")  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "species entry count",
       color = "species entry count",
       x = "",
       y = "") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt")) 
```

```{r}
plot_entry_10_density <- ggplot(spp_entry_pxls_85_2050, aes(x = y)) +
  geom_density(fill = "forestgreen", alpha = 0.5) +
  coord_flip() +
  scale_x_continuous(limits = c(-90, 90)) +
  theme_minimal() +
  labs(x = "",
       y = "") +
  theme(plot.margin = margin(0, 0, 0, 0))
```

```{r}
entry_combined_plot <- plot_entry_10_map + plot_entry_10_density + 
  plot_annotation(title = "Species entry, RCP 8.5 2050") +
  plot_layout(widths = c(4, 1),
              axis_titles = "collect",
              guides = "collect") &
  theme(legend.position = "bottom") 
```

```{r}
# Plot EEZ exit pixels
plot_exit_10_map <- ggplot() +
  geom_sf(data = eez_simple_sf) +
  coord_sf(expand = FALSE) +
  geom_tile(data = spp_exit_pxls_85_2050, aes(x = x, y = y,
                                              fill = n_exit_spp,
                                              color = n_exit_spp)) +
  scale_fill_viridis_c(limits = c(1,40), 
                       breaks = c(1, 2, 4, 8, 16, 28, 40),
                       direction = -1, 
                       trans = "log", # visually stretches the lower value data
                       guide = guide_colorbar(
                       barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                       barheight = unit(0.4, "cm"),
                       ticks = TRUE,
                       ticks.colour = "white",
                       ticks.linewidth = 0.8)) +
  scale_color_viridis_c(limits = c(1,40), 
                        breaks = c(1, 2, 4, 8, 16, 32, 40), 
                        direction = -1,
                        trans = "log", # visually stretches the lower value data
                        guide = "none")  +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() + 
  labs(fill = "species exit count",
       color = "species exit count",
       x = "",
       y = "") +
  theme(legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt"))
```

```{r}
plot_exit_10_density <- ggplot(spp_exit_pxls_85_2050, aes(x = y)) +
  geom_density(fill = "darkmagenta", alpha = 0.5) +
  coord_flip() +
  scale_x_continuous(limits = c(-90, 90)) +
  theme_minimal() +
  labs(x = "",
       y = "") +
  theme(plot.margin = margin(0, 0, 0, 0))
```

```{r}
exit_combined_plot <- plot_exit_10_map + plot_exit_10_density + 
  plot_annotation(title = "Species exit, RCP 8.5 2050") +
  plot_layout(widths = c(4, 1),
              axis_titles = "collect",
              guides = "collect") &
  theme(legend.position = "bottom") 
```

```{r}
# Save entry and exit maps
ggsave(here("figures/iucn_atrisk_spp_entry_10pxl_map_072926.png"), entry_combined_plot, width = 10, height = 5, dpi = 300)
ggsave(here("figures/iucn_atrisk_spp_exit_10pxl_map_072926.png"), exit_combined_plot, width = 10, height = 5, dpi = 300)
```

## Bivariate plot

Code created with assistance from Claude to categorize the raster entry and exit pixels into a bivariate color scheme using the following categories to create a 3x3 grid:
* 0 species entry and 0 species exit
* 0 species entry and 1 species exit
* 1 species entry and 0 species exit
* 1 species entry and 1 species exit
* 2+ species entry and 2+ species exit

```{r}
# Merge the entry and exit dataframes so that there are the same number of pixels in each raster
spp_entry_exit_85_2050 <- full_join(spp_entry_pxls_85_2050, spp_exit_pxls_85_2050) %>% 
  # Replace NA with 0
  replace_na(list(n_entry_spp = 0, n_exit_spp = 0))
```

```{r}
# Create a new column that categorizes the number of species entry and exit into 3 categories: 0, 1, and 2+
spp_entry_exit_cat_85_2050 <- spp_entry_exit_85_2050 %>% 
  # Assign numeric_category
  mutate(entry_cat = case_when(n_entry_spp == 0 ~ 1,
                               n_entry_spp == 1 ~ 2,
                               n_entry_spp >= 2 ~ 3),
         exit_cat = case_when(n_exit_spp == 0 ~ 1,
                              n_exit_spp == 1 ~ 2,
                              n_exit_spp >= 2 ~ 3)) %>% 
  # Create a new column that combines the entry and exit categories into a single category
  mutate(entry_exit_cat = (exit_cat - 1) * 3 + entry_cat)
```

```{r}
# Define numeric categories for the bivariate color scheme
bivar_levels <- data.frame(
  entry_exit_cat    = 1:9,
  label = c(
    "En0_Ex0", "En1_Ex0", "En2_Ex0",   # exits = 0
    "En0_Ex1", "En1_Ex1", "En2_Ex1",   # exits = 1
    "En0_Ex2", "En1_Ex2", "En2_Ex2"    # exits = 2+
  )
)
spp_entry_exit_cat_85_2050 <- left_join(spp_entry_exit_cat_85_2050, bivar_levels)
```

```{r}
# Color palette
# Entries increase left → right (blue axis)
# Exits increase bottom → top (pink/purple axis)
bivar_colors <- c(
  "En0_Ex0" = "#e8e8e8",  # entries=0,   exits=0
  "En1_Ex0" = "#ace4e4",  # entries=1, exits=0
  "En2_Ex0" = "#5ac8c8",  # entries=2+,  exits=0
  "En0_Ex1" = "#dfb0d6",  # entries=0,   exits=1
  "En1_Ex1" = "#a5add3",  # entries=1, exits=1
  "En2_Ex1" = "#5698b9",  # entries=2+,  exits=1
  "En0_Ex2" = "#be64ac",  # entries=0,   exits=2+
  "En1_Ex2" = "#8c62aa",  # entries=1, exits=2+
  "En2_Ex2" = "#3b4994"   # entries=2+,  exits=2+
)
```

```{r}
biplot <- ggplot() +
  geom_sf(data = eez_simple_sf, fill = "#e8e8e8") +
  coord_sf(expand = FALSE) +
  geom_tile(data = spp_entry_exit_cat_85_2050, aes(x = x, y = y,
                                              fill = label,
                                              color = label)) +
  scale_fill_manual(values = bivar_colors, na.value = "white") +
  scale_color_manual(values = bivar_colors, na.value = "white") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Species exits vs entries, RCP 8.5 2050",
       x = "",
       y = "")
```

```{r}
# Create legend for bivariate color scheme
legend_df <- expand.grid(entries = 1:3, exits = 1:3)
legend_df$label <- paste0("En", legend_df$entries - 1, "_Ex", legend_df$exits - 1)

legend_biplot <- ggplot(legend_df, aes(x = entries, y = exits, fill = label)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_manual(values = bivar_colors) +
  scale_x_continuous(breaks = 1:3, labels = c("0", "1", "2+")) +
  scale_y_continuous(breaks = 1:3, labels = c("0", "1", "2+")) +
  labs(x = "Entries →", y = "Exits →") +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text       = element_text(size = 8),
    axis.title      = element_text(size = 9)
  )
```

```{r}
# Combine plot
bivariate_map <- ggdraw() +
  draw_plot(biplot,    0,    0,    1,    1) +
  draw_plot(legend_biplot, 0.75, 0.75, 0.2, 0.25)
```

# Add density plot to bivariate map

```{r}
biplot_density <-ggplot() +
  geom_density(data = spp_exit_pxls_85_2050, aes(x = y),fill = "#be64ac", color = "#be64ac", alpha = 0.5, size = 1) +
  geom_density(data = spp_entry_pxls_85_2050, aes(x = y),fill = "#5698b9", color = "#5698b9", alpha = 0.5, size = 1) +
  coord_flip() +
  scale_x_continuous(limits = c(-90, 90)) +
  theme_minimal() +
  labs(x = "",
       y = "") +
  theme(plot.margin = margin(0, 0, 0, 0))
```

```{r}
biplot_2 <- ggplot() +
  geom_sf(data = eez_simple_sf, fill = "#e8e8e8") +
  coord_sf(expand = FALSE) +
  geom_tile(data = spp_entry_exit_cat_85_2050, aes(x = x, y = y,
                                              fill = label,
                                              color = label)) +
  scale_fill_manual(values = bivar_colors, na.value = "white") +
  scale_color_manual(values = bivar_colors, na.value = "white") +
  scale_y_continuous(limits = c(-90, 90)) +
  theme_minimal() +
  theme(legend.position = "none",
        plot.margin = margin(0, 0, 0, 0),
        panel.spacing = unit(0, "pt")) +
  labs(title = "",
       x = "",
       y = "") 
```

```{r}
biplot_and_density <- biplot_2 + biplot_density + 
  plot_annotation(title = "Species exit vs entry, RCP 8.5 2050") +
  plot_layout(widths = c(4, 1),
              axis_titles = "collect",
              guides = "collect") &
  theme(legend.position = "none") 
```

```{r}
biplot_all <- ggdraw() +
  draw_plot(biplot_and_density,    0,    0,    1,    1) +
  draw_plot(legend_biplot, 0.6, 0.8, 0.2, 0.2)
```

```{r}
# Save entry and exit maps
ggsave(here("figures/iucn_atrisk_spp_entry_exit_density_10pxl_map_073026.png"), biplot_all, width = 10, height = 5, dpi = 300)
```

