---
title: "Plots of Threatened Species Ranges & EEZs"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
library(arrow)
```

This document creates plots for the data calculated in 01_iucn_aquax_eez_overlap.qmd and 02_iucn_aquax_rangearea.qmd.

## Read in threatened species data

```{r}
# Spp in EEZs
spp_eez_pxl_count <- read_csv(here("data/iucn_threatened_spp_eez_pxls_061626.csv"))
```

```{r}
# Spp EEZ entry/exit pixels
spp_entry_pxls <- read_rds(here("data/iucn_threatened_10pxlthres_entry_pxls_061726.rds"))
spp_exit_pxls <- read_rds(here("data/iucn_threatened_10pxlthres_exit_pxls_061726.rds"))
```

```{r}
# Species range area
area_df <- read_csv(here("data/iucn_threatened_spp_range_area_061826.csv"))
```

## Read in EEZ shapefile
```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')

eez_simple <- eez %>% 
  dplyr::select(MRGID, GEONAME)
```

```{r}
# Create a blank raster matching AquaX resolution (EEZs should be large enough to just do 0.05 degree raster)
eez_raster_blank <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05)
```

```{r}
eez_raster_geoname <- rasterize(eez, eez_raster_blank, field = "GEONAME")
```

## Plot species pixels

```{r}
# Convert to equal area projection for plotting
eez_eq <- st_transform(eez, crs = "EPSG:8857")
```

```{r}
spp_entry_85_2050 <- spp_entry_pxls %>% 
    filter(scenario == "RCP85_2050")

spp_exit_85_2050 <- spp_exit_pxls %>% 
    filter(scenario == "RCP85_2050")
```

```{r}
# Convert to equal area projection for plotting
spp_entry_85_2050_r <- rast(spp_entry_85_2050, crs = "EPSG:4326") %>% 
  project("EPSG:8857", method = "near")

spp_exit_85_2050_r <- rast(spp_exit_85_2050, crs = "EPSG:4326") %>% 
  project("EPSG:8857", method = "near")
```

```{r}
# Back to df for ggplot
spp_entry_85_2050df_eq <- as.data.frame(spp_entry_85_2050_r , xy = TRUE) 
spp_exit_85_2050df_eq <- as.data.frame(spp_exit_85_2050_r , xy = TRUE) 
```

```{r}
# Plot EEZ entry pixels
ggplot() +
  geom_sf(data = eez_eq) +
  geom_tile(data = spp_entry_85_2050df_eq, aes(x = x, y = y,
                                               fill = n_entry_spp,
                                               color = n_entry_spp)) +
  scale_fill_viridis_c(limits = c(1,21), direction = -1) +
  scale_color_viridis_c(limits = c(1,21), direction = -1) +
  theme_minimal()
```


```{r}
# Plot EEZ exit pixels
ggplot() +
  geom_sf(data = eez_eq) +
  geom_tile(data = spp_exit_85_2050df_eq, aes(x = x, y = y,
                                               fill = n_exit_spp,
                                               color = n_exit_spp)) +
  scale_fill_viridis_c(limits = c(1,21), direction = -1) +
  scale_color_viridis_c(limits = c(1,21), direction = -1) +
  theme_minimal()
```

## Plot some areas with high numbers of entries/exits
```{r}
ggplot() +
  geom_sf(data = eez_eq %>% filter(MRGID %in% c("8328", "48955", "48950"))) +
  geom_tile(data = spp_entry_85_2050df_eq %>% 
            filter(MRGID %in% c("8328", "48955", "48950")), 
            aes(x = x, y = y,
            fill = n_entry_spp,
            color = n_entry_spp)) +
  scale_fill_viridis_c(limits = c(1,21), direction = -1) +
  scale_color_viridis_c(limits = c(1,21), direction = -1) +
  theme_minimal() +
  labs(title = "EEZs with most threatened species entries under RCP8.5 2050",
       subtitle = "Top 3 EEZs: North Korea, Lincourt Rocks, and Kuril Islands")
```


```{r}
ggplot() +
  geom_sf(data = eez_eq %>% filter(MRGID %in% c("64440", "8485"))) +
  geom_tile(data = spp_exit_85_2050df_eq %>% 
            filter(MRGID %in% c("64440", "8485")), 
            aes(x = x, y = y,
            fill = n_exit_spp,
            color = n_exit_spp)) +
  scale_fill_viridis_c(limits = c(1,21), direction = -1) +
  scale_color_viridis_c(limits = c(1,21), direction = -1) +
  theme_minimal() +
  labs(title = "EEZs with most threatened species exits under RCP8.5 2050",
       subtitle = "Top 2 EEZs: Singapore / Malaysia, Singapore")
```

## Other plots 

```{r}
# How many EEZs are each spp present in for current and future?
eez_spp_counts <- spp_eez_pxl_count %>%  
  group_by(aphiaid) %>% 
  summarize(current_eez_count = sum(n_current >= 10, na.rm = TRUE),
            RCP45_2050_eez_count = sum(n_RCP45_2050 >= 10, na.rm = TRUE),
            RCP45_2100_eez_count = sum(n_RCP45_2100  >= 10, na.rm = TRUE),
            RCP85_2050_eez_count = sum(n_RCP85_2050  >= 10, na.rm = TRUE),
            RCP85_2100_eez_count = sum(n_RCP85_2100  >= 10, na.rm = TRUE)) %>% 
  ungroup() 
```

```{r}
ggplot() +
  geom_point(data = eez_spp_counts, aes(x = current_eez_count, y = RCP85_2050_eez_count),
             alpha = 0.5, color = "maroon") +
  #geom_point(data = eez_spp_counts, aes(x = current_eez_count, y = RCP85_2100_eez_count),
             #alpha = 0.5, color = "forestgreen") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  labs(title = "IUCN Threatened Species EEZ count (RCP 8.5 2050)",
       y = "RCP 85 future EEZ count")
```


```{r}
# Organize dataframe so that's it's easier to select a scenario and compare to current
eez_spp_pxls_long <- spp_eez_pxl_count %>% 
  pivot_longer(cols = starts_with("n_RCP"),
               names_to = "scenario",
               values_to = "pxl_count") %>% 
  mutate(scenario = str_remove(scenario, "n_")) %>% 
  # Replace NA with 0, otherwise will be assigned to "continued presence"
  replace_na(list(pxl_count = 0, n_current = 0)) %>% 
  mutate(spp_change = case_when(n_current < 10 &  pxl_count >= 10 ~ 'entry into new',
                            n_current >= 10  & pxl_count < 10 ~ 'exit from historical',
                            n_current < 10  & pxl_count < 10 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

```{r}
# Get counts of entry into new, exit from historical etc. by EEZ
eez_spp_change <- eez_spp_pxls_long %>% 
  group_by(GEONAME, scenario, spp_change) %>% 
  summarize(spp_change_count = n()) %>% 
  ungroup()
```

```{r}
eez_c_f <- eez_spp_change %>% 
  mutate(spp_change = str_replace_all(spp_change, " ", "_")) %>% 
  pivot_wider(names_from = spp_change, values_from = spp_change_count) %>% 
  replace_na(list(continued_presence = 0, exit_from_historical = 0,
                  entry_into_new = 0, no_presence = 0)) %>% 
  mutate(current_spp = continued_presence + exit_from_historical,
         future_spp = continued_presence + entry_into_new,
         perc_exit = 100 * exit_from_historical/current_spp,
         perc_enter = 100 * entry_into_new/future_spp,
         spp_change = future_spp - current_spp,
         ssi = (2 * continued_presence)/(current_spp + future_spp)) 
```

```{r}
# EEZ centroids
eez_simple_valid <- st_make_valid(eez_simple)
eez_centroid <- st_centroid(eez_simple_valid)  %>%
  cbind(st_coordinates(.)) %>%
  st_drop_geometry()
```

```{r}
# Merge in centroids
eez_c_f_y <- left_join(eez_c_f, eez_centroid)
```


```{r}
ggplot() +
  geom_point(data = eez_c_f_y %>% filter(scenario == "RCP85_2050"),
   aes(color = Y, x = current_spp, y = future_spp), alpha = 0.75, 
             position = position_jitterdodge()) +
  #scale_color_viridis_c() +
  scale_color_distiller(palette = "PRGn", direction = 1, limits = c(-80, 80)) + 
  scale_fill_distiller(palette = "PRGn", direction = 1, limits = c(-80, 80)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(color = "latitude",
       title = "Current and future spp by EEZ",
       y = "future species count",
       x = "current species count") +
  theme_bw() +
  facet_wrap(~scenario)
```

```{r}
ggplot() +
  geom_point(data = eez_c_f_y %>% filter(scenario == "RCP85_2050"),
             aes(color = Y, x = exit_from_historical, y = entry_into_new), alpha = 0.8, 
             position = position_jitterdodge()) +
  #scale_color_viridis_c() +
  scale_color_distiller(palette = "PRGn", direction = 1, limits = c(-80, 80)) + 
  scale_fill_distiller(palette = "PRGn", direction = 1, limits = c(-80, 80)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(#color = "latitude",
       title = "Threatened species loss and gain by EEZ",
       y = "species gain count",
       x = "species loss count") +
  theme_bw() +
  facet_wrap(~scenario)
```

```{r}
ggplot() +
  geom_point(data = left_join(eez_c_f, eez_centroid) %>% filter(scenario == "RCP85_2050"),
             aes(color = Y, x = perc_exit, y = perc_enter), alpha = 0.8, 
             position = position_jitterdodge()) +
  #scale_color_viridis_c() +
  scale_color_distiller(palette = "PRGn", direction = 1, limits = c(-80, 80)) + 
  scale_fill_distiller(palette = "PRGn", direction = 1, limits = c(-80, 80)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray30") +
  labs(#color = "latitude",
       title = "Threatened species loss and gain by EEZ",
       y = "species gain %",
       x = "species loss %") +
  #scale_y_continuous(limits = c(1, 75)) +
  #scale_x_continuous(limits = c(1, 75)) +
  theme_bw() +
  facet_wrap(~scenario)
```

```{r}
# Add sf 
eez_c_f_sf <- left_join(eez_c_f, eez_simple)
```

```{r}
# Threatened species change
ggplot() +
  geom_sf(data = eez_c_f_sf, aes(geometry = geom, fill = spp_change)) +
  scale_color_distiller(palette = "BrBG", direction = 1, limits = c(-75, 75)) + 
  scale_fill_distiller(palette = "BrBG", direction = 1, limits = c(-75, 75)) +
  theme_bw() +
  labs(title = "Threatened species change by EEZ") +
  facet_wrap(~scenario)
```

```{r}
# Threatened species percentage - need to drop overlapping claims
ggplot() +
  geom_sf(data = eez_c_f_sf, aes(geometry = geom, fill = perc_exit)) +
  theme_bw() +
  labs(title = "Threatened species change by EEZ") +
  facet_wrap(~scenario)
```

## Test biplot

```{r}
# Merge the entry and exit dataframes
entry_exit_spp_count_pxl <- full_join(spp_entry_pxls, spp_exit_pxls) 
```

```{r}
entry_exit_spp_sqrt <- entry_exit_spp_count_pxl %>% 
    filter(scenario == "RCP85_2100") %>%
  replace_na(list(n_entry_spp = 0)) %>% 
  replace_na(list(n_exit_spp = 0)) %>% 
  mutate(n_entry_spp_sqrt = sqrt(n_entry_spp),
         n_exit_spp_sqrt = sqrt(n_exit_spp))
```


```{r}
library(biscale)
library(cowplot)
```

```{r}
biplot_data <- bi_class(entry_exit_spp_sqrt, x = n_entry_spp_sqrt, y = n_exit_spp_sqrt,
                        style = "equal", dim = 3)
```

```{r}
biplot_85_2100 <- ggplot() + 
  geom_tile(data = biplot_data, mapping = aes(fill = bi_class,
                                              color = bi_class,
                                              x = x, y = y), show.legend = FALSE) +
  bi_scale_fill(pal = "DkBlue2", dim = 3, rotate_pal = FALSE) +
  bi_scale_color(pal = "DkBlue2", dim = 3, rotate_pal = FALSE) +
  geom_sf(data = eez, fill = NA) +
  labs(x = "",
       y = "",
       subtitle = "RCP 8.5 2100 spp change") +
  bi_theme()
```

```{r}
legend <- bi_legend(pal = "DkBlue2",
                    dim = 3,
                    xlab = "More new spp ",
                    ylab = "More loss spp",
                    size = 8)
```

```{r}
final_plot_85_2100 <- ggdraw() +
  draw_plot(biplot_85_2100, 0, 0, 1, 1) +
  draw_plot(legend, .8, .8, 0.2, 0.2)
```

```{r}
final_plot_85_2100
```

## Plots by taxonomic group

```{r}
iucn_lookup <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/_data_raw/iucn_redlist_2026/iucn_worms_aphia_id_complete.csv")
```

```{r}
iucn_cat <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/_data_raw/iucn_redlist_2026/iucn_species_assessments.csv")
```

```{r}
# Add aphiaid to iucn_cat
iucn_cat_aphiaid <- left_join(iucn_cat, (iucn_lookup %>% dplyr::select(iucn_assessment_id, aphia_id)),
 by = c("assessment_id" = "iucn_assessment_id")) %>% 
  rename(aphiaid = aphia_id) 
```

```{r}
# Merge in iucn_cat to get group for each species
spp_eez_pxl_count_group <- left_join(spp_eez_pxl_count, iucn_cat_aphiaid)
```

```{r}
# Get count of number of EEZ by group and count of IUCN cat by group (create figure similar to bar chart)
group_eez_count <- spp_eez_pxl_count_group %>% 
  group_by(group, aphiaid, redlist_category) %>% 
  summarize(current_eez_count = sum(n_current >= 10, na.rm = TRUE),
            RCP45_2050_eez_count = sum(n_RCP45_2050 >= 10, na.rm = TRUE),
            RCP45_2100_eez_count = sum(n_RCP45_2100 >= 10, na.rm = TRUE),
            RCP85_2050_eez_count = sum(n_RCP85_2050 >= 10, na.rm = TRUE),
            RCP85_2100_eez_count = sum(n_RCP85_2100 >= 10, na.rm = TRUE)) %>% 
  ungroup()

group_iucncat_count <- group_eez_count %>% 
  group_by(group, redlist_category) %>% 
  summarize(redlist_count = n()) %>% 
  ungroup()
```


```{r}
ggplot() +
  geom_density(data = group_eez_count, aes(x = current_eez_count, fill = group), alpha = 0.5) +
  labs(title = "Distribution of current EEZ counts by taxonomic group",
       x = "current EEZ count") +
  facet_wrap(~group) +
  theme_bw()
```

```{r}
ggplot() +
  geom_density(data = group_eez_count, aes(x = RCP85_2100_eez_count, fill = group), alpha = 0.5) +
  labs(title = "Distribution of RCP85_2100 EEZ counts by taxonomic group",
       x = "RCP85_2100 EEZ count") +
  facet_wrap(~group) +
  theme_bw()
```


```{r}
# Assign EEZ count thresholds based on plotted density (drop not comprehensively assessed)
group_eez_count_threshold <- group_eez_count %>% 
  filter(group != "not comprehensively assessed") %>%
  mutate(current_eez_count_cat = case_when(current_eez_count == 1 ~ "1",
                                          current_eez_count == 2 ~ "2",
                                           current_eez_count > 2 & current_eez_count <= 10 ~ "3-10",
                                           current_eez_count > 10 & current_eez_count <= 25 ~ "11-25",
                                           current_eez_count > 25 & current_eez_count <= 100 ~ "26-100",
                                           current_eez_count > 100 ~ ">100"),
         RCP85_2050_eez_count_cat = case_when(RCP85_2050_eez_count == 1 ~ "1",
                                            RCP85_2050_eez_count == 2 ~ "2",
                                           RCP85_2050_eez_count > 2 & RCP85_2050_eez_count <= 10 ~ "3-10",
                                           RCP85_2050_eez_count > 10 & RCP85_2050_eez_count <= 25 ~ "11-25",
                                           RCP85_2050_eez_count > 25 & RCP85_2050_eez_count <= 100 ~ "26-100",
                                           RCP85_2050_eez_count > 100 ~ ">100")) %>% 

  mutate(general_group = case_when(group %in% c("hagfishes", "sturgeons", "tunas_and_billfishes", "tarpons_and_ladyfishes")~ "pelagic_fishes",
                                   group %in% c("angelfishes", "blennies", "butterfly_fishes", "groupers", "pufferfishes", "seabreams_porgies_picarels", "surgeonfishes", "wrasses_and_parrotfishes") ~ "reef_fishes",
                                   group %in% c("marine_turtles", "seasnakes") ~ "marine_turtles_and_seasnakes",
                                   TRUE ~ group))
```

```{r}
cat_summary <- group_eez_count_threshold %>% 
  group_by(group, current_eez_count_cat) %>% 
  summarise(n_cat = n()) %>% 
  ungroup()

# Get total number in each group
group_count <- group_eez_count_threshold %>% 
  group_by(group) %>% 
  summarise(n_group = n()) %>% 
  ungroup()

# Merge in group count to get percentage
cat_summary_perc <- left_join(cat_summary, group_count) %>% 
  mutate(perc_cat = n_cat/n_group * 100)
```

```{r}
group_eez_count_threshold$current_eez_count_cat <- factor(group_eez_count_threshold$current_eez_count_cat, 
levels = c("1", "2", "3-10", "11-25", "26-100", ">100"))

group_eez_count_threshold$RCP85_2050_eez_count_cat <- factor(group_eez_count_threshold$RCP85_2050_eez_count_cat, 
levels = c("1", "2", "3-10", "11-25", "26-100", ">100"))
```

```{r}
library(forcats)
# Plot a bar graph by group with fill as the EEZ count category
ggplot() +
  geom_bar(data = group_eez_count_threshold, aes(y = fct_rev(general_group), fill = fct_rev(current_eez_count_cat)), position = "fill") +
  scale_fill_viridis_d() +
  labs(title = "Current EEZ count category by IUCN group",
       y = "",
       x = "proportion",
       fill = "EEZ count") +
  theme_bw()
```

```{r}
# Plot a bar graph by group with fill as the EEZ count category
ggplot() +
  geom_bar(data = group_eez_count_threshold, aes(y = fct_rev(general_group), fill = fct_rev(RCP85_2050_eez_count_cat)), position = "fill") +
  scale_fill_viridis_d() +
  labs(title = "RCP 8.5 2050 EEZ count category by IUCN group",
       y = "",
       x = "proportion",
       fill = "EEZ count") +
  theme_bw()
```

## Set up dataframe so that it's easier to plot current vs future side by side for each group 
```{r}
eez_count_group_simple <- group_eez_count_threshold %>% 
  dplyr::select(general_group, current_eez_count_cat, RCP85_2050_eez_count_cat) %>% 
  pivot_longer(cols = c(current_eez_count_cat, RCP85_2050_eez_count_cat), names_to = "scenario", values_to = "eez_count_cat") %>% 
  mutate(scenario = str_replace(scenario, "current_eez_count_cat", "current"),
         scenario = str_replace(scenario, "RCP85_2050_eez_count_cat", "RCP85_2050"))
```

```{r}
# Plot a bar graph by group with fill as the EEZ count category
ggplot() +
  geom_bar(data = eez_count_group_simple, aes(y = fct_rev(general_group), fill = fct_rev(eez_count_cat)), position = "fill") +
  scale_fill_viridis_d() +
  labs(title = "Threatened species EEZ count by IUCN comprehensive group",
       y = "",
       x = "proportion",
       fill = "EEZ count") +
  theme_bw() + 
  theme(legend.position = "bottom") +
  facet_wrap(~scenario)
```

## Create jitter plot of the difference btwn RCP85_2050 and current EEZ count by group 

```{r}
group_eez_count_diff <- group_eez_count_threshold %>% 
  mutate(eez_count_diff = RCP85_2050_eez_count - current_eez_count) %>% 
  dplyr::select(group, general_group, aphiaid, current_eez_count,RCP85_2050_eez_count, eez_count_diff)
```

```{r}
ggplot() +
  geom_jitter(data = group_eez_count_diff, aes(x = eez_count_diff, y = fct_rev(general_group), color = general_group),
   alpha = 0.6, position = position_jitterdodge(dodge.width = 1, jitter.width = .25),) +
  labs(title = "EEZ counts change by IUCN comprehensive group",
       x = "EEZ count change (RCP85 2050 - current)",
       y = "") +
  theme_bw() +
  theme(legend.position = "none") 
```

## IUCN threats
For those species experiencing range shifts, what are the most common IUCN threats? Are there particular threats that are more common for species experiencing range loss vs gain?
https://www.iucnredlist.org/resources/threat-classification-scheme

According to the IUCN, threats marked with timing as “ongoing”, severity as “rapid” or “very rapid” declines and scopes of “majority” or “whole” are typically considered major threats to a species. IUCN stopped using scores in 2022 due to lack of sceintific method.

```{r}
library(iucnredlist) ### remotes::install_github("IUCN-UK/iucnredlist")
```

```{r}
api_key <- Sys.getenv("IUCN_API_KEY") ### usethis::edit_r_environ() to add key to R environment file
api <- init_api(api_key)
```

```{r}
iucn_cat_threat <- iucn_cat %>% 
  filter(redlist_category %in% c("CR", "EN", "VU"))
```

```{r}
# Get IUCN ID data 
iucn_id_data <- iucn_cat_threat$assessment_id
```

```{r}
# still need to run this through assesment_data_many
iucn_id_data <- assessment_data_many(api, iucn_id_data[1:5], wait_time = 0.5)
```

```{r}
# run element_names() to see what categories can be extracted
iucn_threats <- extract_element(iucn_id_data, "threats")
```

Threat information is pretty complicated, so maybe will focus on a subset of species for now. Perhaps exiting a certain EEZ. 

```{r}
# Singapore / Malaysia & Malaysia EEZs has the most threatened species exits under RCP8.5 2050 - get list of species exiting this EEZ
spp_exit_eez_top <- spp_eez_pxl_count_group %>% 
  filter(MRGID %in% c("64440", "8485")) %>%
  filter(is.na(n_RCP85_2050) & n_current > 0)
```

```{r}
# Get IUCN IDs for these species
iucn_ids_exit_eez_top <- unique(spp_exit_eez_top$assessment_id)
```

```{r}
# still need to run this through assesment_data_many
iucn_id_data_eez_top <- assessment_data_many(api, iucn_ids_exit_eez_top, wait_time = 0.5)
```

```{r}
# run element_names() to see what categories can be extracted
iucn_threats_eez_top <- extract_element(iucn_id_data_eez_top, "threats") %>% 
  mutate(assessment_id = as.numeric(assessment_id))
```

```{r}
# Merge in the threat data to the spp_exit_eez_top dataframe to see if there are particular threats that are more common for species exiting this EEZ
spp_exit_eez_top_threats <- inner_join(spp_exit_eez_top, iucn_threats_eez_top, by = c("assessment_id" = "assessment_id"))
```


```{r}
spp_exit_eez_top_threats_summary <- spp_exit_eez_top_threats %>% 
  group_by(MRGID, GEONAME, code, description) %>% 
  summarize(n_spp = n()) %>% 
  ungroup() %>% 
  arrange(desc(n_spp))

```

```{r}
# Try filtering for only severe threats (scope of majority or whole, severity of rapid or very rapid, timing of ongoing)
# 7 of the 75 species exiting the Singapore/Malaysia EEZ under RCP8.5 2050 are experiencing severe threats according to IUCN criteria
# Threats are all connected to fishing and harvest, along with shipping lanes
spp_exit_eez_top_severe_threats_summary <- spp_exit_eez_top_threats %>% 
  filter(scope.y %in% c("Majority (50-90%)", "Whole (>90%)") &
         severity %in% c("Rapid Declines", "Very Rapid Declines") &
         timing == "Ongoing") %>%
  group_by(MRGID, GEONAME, code, description) %>% 
  summarize(n_spp = n()) %>% 
  ungroup() %>% 
  arrange(desc(n_spp))

```


```{r}
# What about all the threatened species leaving EEZs under RCP8.5 2050
spp_exit_eez <- spp_eez_pxl_count_group %>% 
  filter(is.na(n_RCP85_2050) & n_current > 0)
```

```{r}
# Get IUCN IDs for these species
iucn_ids_exit_eez <- unique(spp_exit_eez$assessment_id)
```

```{r}
# still need to run this through assesment_data_many
iucn_id_data_eez <- assessment_data_many(api, iucn_ids_exit_eez, wait_time = 0.5)
```

```{r}
# run element_names() to see what categories can be extracted
iucn_threats_eez <- extract_element(iucn_id_data_eez, "threats") %>% 
  mutate(assessment_id = as.numeric(assessment_id))
```

```{r}
# Merge in the threat data to the spp_exit_eez_top dataframe to see if there are particular threats that are more common for species exiting this EEZ
spp_exit_eez_threats <- inner_join(spp_exit_eez, iucn_threats_eez, by = c("assessment_id" = "assessment_id"))
```

```{r}
# Filter for the severe threats (33 threatened species experiencing severe threats)
spp_exit_eez_severe_threats <- spp_exit_eez_threats %>% 
  filter(scope.y %in% c("Majority (50-90%)", "Whole (>90%)") &
         severity %in% c("Rapid Declines", "Very Rapid Declines") &
         timing == "Ongoing")  

spp_exit_eez_severe_threats_summary <- spp_exit_eez_severe_threats %>% 
  group_by(MRGID, GEONAME, code, description) %>% 
  summarize(n_spp = n()) %>% 
  ungroup() %>% 
  arrange(desc(n_spp))
```


```{r}
# IDs for severe threat species that are leaving EEZS under RCP8.5 2050
iucn_ids_exit_eez_severe_threats <- unique(spp_exit_eez_severe_threats$aphiaid)
```