---
title: "AquaX species pixels entering EEZs"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(terra)
library(sf)
library(here)
library(janitor)
library(arrow)
```

This script prepares the data for the count of pixels in EEZs for threatened IUCN redlist species.

## Load functions

```{r}
source(here("common_fxns.R"))
```

## EEZ 

```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
# Create a blank raster matching AquaX resolution (EEZs should be large enough to just do 0.05 degree raster)
eez_raster_blank <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05)
```

```{r}
eez_raster_geoname <- rasterize(eez, eez_raster_blank, field = "GEONAME")
```

```{r}
eez_raster_mrgid <- rasterize(eez, eez_raster_blank, field = "MRGID")
```

```{r}
eez_simple <- eez %>% 
  dplyr::select(MRGID, GEONAME)
```


## Prep AquaX redlist data

```{r}
# read in meta data on the aquax sample of redlist species
load("/home/shares/data-aquax/iucn_request/meta.Rdata")
```

```{r}
# IUCN threat info
iucn <- read.csv("/home/shares/usgs-cap-rangeshifts/data_iucn/iucn_to_worms_match_redlist_category.csv")
```

```{r}
meta_clean <- meta %>% 
  mutate(worms_name = tolower(ScientificName)) %>% 
  select(!ScientificName) %>% 
  # simplify for now for easier viewing
  select(AphiaID, SPID, worms_name, GROUP, NR_CR, range_pixels, SDM)
```

```{r}
# Merge in IUCN info based on the worms_name
meta_iucn <- left_join(meta_clean, 
                       # merge in IUCN data and filter for species (no subpops)
                       iucn %>%  filter(species == TRUE),
                       by = "worms_name")
```

```{r}
# Drop those without IUCN sid (get 13,132 species remaining)
meta_iucn_no_na <- meta_iucn %>% 
  filter(!is.na(iucn_sid))
```

```{r}
# May be duplicates (90 cases where AquaX/worms is two or more species in IUCN)
# Might just remove these for now
df_worms_iucn_dupes <- meta_iucn_no_na %>% 
  get_dupes(worms_name)

dupes <- unique(df_worms_iucn_dupes$worms_name)
```

```{r}
# Drop those 90 worms species that are representing 2+ IUCN species 
# 12,940 species remaining
meta_iucn_no_na_no_dupes <- meta_iucn_no_na %>% 
  filter(!worms_name %in% dupes)
```

```{r}
# For now, focusing on only threatened species (1,013 species)
meta_iucn_threatened <- meta_iucn_no_na_no_dupes %>% 
  filter(iucn_category %in% c("Critically Endangered",
                              "Endangered", "Vulnerable"))
```

```{r}
# Get AphiaID of threatened species
aphiaid_threatened <- meta_iucn_threatened$AphiaID
```

## Prep AquaX data
Here, there are just the 12,238 files -not totally sure why the files and meta data don't exactly line up. But when I filter for the 1,013 threatened AphiaIDs, I get 906 species.

```{r}
spp_fs <- list.files('/home/shares/data-aquax/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(meta, by = c('aphiaid' = 'AphiaID'))
```

```{r}
# Filter for the AphiaID of the threatend species (906)
df_threatened <- df %>% 
  filter(aphiaid %in% aphiaid_threatened)
```

## Test out subsample

```{r}
set.seed(123)
test_spp_30 <- slice_sample(df_threatened, n = 30)
files_30 <- test_spp_30$f
aphiaid_30 <- test_spp_30$aphiaid
```

```{r}
test_f <- read_parquet("/home/shares/data-aquax/SDM/FINAL_EMSDM_EMMEAN_SP_1006832.parquet")
```

```{r}
test_r <- rast(test_f, crs = "EPSG:4326") %>%
    extend(ext(eez_raster_mrgid))

test_binary <- ifel(test_r[["Current"]]    >= 712, 1, NA)
```

```{r}
current_counts <- zonal(test_binary, eez_raster_mrgid, fun = "sum", na.rm = TRUE)
```

## Test out filtering method

```{r}
eez_spp_pxls <- read_csv(here("data/iucn_threatened_spp_eez_pxls_iucncat_050826.csv"))
```

```{r}
# How many EEZs are each spp present in for current and future?
eez_spp_counts <- eez_spp_pxls %>% 
  group_by(aphiaid, iucn_category_code) %>% 
  summarize(current_eez_count = sum(n_current > 0, na.rm = TRUE),
            RCP45_2050_eez_count = sum(n_RCP45_2050 >0, na.rm = TRUE),
            RCP45_2100_eez_count = sum(n_RCP45_2100 > 0, na.rm = TRUE),
            RCP85_2050_eez_count = sum(n_RCP85_2050 > 0, na.rm = TRUE),
            RCP85_2100_eez_count = sum(n_RCP85_2100 > 0, na.rm = TRUE)) %>% 
  ungroup() 

```

```{r}
# Organize dataframe so that's it's easier to select a scenario and compare to current
eez_spp_pxls_long <- eez_spp_pxls %>% 
  pivot_longer(cols = starts_with("n_RCP"),
               names_to = "scenario",
               values_to = "pxl_count") %>% 
  mutate(scenario = str_remove(scenario, "n_")) %>% 
  mutate(spp_change = case_when(is.na(n_current) &  pxl_count > 0 ~ 'entry into new',
                            n_current > 0  & is.na(pxl_count) ~ 'exit from historical',
                            is.na(n_current)  & is.na(pxl_count) == 0 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

```{r}
# Get counts of entry into new, exit from historical etc. by EEZ
eez_spp_entry <- eez_spp_pxls_long %>% 
  filter(spp_change == "entry into new")
```

```{r}
# Create dataframe of EEZ raster
eez_raster_df <- as.data.frame(eez_raster_geoname, xy = TRUE)
```

```{r}
# Join to the test df
test_f_eez <- inner_join(test_f, eez_raster_df)

test_f_us <- test_f_eez %>% 
  filter(GEONAME == "United States Exclusive Economic Zone") %>% 
  filter(RCP45_2050 >= cutoff) %>% 
  mutate(bi_RCP45_2050 = 1)
```

```{r}
test_r_us_r <- rast(test_f_us,  crs = "EPSG:4326")
```

```{r}
plot(test_r_us_r$bi_RCP45_2050)
```


```{r}
extract_eez_entry_pixels_df <- function(id, df, s) {
  spp_filter <- df %>% filter(aphiaid == id)
  
  if (nrow(spp_filter) == 0) {
    warning("No rows found for aphiaid: ", id)
    return(NULL)
  }
  
  spp <- read_parquet(spp_filter$f) %>%
    mutate(aphiaid = id) %>% 
    dplyr::select(x, y, Current, all_of(s), cutoff)
  
  # Filter the df of entry into for this species & scenario
  entry_spp_df <- eez_spp_entry %>% 
    filter(aphiaid == id) %>% 
    filter(scenario == s)
  
  # Get the EEZs for which the species & scenario is entry
  spp_entry_eezs <- unique(entry_spp_df$GEONAME)
  
  # Join the eez dataframe and filter for the spp_entry_eezs
  spp_eez <- inner_join(spp, eez_raster_df) %>% 
    filter(GEONAME %in% spp_entry_eezs) %>% 
    filter(.data[[s]] >= cutoff) %>% 
    mutate(binary = 1) %>% 
    bind_rows()
  
  return(spp_eez)
}
```

```{r}
system.time({
  eez_list_count_df <- parallel::mclapply(aphiaid_30, FUN = extract_eez_entry_pixels_df, 
                                       df = test_spp_30, 
                                       s = "RCP45_2050",
                                       mc.cores = 10)
})
```

```{r}
eez_count_df <- bind_rows(eez_list_count_df)
```

```{r}
# Sum by XY cell, create raster
eez_sum_df <- eez_count_df %>% 
  mutate(scenario = "RCP45_2050") %>% 
  group_by(x, y, GEONAME,scenario) %>% 
  summarize(n_entry_spp = sum(binary, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
eez_sum_raster_30 <- rast(eez_sum_df, crs = "EPSG:4326")
```

```{r}
plot(eez_sum_raster_30)
```

## Try full sample
Going scenario by scenario - starting with RCP45_2050

```{r}
files <- df_threatened$f
aphiaid <- df_threatened$aphiaid
```

```{r}
# Try // - took ~1 min for 906 species
system.time({
  eez_pixels_threatened <- parallel::mclapply(aphiaid, FUN = extract_eez_entry_pixels_df, 
                                            df = df_threatened,
                                            s = "RCP45_2050",
                                            mc.cores = 20)
})
```

```{r}
eez_count_threatened_df <- bind_rows(eez_pixels_threatened)
```

```{r}
# Sum by XY cell, create raster
eez_sum_threatened_df <- eez_count_threatened_df  %>% 
  mutate(scenario = "RCP45_2050") %>% 
  group_by(x, y, GEONAME, scenario) %>% 
  summarize(n_entry_spp = sum(binary, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
eez_sum_threatened_r <- rast(eez_sum_threatened_df, crs = "EPSG:4326")
```

```{r}
plot(eez_sum_threatened_r)
```

```{r}
# Will need to go back and check this - where are those entering Russia EEZ?
ggplot() +
  geom_sf(data = eez) +
  geom_sf(data = eez %>% filter(GEONAME == ""))
  geom_tile(data = eez_sum_threatened_df, aes(x = x, y = y,
                                               fill = n_entry_spp)) +
  scale_fill_viridis_c(limits = c(0,15), direction = -1) +
  theme_bw()
```

```{r}
# Try // - took ~1 min for 906 species
system.time({
  eez_pixels_threatened_45_2100 <- parallel::mclapply(aphiaid, 
                                                      FUN = extract_eez_entry_pixels_df, 
                                                      df = df_threatened,
                                                      s = "RCP45_2100",
                                                      mc.cores = 20)
})
```

```{r}
eez_count_threatened_45_2100_df <- bind_rows(eez_pixels_threatened_45_2100)
```

```{r}
# Sum by XY cell, create raster
eez_sum_threatened_45_2100_df <- eez_count_threatened_45_2100_df  %>% 
  mutate(scenario = "RCP45_2100") %>% 
  group_by(x, y, GEONAME, scenario) %>% 
  summarize(n_entry_spp = sum(binary, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
eez_sum_threatened_45_2100_r <- rast(eez_sum_threatened_45_2100_df, crs = "EPSG:4326")
```

```{r}
# Will need to go back and check this - where are those entering Russia EEZ?
ggplot() +
  geom_sf(data = eez) +
  geom_tile(data = eez_sum_threatened_45_2100_df, aes(x = x, y = y,
                                               fill = n_entry_spp)) +
  scale_fill_viridis_c(limits = c(0,15), direction = -1) +
  theme_bw()
```

```{r}
# Try // - took ~1 min for 906 species
system.time({
  eez_pixels_threatened_85_2050 <- parallel::mclapply(aphiaid, 
                                                      FUN = extract_eez_entry_pixels_df, 
                                                      df = df_threatened,
                                                      s = "RCP85_2050",
                                                      mc.cores = 20)
})
```

```{r}
eez_count_threatened_85_2050_df <- bind_rows(eez_pixels_threatened_85_2050)
```

```{r}
# Sum by XY cell, create raster
eez_sum_threatened_85_2050_df <- eez_count_threatened_85_2050_df  %>% 
  mutate(scenario = "RCP85_2050") %>% 
  group_by(x, y, GEONAME, scenario) %>% 
  summarize(n_entry_spp = sum(binary, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
# Will need to go back and check this - where are those entering Russia EEZ?
ggplot() +
  geom_sf(data = eez) +
  geom_tile(data = eez_sum_threatened_85_2050_df, aes(x = x, y = y,
                                               fill = n_entry_spp)) +
  geom_sf(data = eez %>% filter(GEONAME == "Russian Exclusive Economic Zone"),
          color = "red", fill = NA) +
  scale_fill_viridis_c(limits = c(0,15), direction = -1) +
  theme_bw()
```

```{r}
# Try // - took ~1 min for 906 species
system.time({
  eez_pixels_threatened_85_2100 <- parallel::mclapply(aphiaid, 
                                                      FUN = extract_eez_entry_pixels_df, 
                                                      df = df_threatened,
                                                      s = "RCP85_2100",
                                                      mc.cores = 20)
})
```

```{r}
eez_count_threatened_85_2100_df <- bind_rows(eez_pixels_threatened_85_2100)
```

```{r}
# Sum by XY cell, create raster
eez_sum_threatened_85_2100_df <- eez_count_threatened_85_2100_df  %>% 
  mutate(scenario = "RCP85_2100") %>% 
  group_by(x, y, GEONAME, scenario) %>% 
  summarize(n_entry_spp = sum(binary, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
ggplot() +
  geom_tile(data = eez_sum_threatened_85_2050_df, aes(x = x, y = y,
                                               fill = n_entry_spp,
                                               color = n_entry_spp)) +
  #geom_sf(data = eez %>% filter(GEONAME == "Russian Exclusive Economic Zone"),
          #color = "red", fill = NA) +
  scale_fill_viridis_c(limits = c(0,25), direction = -1) +
  scale_color_viridis_c(limits = c(0,25), direction = -1) +
  geom_sf(data = eez, fill = NA) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r}
russia_85_2100 <- eez_sum_threatened_85_2100_df %>% 
  filter(GEONAME == "Russian Exclusive Economic Zone")
```

```{r}
entry_spp_count_pxl <- rbind(eez_sum_threatened_df,
                             eez_sum_threatened_45_2100_df,
                             eez_sum_threatened_85_2050_df,
                             eez_sum_threatened_85_2100_df)
```

```{r}
ggplot() +
  geom_tile(data = entry_spp_count_pxl, aes(x = x, y = y,
                                               fill = log(n_entry_spp),
                                            color = log(n_entry_spp))) +
  #geom_sf(data = eez %>% filter(GEONAME == "Russian Exclusive Economic Zone"),
          #color = "red", fill = NA) +
  scale_fill_viridis_c( direction = -1) +
  scale_color_viridis_c( direction = -1) +
  facet_wrap(~scenario) +
  geom_sf(data = eez, fill = NA) +
  theme_bw() +
  theme(legend.position= "bottom") +
  labs(fill = "log of new species count by pixel",
       color = "log of new species count by pixel")
```

## Try the same thing but with species leaving pixels 

```{r}
eez_spp_exit <- eez_spp_pxls_long %>% 
  filter(spp_change == "exit from historical")
```


```{r}
extract_eez_exit_pixels_df <- function(id, df, s) {
  spp_filter <- df %>% filter(aphiaid == id)
  
  if (nrow(spp_filter) == 0) {
    warning("No rows found for aphiaid: ", id)
    return(NULL)
  }
  
  spp <- read_parquet(spp_filter$f) %>%
    mutate(aphiaid = id) %>% 
    dplyr::select(x, y, Current, all_of(s), cutoff)
  
  # Filter the df of entry into for this species & scenario
  exit_spp_df <- eez_spp_exit %>% 
    filter(aphiaid == id) %>% 
    filter(scenario == s)
  
  # Get the EEZs for which the species & scenario is entry
  spp_exit_eezs <- unique(exit_spp_df$GEONAME)
  
  # Join the eez dataframe and filter for the spp_entry_eezs
  spp_eez <- inner_join(spp, eez_raster_df) %>% 
    filter(GEONAME %in% spp_exit_eezs) %>% 
    filter(Current >= cutoff) %>% 
    mutate(binary_loss = 1) %>% 
    bind_rows()
  
  return(spp_eez)
}
```

```{r}
# Try // - took ~1 min for 906 species
system.time({
  eez_pixels_exit_45_2050 <- parallel::mclapply(aphiaid, 
                                                      FUN = extract_eez_exit_pixels_df, 
                                                      df = df_threatened,
                                                      s = "RCP45_2050",
                                                      mc.cores = 20)
})
```

```{r}
eez_count_exit_45_2050_df <- bind_rows(eez_pixels_exit_45_2050)
```

```{r}
# Sum by XY cell, create raster
eez_sum_exit_45_2050_df <- eez_count_exit_45_2050_df  %>% 
  mutate(scenario = "RCP45_2050") %>% 
  group_by(x, y, GEONAME, scenario) %>% 
  summarize(n_exit_spp = sum(binary_loss, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
ggplot() +
  geom_tile(data = eez_sum_exit_45_2050_df , aes(x = x, y = y,
                                               fill = n_exit_spp,
                                               color = n_exit_spp)) +
  #geom_sf(data = eez %>% filter(GEONAME == "Russian Exclusive Economic Zone"),
          #color = "red", fill = NA) +
  scale_fill_viridis_c(limits = c(0,20), direction = -1) +
  scale_color_viridis_c(limits = c(0,20), direction = -1) +
  geom_sf(data = eez, fill = NA) +
  theme_bw()+
  theme(legend.position = "bottom")
```

```{r}
# Try // - took ~1 min for 906 species
system.time({
  eez_pixels_exit_45_2100 <- parallel::mclapply(aphiaid, 
                                                      FUN = extract_eez_exit_pixels_df, 
                                                      df = df_threatened,
                                                      s = "RCP45_2100",
                                                      mc.cores = 20)
})
```

```{r}
eez_count_exit_45_2100_df <- bind_rows(eez_pixels_exit_45_2100)
```

```{r}
# Sum by XY cell, create raster
eez_sum_exit_45_2100_df <- eez_count_exit_45_2100_df  %>% 
  mutate(scenario = "RCP45_2100") %>% 
  group_by(x, y, GEONAME, scenario) %>% 
  summarize(n_exit_spp = sum(binary_loss, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
ggplot() +
  geom_tile(data = eez_sum_exit_45_2100_df , aes(x = x, y = y,
                                               fill = n_exit_spp,
                                               color = n_exit_spp)) +
  #geom_sf(data = eez %>% filter(GEONAME == "Russian Exclusive Economic Zone"),
          #color = "red", fill = NA) +
  scale_fill_viridis_c(limits = c(0,20), direction = -1) +
  scale_color_viridis_c(limits = c(0,20), direction = -1) +
  geom_sf(data = eez, fill = NA) +
  theme_bw()+
  theme(legend.position = "bottom")
```

```{r}
# Try // - took ~1 min for 906 species
system.time({
  eez_pixels_exit_85_2050 <- parallel::mclapply(aphiaid, 
                                                      FUN = extract_eez_exit_pixels_df, 
                                                      df = df_threatened,
                                                      s = "RCP85_2050",
                                                      mc.cores = 20)
})
```

```{r}
eez_count_exit_85_2050_df <- bind_rows(eez_pixels_exit_85_2050)
```

```{r}
# Sum by XY cell, create raster
eez_sum_exit_85_2050_df <- eez_count_exit_85_2050_df  %>% 
  mutate(scenario = "RCP85_2050") %>% 
  group_by(x, y, GEONAME, scenario) %>% 
  summarize(n_exit_spp = sum(binary_loss, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
ggplot() +
  geom_tile(data = eez_sum_exit_85_2050_df , aes(x = x, y = y,
                                               fill = n_exit_spp,
                                               color = n_exit_spp)) +
  scale_fill_viridis_c(limits = c(0,20), direction = -1) +
  scale_color_viridis_c(limits = c(0,20), direction = -1) +
  geom_sf(data = eez, fill = NA) +
  theme_bw()+
  theme(legend.position = "bottom")
```
```{r}
# Try // - took ~1 min for 906 species
system.time({
  eez_pixels_exit_85_2100 <- parallel::mclapply(aphiaid, 
                                                      FUN = extract_eez_exit_pixels_df, 
                                                      df = df_threatened,
                                                      s = "RCP85_2100",
                                                      mc.cores = 20)
})
```

```{r}
eez_count_exit_85_2100_df <- bind_rows(eez_pixels_exit_85_2100)
```

```{r}
# Sum by XY cell, create raster
eez_sum_exit_85_2100_df <- eez_count_exit_85_2100_df  %>% 
  mutate(scenario = "RCP85_2100") %>% 
  group_by(x, y, GEONAME, scenario) %>% 
  summarize(n_exit_spp = sum(binary_loss, na.rm = TRUE)) %>% 
  ungroup()
```

```{r}
ggplot() +
  geom_tile(data = eez_sum_exit_85_2100_df , aes(x = x, y = y,
                                               fill = n_exit_spp,
                                               color = n_exit_spp)) +
  scale_fill_viridis_c(limits = c(0,25), direction = -1) +
  scale_color_viridis_c(limits = c(0,25), direction = -1) +
  geom_sf(data = eez, fill = NA) +
  theme_bw()+
  theme(legend.position = "bottom")
```

```{r}
exit_spp_count_pxl <- rbind(eez_sum_exit_45_2050_df,
                             eez_sum_exit_45_2100_df,
                             eez_sum_exit_85_2050_df,
                             eez_sum_exit_85_2100_df)
```

```{r}
ggplot() +
  geom_tile(data = exit_spp_count_pxl, aes(x = x, y = y,
                                               fill = log(n_exit_spp),
                                            color = log(n_exit_spp))) +
  #geom_sf(data = eez %>% filter(GEONAME == "Russian Exclusive Economic Zone"),
          #color = "red", fill = NA) +
  scale_fill_viridis_c(direction = -1) +
  scale_color_viridis_c(direction = -1) +
  facet_wrap(~scenario) +
  geom_sf(data = eez, fill = NA) +
  theme_bw() +
  theme(legend.position= "bottom") +
  labs(fill = "log of loss species count by pixel",
       color = "log of loss species count by pixel")
```

## Test biplot

```{r}
# Merge the entry and exit dataframes
entry_exit_spp_count_pxl <- full_join(entry_spp_count_pxl, exit_spp_count_pxl) 
```

```{r}
entry_exit_spp_sqrt <- entry_exit_spp_count_pxl %>% 
  replace_na(list(n_entry_spp = 0)) %>% 
  replace_na(list(n_exit_spp = 0)) %>% 
  mutate(n_entry_spp_sqrt = sqrt(n_entry_spp),
         n_exit_spp_sqrt = sqrt(n_exit_spp))
```


```{r}
library(biscale)
library(cowplot)
```

```{r}
biplot_data <- bi_class(entry_exit_spp_sqrt, x = n_entry_spp_sqrt, y = n_exit_spp_sqrt,
                        style = "equal", dim = 3)
```

```{r}
biplot_85_2100 <- ggplot() + 
  geom_tile(data = biplot_data, mapping = aes(fill = bi_class,
                                              color = bi_class,
                                              x = x, y = y), show.legend = FALSE) +
  bi_scale_fill(pal = "DkBlue2", dim = 3, rotate_pal = FALSE) +
  bi_scale_color(pal = "DkBlue2", dim = 3, rotate_pal = FALSE) +
  geom_sf(data = eez, fill = NA) +
  labs(x = "",
       y = "",
       subtitle = "RCP 8.5 2100 spp change") +
  bi_theme()
```

```{r}
legend <- bi_legend(pal = "DkBlue2",
                    dim = 3,
                    xlab = "More new spp ",
                    ylab = "More loss spp",
                    size = 8)
```

```{r}
final_plot_85_2100 <- ggdraw() +
  draw_plot(biplot_85_2100, 0, 0, 1, 1) +
  draw_plot(legend, .8, .8, 0.2, 0.2)
```

```{r}
final_plot_85_2100
```


## Plot entry and exit spp

```{r}
change_spp_count_pxl <- entry_exit_spp_count_pxl %>% 
  replace_na(list(n_entry_spp = 0)) %>% 
  replace_na(list(n_exit_spp = 0)) %>% 
  mutate(spp_change = n_entry_spp - n_exit_spp)
```

```{r}
ggplot() +
  geom_tile(data = change_spp_count_pxl, mapping = aes(x = x, y = y,
                                      fill = spp_change,
                                      color = spp_change)) +
  scale_color_distiller(palette = "BrBG", direction = 1, limits = c(-25, 25)) + 
  scale_fill_distiller(palette = "BrBG", direction = 1, limits = c(-25, 25)) +
  facet_wrap(~scenario) +
  geom_sf(data = eez, fill = NA) +
  theme_bw() +
  theme(legend.position= "bottom") +
  labs(fill = "EEZ new species - loss species",
       color = "EEZ new species - loss species")
```

```{r}
ggplot() +
  geom_sf(data = eez, fill = "grey") +
  geom_tile(data = change_spp_count_pxl %>% filter(scenario == "RCP85_2100"),
            mapping = aes(x = x, y = y,
                                      fill = spp_change,
                                      color = spp_change)) +
  scale_color_distiller(palette = "BrBG", direction = 1, limits = c(-25, 25)) + 
  scale_fill_distiller(palette = "BrBG", direction = 1, limits = c(-25, 25)) +
  #scale_color_viridis_c(option = "turbo", limits = c(-25, 25), direction = -1) +
  #scale_fill_viridis_c(option = "turbo", limits = c(-25, 25), direction = -1) +
  facet_wrap(~scenario) +
  geom_sf(data = eez, fill = NA) +
  theme_bw() +
  theme(legend.position= "bottom") +
  labs(fill = "EEZ new species - loss species",
       color = "EEZ new species - loss species")
```
