---
title: "Species entry/exit pixels - sensitivity analysis"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
library(arrow)
```

This document builds on 01_iucn_aquax_eez_overlap.qmd which identifies the number of pixels entering and exiting EEZs for each species. However, in that document, there is no threshold for what qualifies as a species being present in an EEZ. Here, we test thresholds of 10 pixels and 100 pixels to see how that changes the number of species entering/exiting EEZs. We'll use RCP 4.5 2050 here.

```{r}
# Load functions for processing AquaX data
source(here( "common_fxns.R"))
```

```{r}
iucn_lookup <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/data/iucn_redlist/iucn_worms_aphia_id_complete.csv")
```

```{r}
iucn_new_cat <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/data/iucn_redlist/iucn_species_assessments.csv")
```

```{r}
# Spp in EEZs
spp_eez_pxl_count <- read_csv(here("data/iucn_threatened_spp_eez_pxls_052126.csv"))
```

# IUCN threatened species

```{r}
# Filter the iucn_lookup for species with an SDM
iucn_lookup_sdm <- iucn_lookup %>% filter(aquax == "SDM")
```

```{r}
# Filter iucn categories for threatened species (CR, EN, VU)
iucn_threatened <- iucn_new_cat %>% 
  filter(redlist_category %in% c("CR","EN", "VU"))
```

```{r}
# Merge in the iucn_lookup_sdm to get the AphiaID for the threatened species
aphiaid_threatened <- inner_join(iucn_lookup_sdm, iucn_threatened, by = c("iucn_assessment_id" = "assessment_id")) 
```

```{r}
# Get aphiaids for the threatened species
aphiaid_threat <- unique(aphiaid_threatened$aphia_id)
```

## Prep AquaX data
We'll be working with the 1,047 thretened aphiaids

```{r}
# read in meta data on the aquax sample of redlist species
load("/home/shares/data-aquax/iucn_request/meta.Rdata")
```

```{r}
meta_clean <- meta %>% 
  mutate(worms_name = tolower(ScientificName)) %>% 
  select(!ScientificName) %>% 
  # simplify for now for easier viewing
  select(AphiaID, SPID, worms_name, GROUP, NR_CR, range_pixels, SDM)
```

```{r}
spp_fs <- list.files('/home/shares/data-aquax/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(meta, by = c('aphiaid' = 'AphiaID'))
```

```{r}
# Filter for the AphiaID of the threatend species - get 987 species in AquaX data (even though the 1,047 species were in the iucn_lookup_sdm with AquaX SDMs)
df_threatened <- df %>% 
  filter(aphiaid %in% aphiaid_threat)
```

```{r}
files <- df_threatened$f
aphiaid <- df_threatened$aphiaid
```

## EEZ 

```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
eez_simple <- eez |>
  select(MRGID, GEONAME, geom)
```

```{r}
# Create a blank raster matching AquaX resolution (EEZs should be large enough to just do 0.05 degree raster)
eez_raster_blank <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05,
                  crs = "EPSG:4326")
```

```{r}
eez_raster_geoname <- rasterize(eez_simple, eez_raster_blank, field = "GEONAME")
```

```{r}
eez_raster_mrgid <- rasterize(eez_simple, eez_raster_blank, field = "MRGID")
```

# Identify which EEZs have species entering/exiting - different thresholds

```{r}
# Organize dataframe so that's it's easier to select a scenario and compare to current
eez_spp_pxls_long_1 <- spp_eez_pxl_count %>% 
  pivot_longer(cols = starts_with("n_RCP"),
               names_to = "scenario",
               values_to = "pxl_count") %>% 
  mutate(scenario = str_remove(scenario, "n_"),
        pxl_thres = 1) %>% 
  mutate(spp_change = case_when(is.na(n_current) &  pxl_count > 0 ~ 'entry into new',
                            n_current > 0  & is.na(pxl_count) ~ 'exit from historical',
                            is.na(n_current)  & is.na(pxl_count) == 0 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

```{r}
# Organize dataframe so that's it's easier to select a scenario and compare to current
eez_spp_pxls_long_10 <- spp_eez_pxl_count %>% 
  pivot_longer(cols = starts_with("n_RCP"),
               names_to = "scenario",
               values_to = "pxl_count") %>% 
  mutate(scenario = str_remove(scenario, "n_"),
        pxl_thres = 10) %>% 
  mutate(spp_change = case_when(n_current < 10 &  pxl_count >= 10 ~ 'entry into new',
                            n_current >= 10  & pxl_count < 10 ~ 'exit from historical',
                            n_current < 10  & pxl_count < 10 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

```{r}
# Organize dataframe so that's it's easier to select a scenario and compare to current
eez_spp_pxls_long_100 <- spp_eez_pxl_count %>% 
  pivot_longer(cols = starts_with("n_RCP"),
               names_to = "scenario",
               values_to = "pxl_count") %>% 
  mutate(scenario = str_remove(scenario, "n_"),
        pxl_thres = 100) %>% 
  mutate(spp_change = case_when(n_current < 100 &  pxl_count >= 100 ~ 'entry into new',
                            n_current >= 100  & pxl_count < 100 ~ 'exit from historical',
                            n_current < 100  & pxl_count < 100 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

```{r}
# Combine dataframea and compare summary stats
eez_spp_pxls_long <- bind_rows(eez_spp_pxls_long_1, eez_spp_pxls_long_10, eez_spp_pxls_long_100) %>% 
    filter(scenario == "RCP85_2050") %>%
    group_by(pxl_thres, scenario, spp_change) %>%
    summarise(n_spp_x_eez = n())
```

```{r}
# Plot number of species entering/exiting EEZs by threshold - fewer species entering/exiting under 10 threshold than under 100 threshold
ggplot(eez_spp_pxls_long %>% 
    filter(spp_change != "continued presence"), aes(x = as.character(pxl_thres), y = n_spp_x_eez, fill = spp_change)) +
  geom_col(position = "dodge") +
  labs(x = "Pixel threshold for presence in EEZ", y = "Number of species x EEZs", fill = "Change in presence",
  title = "Species presence change - RCP 8.5 2050") +
  theme_minimal()
```

# Identify pixels entering/exiting EEZs - 10 pixel threshold
```{r}
# Get counts of entry into new by EEZ for 10 pixel threshold
eez_spp_entry_10 <- eez_spp_pxls_long_10 %>% 
  filter(spp_change == "entry into new")
```

```{r}
# Create dataframe of EEZ raster
eez_raster_df <- as.data.frame(eez_raster_geoname, xy = TRUE)
```

```{r}
# Try // - took ~1 min for 987 species
system.time({
  eez_pixels_threatened_10 <- parallel::mclapply(aphiaid, FUN = extract_eez_entry_pixels_df, 
                                            df = df_threatened,
                                            s = "RCP45_2050",
                                            eez_spp_entry = eez_spp_entry_10,
                                            eez_raster_df = eez_raster_df,
                                            mc.cores = 20)
})
```