---
title: "Additional plots"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
library(arrow)
library(patchwork)
library(cowplot)
library(collapse)
```

This document creates some more focused plots for the data calculated in 01_iucn_aquax_eez_overlap_db.qmd and 02_iucn_aquax_rangearea.qmd.

## Read in threatened species data

```{r}
# Spp in EEZs
spp_eez_pxl_count <- read_csv(here("data/iucn_threatened_spp_eez_pxls_061626.csv"))
```

```{r}
# Spp EEZ entry/exit pixels
spp_entry_pxls <- read_rds(here("data/iucn_atrisk_10pxlthres_entry_pxls_072826.rds"))
spp_exit_pxls <- read_rds(here("data/iucn_atrisk_10pxlthres_exit_pxls_072926.rds"))
```

```{r}
# Species range area
area_df <- read_csv(here("data/iucn_threatened_spp_range_area_061826.csv"))
```

## EEZs

```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
eez <- st_read('/Users/abowd/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
# Get list of EEZ IDs that are not overlapping claims (i.e., POL_TYPE = 200NM)
eez_200nm_list <- eez %>% 
  filter(POL_TYPE == "200NM") %>% 
  pull(MRGID)
```

```{r}
eez_simple <- eez %>% 
  dplyr::select(MRGID, GEONAME) %>% 
  st_drop_geometry() %>% 
  filter(MRGID %in% eez_200nm_list)

eez_simple_sf <- eez %>% 
  dplyr::select(MRGID, GEONAME) %>% 
  filter(MRGID %in% eez_200nm_list)
```

```{r}
# Create a blank raster matching AquaX resolution (EEZs should be large enough to just do 0.05 degree raster)
eez_raster_blank <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05,
                  crs = "EPSG:4326")
```

```{r}
eez_raster_mrgid <- rasterize(eez_simple_sf, eez_raster_blank, field = "MRGID")
```

```{r}
# Create dataframe of EEZ raster
eez_raster_df <- as.data.frame(eez_raster_mrgid, xy = TRUE) %>% 
  # Round so that the X and Y match the AquaX data (which is rounded to 3 decimal places)
  mutate(x = round(x, 3),
         y = round(y, 3))
```

```{r}
ggplot() +
  geom_sf(data = eez_simple_sf) +
  coord_sf(expand = FALSE) +
  geom_tile(data = eez_raster_df, aes(x = x, y = y,
                              fill = MRGID,
                              color = MRGID)) 
```

## Try out rainbow plot by IUCN status

```{r}
iucn_lookup <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/_data_raw/iucn_redlist_2026/iucn_worms_aphia_id_complete.csv")
```

```{r}
iucn_cat <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/_data_raw/iucn_redlist_2026/iucn_species_assessments.csv")
```

```{r}
# Add aphiaid to iucn_cat
iucn_cat_aphiaid <- left_join(iucn_cat, (iucn_lookup %>% dplyr::select(iucn_assessment_id, aphia_id)),
 by = c("assessment_id" = "iucn_assessment_id")) %>% 
  rename(aphiaid = aphia_id) 
```

```{r}
# Merge in iucn_cat to get group for each species
spp_eez_pxl_count_group <- left_join(spp_eez_pxl_count, iucn_cat_aphiaid)
```

```{r}
# What's the count of each group 
spp_eez_group_count <- spp_eez_pxl_count_group %>% 
  mutate(general_group = case_when(group %in% c("hagfishes", "sturgeons", "tunas_and_billfishes", "tarpons_and_ladyfishes")~ "pelagic_fishes",
                                   group %in% c("angelfishes", "blennies", "butterfly_fishes", "groupers", "pufferfishes", "seabreams_porgies_picarels", "surgeonfishes", "wrasses_and_parrotfishes") ~ "reef_fishes",
                                   #group %in% c("marine_turtles", "seasnakes") ~ "marine_turtles_and_seasnakes",
                                   TRUE ~ group)) %>% 
  group_by(general_group) %>% 
  summarise(n_spp = length(unique(aphiaid))) %>% 
  ungroup()
```

```{r}
# Get count of number of EEZ by group and count of IUCN cat by group (create figure similar to bar chart)
group_eez_count <- spp_eez_pxl_count_group %>% 
  # Filter for 
  filter(MRGID %in% eez_200nm_list) %>%
  group_by(group, aphiaid, redlist_category) %>% 
  summarize(current_eez_count = sum(n_current >= 10, na.rm = TRUE),
            RCP45_2050_eez_count = sum(n_RCP45_2050 >= 10, na.rm = TRUE),
            RCP45_2100_eez_count = sum(n_RCP45_2100 >= 10, na.rm = TRUE),
            RCP85_2050_eez_count = sum(n_RCP85_2050 >= 10, na.rm = TRUE),
            RCP85_2100_eez_count = sum(n_RCP85_2100 >= 10, na.rm = TRUE)) %>% 
  ungroup()

group_iucncat_count <- group_eez_count %>% 
  group_by(group, redlist_category) %>% 
  summarize(redlist_count = n()) %>% 
  ungroup()
```


```{r}
ggplot() +
  geom_density(data = group_eez_count, aes(x = current_eez_count, fill = redlist_category), alpha = 0.5) +
  labs(title = "Distribution of current EEZ counts by IUCN redlist category",
       x = "current EEZ count") +
  facet_wrap(~redlist_category) +
  theme_bw()
```



```{r}
# Assign EEZ count thresholds based on plotted density (drop not comprehensively assessed)
group_eez_count_threshold <- group_eez_count %>% 
  #filter(group != "not comprehensively assessed") %>%
  # Drop those with current and RCP85_2050 EEZ counts of 0 (i.e., not in any EEZ)
  filter(current_eez_count != 0 | RCP85_2050_eez_count != 0) %>% 
  mutate(current_eez_count_cat = case_when(current_eez_count == 1 ~ "1",
                                          current_eez_count == 2 ~ "2",
                                           current_eez_count > 2 & current_eez_count <= 5 ~ "3-5",
                                           current_eez_count > 5 & current_eez_count <= 15 ~ "6-15",
                                           current_eez_count > 15 & current_eez_count <= 25 ~ "16-25",
                                           current_eez_count > 25 & current_eez_count <= 50 ~ "26-50",
                                           current_eez_count > 50 & current_eez_count <= 100 ~ "51-100",
                                           current_eez_count > 100 ~ ">100",
                                           TRUE ~ "0"),
         RCP85_2050_eez_count_cat = case_when(RCP85_2050_eez_count == 1 ~ "1",
                                            RCP85_2050_eez_count == 2 ~ "2",
                                           RCP85_2050_eez_count > 2 & RCP85_2050_eez_count <= 5 ~ "3-5",
                                           RCP85_2050_eez_count > 5 & RCP85_2050_eez_count <= 15 ~ "6-15",
                                           RCP85_2050_eez_count > 15 & RCP85_2050_eez_count <= 25 ~ "16-25",
                                           RCP85_2050_eez_count > 25 & RCP85_2050_eez_count <= 50 ~ "26-50",
                                           RCP85_2050_eez_count > 50 & RCP85_2050_eez_count <= 100 ~ "51-100",
                                           RCP85_2050_eez_count > 100 ~ ">100",
                                           TRUE ~ "0")) %>%

  mutate(general_group = case_when(group %in% c("hagfishes", "sturgeons", "tunas_and_billfishes", "tarpons_and_ladyfishes")~ "pelagic_fishes",
                                   group %in% c("angelfishes", "blennies", "butterfly_fishes", "groupers", "pufferfishes", "seabreams_porgies_picarels", "surgeonfishes", "wrasses_and_parrotfishes") ~ "reef_fishes",
                                   #group %in% c("marine_turtles", "seasnakes") ~ "marine_turtles_and_seasnakes",
                                   TRUE ~ group))
```

```{r}
cat_summary <- group_eez_count_threshold %>% 
  group_by(redlist_category, current_eez_count_cat) %>% 
  summarise(n_cat = n()) %>% 
  ungroup()

# Get total number in each group
group_count <- group_eez_count_threshold %>% 
  group_by(redlist_category) %>% 
  summarise(n_group = n()) %>% 
  ungroup()

# Merge in group count to get percentage
cat_summary_perc <- left_join(cat_summary, group_count) %>% 
  mutate(perc_cat = n_cat/n_group * 100)
```

```{r}
group_eez_count_threshold$current_eez_count_cat <- factor(group_eez_count_threshold$current_eez_count_cat, 
levels = c("0", "1", "2", "3-5", "6-15", "16-25", "26-50", "51-100", ">100"))

group_eez_count_threshold$RCP85_2050_eez_count_cat <- factor(group_eez_count_threshold$RCP85_2050_eez_count_cat, 
levels = c("0", "1", "2", "3-5", "6-15", "16-25", "26-50", "51-100", ">100"))

group_eez_count_threshold$redlist_category <- factor(group_eez_count_threshold$redlist_category,
levels = c("CR", "EN", "VU", "NT"))
```

```{r}
library(forcats)
# Plot a bar graph by group with fill as the EEZ count category
ggplot() +
  geom_bar(data = group_eez_count_threshold, aes(y = fct_rev(redlist_category), fill = fct_rev(current_eez_count_cat)), position = "fill") +
  scale_fill_viridis_d() +
  labs(title = "Current EEZ count category by IUCN category",
       y = "",
       x = "proportion",
       fill = "EEZ count") +
  theme_bw()
```

```{r}
# Plot a bar graph by group with fill as the EEZ count category
ggplot() +
  geom_bar(data = group_eez_count_threshold, aes(y = fct_rev(redlist_category), fill = fct_rev(RCP85_2050_eez_count_cat)), position = "fill") +
  scale_fill_viridis_d() +
  labs(title = "RCP 8.5 2050 EEZ count category by IUCN category",
       y = "",
       x = "proportion",
       fill = "EEZ count") +
  theme_bw()
```

## Set up dataframe so that it's easier to plot current vs future side by side for each group 
```{r}
eez_count_group_simple <- group_eez_count_threshold %>% 
  dplyr::select(redlist_category, general_group, current_eez_count_cat, RCP85_2050_eez_count_cat) %>% 
  pivot_longer(cols = c(current_eez_count_cat, RCP85_2050_eez_count_cat), names_to = "scenario", values_to = "eez_count_cat") %>% 
  mutate(scenario = str_replace(scenario, "current_eez_count_cat", "current"),
         scenario = str_replace(scenario, "RCP85_2050_eez_count_cat", "RCP85_2050"))
```

```{r}
eez_count_summary_taxon <- eez_count_group_simple %>%
  filter(general_group != "not comprehensively assessed") %>%
  count(scenario, general_group, eez_count_cat, name = "n") %>%
  group_by(scenario, general_group) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

eez_count_summary_iucn <- eez_count_group_simple %>%
  count(scenario, redlist_category, eez_count_cat, name = "n") %>%
  group_by(scenario, redlist_category) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()
```

```{r}
# Plot a bar graph by group with fill as the EEZ count category
taxon_bar_plot <- ggplot() +
  geom_bar(data = eez_count_group_simple %>% 
  filter(general_group != "not comprehensively assessed"), aes(y = fct_rev(general_group), fill = fct_rev(eez_count_cat)), position = "fill") +
  scale_fill_viridis_d(option = "turbo") +
  labs(title = "",
       y = "",
       x = "proportion",
       fill = "EEZ count") +
  theme_bw() + 
  theme(legend.position = "bottom") +
  facet_wrap(~scenario)
```

```{r}
# Plot a bar graph by group with fill as the EEZ count category
iucn_bar_plot <- ggplot() +
  geom_bar(data = eez_count_group_simple, aes(y = fct_rev(redlist_category), fill = fct_rev(eez_count_cat)), position = "fill") +
  scale_fill_viridis_d(option = "turbo") +
  labs(title = "",
       y = "",
       x = "proportion",
       fill = "EEZ count") +
  theme_bw() + 
  theme(legend.position = "bottom") +
  facet_wrap(~scenario)
```

```{r}
group_eez_count_diff <- group_eez_count_threshold %>% 
  mutate(eez_count_diff = RCP85_2050_eez_count - current_eez_count) #%>% 
  #dplyr::select(redlist_category, aphiaid, current_eez_count,RCP85_2050_eez_count, eez_count_diff)
```

```{r}
taxon_jitter_plot_only <- ggplot() +
  geom_jitter(data = group_eez_count_diff %>% filter(general_group != "not comprehensively assessed"),
   aes(x = eez_count_diff, y = fct_rev(general_group), alpha = 0.6),
   color = "gray10", position = position_jitterdodge(dodge.width = 0, jitter.width = .6),) +
  scale_y_discrete() + 
  #scale_x_continuous(limits = c(-25, 10)) +
  labs(title = "EEZ region counts by IUCN comprehensive group",
       x = "EEZ region count change\n(RCP8.5 2050 - current)",
       y = "") +
  theme_bw() +
  theme(legend.position = "none")
```

```{r}
iucn_jitter_plot_only <- ggplot() +
  geom_jitter(data = group_eez_count_diff, aes(x = eez_count_diff, y = fct_rev(redlist_category), alpha = 0.6),
   color = "gray10", position = position_jitterdodge(dodge.width = 0, jitter.width = .6),) +
  scale_y_discrete() + 
  #scale_x_continuous(limits = c(-25, 10)) +
  labs(title = "EEZ region counts by IUCN redlst category",
       x = "EEZ region count change\n(RCP8.5 2050 - current)",
       y = "") +
  theme_bw() +
  theme(legend.position = "none")
```

```{r}
# Save only the jitter plot
ggsave(here("figures/iucn_threatened_200nmonly_taxon_jitterplot_072126.png"), taxon_jitter_plot_only, width = 7, height = 5, dpi = 300)

ggsave(here("figures/iucn_threatened_200nmonly_iucncat_jitterplot_072126.png"), iucn_jitter_plot_only, width = 7, height = 5, dpi = 300)
```

```{r}
# Save only the bar plot
ggsave(here("figures/iucn_threatened_200nmonly_taxon_barplot_072126.png"), taxon_bar_plot, width = 7, height = 5, dpi = 300)

ggsave(here("figures/iucn_threatened_200nmonly_iucncat_barplot_072126.png"), iucn_bar_plot, width = 7, height = 5, dpi = 300)
```

## Get list of species entries by EEZ

```{r}
# Organize dataframe so that's it's easier to select a scenario and compare to current
eez_spp_pxls_long <- spp_eez_pxl_count %>% 
  pivot_longer(cols = starts_with("n_RCP"),
               names_to = "scenario",
               values_to = "pxl_count") %>% 
  mutate(scenario = str_remove(scenario, "n_")) %>% 
  # Replace NA with 0, otherwise will be assigned to "continued presence"
  replace_na(list(pxl_count = 0, n_current = 0)) %>% 
  # Set 10 pixel threshold (at least ten pixels need in EEZ for species presence)
  mutate(spp_change = case_when(n_current < 10 &  pxl_count >= 10 ~ 'entry into new',
                            n_current >= 10  & pxl_count < 10 ~ 'exit from historical',
                            n_current < 10  & pxl_count < 10 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

```{r}
# Which species have the most EEZ region entries?
spp_eez_entry_count <- eez_spp_pxls_long %>% 
  filter(MRGID %in% eez_200nm_list) %>%
  filter(spp_change == "entry into new") %>% 
  group_by(aphiaid, scenario) %>% 
  summarise(n_eez_entry = n()) %>% 
  ungroup()

# Which species have the most EEZ region exits?
spp_eez_exit_count <- eez_spp_pxls_long %>% 
  filter(MRGID %in% eez_200nm_list) %>%
  filter(spp_change == "exit from historical") %>% 
  group_by(aphiaid, scenario) %>% 
  summarise(n_eez_exit = n()) %>% 
  ungroup()

# Get list of EEZ exits 100 and over
spp_exit_100 <- spp_eez_exit_count %>% 
  filter(n_eez_exit >= 100,
         scenario == "RCP85_2050")

spp_exit_100_list <- spp_exit_100$aphiaid
```

```{r}
# Area of 100 + species
area_df_100 <- area_df %>% 
  filter(aphiaid %in% spp_exit_100_list) %>% 
  filter(scenario == "RCP85_2050") %>%
  mutate(perc_of_current = scenario_area_km2 / current_area_km2 * 100)
```

```{r}
# Merge in IUCN info
spp_eez_entry_count_iucn <- left_join(spp_eez_entry_count, iucn_cat_aphiaid) %>% 
  filter(group != "not comprehensively assessed") %>%
  mutate(general_group = case_when(group %in% c("hagfishes", "sturgeons", "tunas_and_billfishes", "tarpons_and_ladyfishes")~ "pelagic_fishes",
                                   group %in% c("angelfishes", "blennies", "butterfly_fishes", "groupers", "pufferfishes", "seabreams_porgies_picarels", "surgeonfishes", "wrasses_and_parrotfishes") ~ "reef_fishes",
                                  group %in% c("marine_turtles", "seasnakes") ~ "marine_reptiles",
                                   TRUE ~ group)) %>% 
  group_by(general_group, scenario, n_eez_entry) %>% 
  summarize(n_spp = n()) 
```

```{r}
spp_entry_summary <- spp_eez_entry_count_iucn %>% 
  group_by(scenario, n_eez_entry) %>% 
  summarize(n_spp = sum(n_spp))
```

```{r}
ggplot() +
  geom_bar(data = spp_eez_entry_count_iucn %>% filter(scenario == "RCP85_2050"),
            aes(x = n_eez_entry, y = n_spp, fill = general_group), stat = "identity") +
  scale_fill_viridis_d(option = "turbo", direction = -1) +
  scale_x_sqrt(breaks = c(1, 5,10, 20, 30, 40, 50)) +
  #scale_x_continuous(breaks = c(1, 2, 3, 4, 5, 10, 20, 30, 40, 50)) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(title = "Number EEZs entered by IUCN comprehensive group",
       x = "Total number of EEZs entered",
       y = "Species count",
       fill = "")
```

```{r}
# Merge in IUCN info
spp_eez_exit_count_iucn <- left_join(spp_eez_exit_count, iucn_cat_aphiaid) %>% 
  filter(group != "not comprehensively assessed") %>%
  mutate(general_group = case_when(group %in% c("hagfishes", "sturgeons", "tunas_and_billfishes", "tarpons_and_ladyfishes")~ "pelagic_fishes",
                                   group %in% c("angelfishes", "blennies", "butterfly_fishes", "groupers", "pufferfishes", "seabreams_porgies_picarels", "surgeonfishes", "wrasses_and_parrotfishes") ~ "reef_fishes",
                                   group %in% c("marine_turtles", "seasnakes") ~ "marine_reptiles",
                                   TRUE ~ group)) %>% 
  group_by(general_group, scenario, n_eez_exit) %>% 
  summarize(n_spp = n())
```

```{r}
spp_exit_summary <- spp_eez_exit_count_iucn %>% 
  group_by(scenario, n_eez_exit) %>% 
  summarize(n_spp = sum(n_spp))
```

```{r}
ggplot() +
  geom_bar(data = spp_eez_exit_count_iucn %>% filter(scenario == "RCP85_2050"),
            aes(x = n_eez_exit, y = n_spp, fill = general_group), stat = "identity") +
  scale_fill_viridis_d(option = "turbo", direction = -1) +
  scale_x_sqrt(breaks = c(1, 10, 20, 40, 80, 120, 160)) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(title = "Number of EEZs left by IUCN comprehensive group",
       x = "Total number of EEZs exited",
       y = "Species count",
       fill = "")
  
```

```{r}
spp_eez_entry_clean <- spp_eez_entry_count_iucn %>% 
  filter(scenario == "RCP85_2050") %>% 
  mutate(n_eez_entry_group = ifelse(n_eez_entry >= 10, "10+", as.character(n_eez_entry)))
```

```{r}
spp_eez_entry_clean$n_eez_entry_group <- factor(spp_eez_entry_clean$n_eez_entry_group, 
levels = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10+"))
```

```{r}
ggplot() +
  geom_bar(data = spp_eez_entry_clean,
            aes(x = n_eez_entry_group, y = n_spp, fill = general_group), stat = "identity") +
  scale_fill_viridis_d(option = "turbo", direction = -1) +
  scale_x_discrete(breaks = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10+")) +
  theme_minimal() +
  theme(legend.position = "right") +
  labs(title = "Number EEZs entered by IUCN comprehensive group",
       x = "Total number of EEZs entered",
       y = "Species count",
       fill = "")
  
```


# For species exit bar graph, lump 20+ into one bar
```{r}
spp_eez_exit_clean <- spp_eez_exit_count_iucn %>% 
  filter(scenario == "RCP85_2050") %>% 
  mutate(n_eez_exit_group = ifelse(n_eez_exit >= 10, "10+", as.character(n_eez_exit)))
```

```{r}
spp_eez_exit_clean$n_eez_exit_group <- factor(spp_eez_exit_clean$n_eez_exit_group, 
levels = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20+"))
```

```{r}
spp_eez_exit_clean$n_eez_exit_group <- factor(spp_eez_exit_clean$n_eez_exit_group, 
levels = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10+"))
```

```{r}
ggplot() +
  geom_bar(data = spp_eez_exit_clean,
            aes(x = n_eez_exit_group, y = n_spp, fill = general_group), stat = "identity") +
  scale_fill_viridis_d(option = "turbo", direction = -1) +
  scale_x_discrete(breaks = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10+")) +
  theme_minimal() +
  theme(legend.position = "right") +
  labs(title = "Number of EEZs left by IUCN comprehensive group",
       x = "Total number of EEZs exited",
       y = "Species count",
       fill = "")
  
```

## Both species entries and exits on faceted plot

```{r}
spp_eez_exit_entry_clean <- rbind(spp_eez_exit_clean %>% 
                                  dplyr::select(general_group, scenario, n_eez_exit_group, n_spp) %>% 
                                  mutate(change = "Exits") %>% 
                                  rename(n_eez_group = n_eez_exit_group),
                                  spp_eez_entry_clean %>% 
                                  dplyr::select(general_group, scenario, n_eez_entry_group, n_spp) %>% 
                                  mutate(change = "Entries") %>% 
                                  rename(n_eez_group = n_eez_entry_group))
```

```{r}
spp_eez_exit_entry_clean$n_eez_group <- factor(spp_eez_exit_entry_clean$n_eez_group, 
levels = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10+"))
```

```{r}
entry_exit_barplot <- ggplot() +
  geom_bar(data = spp_eez_exit_entry_clean,
            aes(x = n_eez_group, y = n_spp, fill = general_group), stat = "identity") +
  scale_fill_viridis_d(option = "turbo", direction = -1) +
  scale_x_discrete(breaks = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10+")) +
  theme_minimal() +
  facet_wrap(~change) + 
  theme(legend.position = "bottom") +
  labs(title = "EEZ entries and exits by IUCN comprehensive group",
       x = "EEZ count",
       y = "Species count",
       fill = "")
  
```

```{r}
# Save only the bar plot
ggsave(here("figures/iucn_atrisk_taxon_entry_exit_073026.png"), entry_exit_barplot, width = 7, height = 5, dpi = 300)

```

## Plot North Korea Species Entries
```{r}
library(rnaturalearth)
north_korea <- ne_countries(country = "North Korea", scale = "large", returnclass = "sf")
south_korea <- ne_countries(country = "South Korea", scale = "large", returnclass = "sf")
japan <- ne_countries(country = "Japan", scale = "large", returnclass = "sf")
```

```{r}
nk_entry_pixels <- spp_entry_pxls %>% 
  filter(MRGID == 8328) 

nk_entry_rcp85_2050 <- nk_entry_pixels %>% 
  filter(scenario == "RCP85_2050")
```

```{r}
sk_entry_rcp85_2050 <- spp_entry_pxls %>% 
  filter(scenario == "RCP85_2050") %>% 
  filter(MRGID == 8327) 
```

```{r}
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8328 )) +
  coord_sf(expand = FALSE) +
  geom_tile(data = nk_entry_rcp85_2050, aes(x = x, y = y,
                                               fill = n_entry_spp,
                                               color = n_entry_spp)) +
  scale_fill_viridis_c(option = "turbo",
                        limits = c(1,21), 
                        breaks = c(1, 5, 10, 15, 21)) +
  scale_color_viridis_c(option = "turbo",
                        limits = c(1,21), 
                        breaks = c(1, 5, 10, 15, 21)) +
  geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5) +
  labs(title = "North Korea EEZ species entries (RCP8.5 2050)",
       fill = "Species entry count",
       color = "Species entry count") +
  theme_minimal() 
```

```{r}
ggplot() +
  geom_sf(data = eez %>% filter(MRGID %in% c(8328, 8327))) +
  coord_sf(expand = FALSE) +
  geom_tile(data = nk_entry_rcp85_2050, aes(x = x, y = y,
                                               fill = n_entry_spp,
                                               color = n_entry_spp)) +
    geom_tile(data = sk_entry_rcp85_2050, aes(x = x, y = y,
                                               fill = n_entry_spp,
                                               color = n_entry_spp)) +
  scale_fill_viridis_c(option = "turbo",
                        limits = c(1,21), 
                        breaks = c(1, 5, 10, 15, 21)) +
  scale_color_viridis_c(option = "turbo",
                        limits = c(1,21), 
                        breaks = c(1, 5, 10, 15, 21)) +
  geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5) +
  geom_sf(data = south_korea, fill = "cornsilk2", color = "black", size = 0.5) +
  labs(title = "North & South Korea EEZ species entries (RCP8.5 2050)",
       fill = "Species entry count",
       color = "Species entry count") +
  theme_minimal() 
```


```{r}
# Which species has less than 10 pixels in current and the largest number of pixels in NK in RCP 8.5 2050?
nk_spp_future <- eez_spp_pxls_long %>% 
  filter(MRGID == 8328) %>% 
  filter(scenario == "RCP85_2050") %>%
  filter(spp_change == "entry into new") 

# Merge in IUCN info
nk_spp_future_iucn <- left_join(nk_spp_future, iucn_cat_aphiaid) %>% 
  #filter(group != "not comprehensively assessed") %>%
  mutate(general_group = case_when(group %in% c("hagfishes", "sturgeons", "tunas_and_billfishes", "tarpons_and_ladyfishes")~ "pelagic_fishes",
                                   group %in% c("angelfishes", "blennies", "butterfly_fishes", "groupers", "pufferfishes", "seabreams_porgies_picarels", "surgeonfishes", "wrasses_and_parrotfishes") ~ "reef_fishes",
                                   group %in% c("marine_turtles", "seasnakes") ~ "marine_reptiles",
                                   TRUE ~ group)) %>% 
  arrange(desc(pxl_count))
```

```{r}
# Which species has less than 10 pixels in current and the largest number of pixels in Russia in RCP 8.5 2050?
russia_spp_future <- eez_spp_pxls_long %>% 
  filter(MRGID == 5690) %>% 
  filter(scenario == "RCP85_2050") %>%
  filter(spp_change == "entry into new") 

# Merge in IUCN info
russia_spp_future_iucn <- left_join(russia_spp_future, iucn_cat_aphiaid) %>% 
  #filter(group != "not comprehensively assessed") %>%
  mutate(general_group = case_when(group %in% c("hagfishes", "sturgeons", "tunas_and_billfishes", "tarpons_and_ladyfishes")~ "pelagic_fishes",
                                   group %in% c("angelfishes", "blennies", "butterfly_fishes", "groupers", "pufferfishes", "seabreams_porgies_picarels", "surgeonfishes", "wrasses_and_parrotfishes") ~ "reef_fishes",
                                   group %in% c("marine_turtles", "seasnakes") ~ "marine_reptiles",
                                   TRUE ~ group)) %>% 
  arrange(desc(pxl_count))
```

```{r}
# Are any russia species entries also in North Korea? (i.e., same aphiaid)
russia_nk_spp_overlap <- inner_join(russia_spp_future_iucn, nk_spp_future_iucn, by = "aphiaid")
```

```{r}
# Are any of russia species in the spp_exit_100_list? 
russia_exit_100 <- russia_spp_future_iucn %>% 
  filter(aphiaid %in% spp_exit_100_list)

#NK
nk_exit_100 <- nk_spp_future_iucn %>% 
  filter(aphiaid %in% spp_exit_100_list)
```

```{r}
# What about the Singapore EEZ? (MRGID = 8485)
singapore_spp_future <- eez_spp_pxls_long %>% 
  filter(MRGID == 8485) %>%
  filter(scenario == "RCP85_2050") %>%
  filter(spp_change == "entry into new")

singapore_spp_future_iucn <- left_join(singapore_spp_future, iucn_cat_aphiaid) %>% 
  #filter(group != "not comprehensively assessed") %>%
  mutate(general_group = case_when(group %in% c("hagfishes", "sturgeons", "tunas_and_billfishes", "tarpons_and_ladyfishes")~ "pelagic_fishes",
                                   group %in% c("angelfishes", "blennies", "butterfly_fishes", "groupers", "pufferfishes", "seabreams_porgies_picarels", "surgeonfishes", "wrasses_and_parrotfishes") ~ "reef_fishes",
                                   group %in% c("marine_turtles", "seasnakes") ~ "marine_reptiles",
                                   TRUE ~ group)) %>% 
  arrange(desc(pxl_count))
```

```{r}
singapore_exit_100 <- singapore_spp_future_iucn %>% 
  filter(aphiaid %in% spp_exit_100_list)
```

## Top species exits 

```{r}
# Christmas Island
christmas_island_spp <- eez_spp_pxls_long %>% 
  filter(MRGID == 8309) %>%
  filter(scenario == "RCP85_2050") %>%
  filter(spp_change == "exit from historical")

ci_spp_exit_cat <- left_join(christmas_island_spp, iucn_cat_aphiaid) %>% 
  #filter(group != "not comprehensively assessed") %>%
  mutate(general_group = case_when(group %in% c("hagfishes", "sturgeons", "tunas_and_billfishes", "tarpons_and_ladyfishes")~ "pelagic_fishes",
                                   group %in% c("angelfishes", "blennies", "butterfly_fishes", "groupers", "pufferfishes", "seabreams_porgies_picarels", "surgeonfishes", "wrasses_and_parrotfishes") ~ "reef_fishes",
                                   group %in% c("marine_turtles", "seasnakes") ~ "marine_reptiles",
                                   TRUE ~ group)) %>% 
  arrange(desc(n_current))

ci_exit_100 <- ci_spp_exit_cat %>% 
  filter(aphiaid %in% spp_exit_100_list)
```

```{r}
# Howland & Baker Islands - US
hb_island_spp <- eez_spp_pxls_long %>% 
  filter(MRGID == 8451) %>%
  filter(scenario == "RCP85_2050") %>%
  filter(spp_change == "exit from historical")

hb_spp_exit_cat <- left_join(hb_island_spp, iucn_cat_aphiaid) %>% 
  #filter(group != "not comprehensively assessed") %>%
  mutate(general_group = case_when(group %in% c("hagfishes", "sturgeons", "tunas_and_billfishes", "tarpons_and_ladyfishes")~ "pelagic_fishes",
                                   group %in% c("angelfishes", "blennies", "butterfly_fishes", "groupers", "pufferfishes", "seabreams_porgies_picarels", "surgeonfishes", "wrasses_and_parrotfishes") ~ "reef_fishes",
                                   group %in% c("marine_turtles", "seasnakes") ~ "marine_reptiles",
                                   TRUE ~ group)) %>% 
  arrange(desc(n_current))

hb_exit_100 <- hb_spp_exit_cat %>% 
  filter(aphiaid %in% spp_exit_100_list)
```

```{r}
# Bassas da India - France
bdi_spp <- eez_spp_pxls_long %>% 
  filter(MRGID == 8340) %>%
  filter(scenario == "RCP85_2050") %>%
  filter(spp_change == "exit from historical")

bdi_spp_exit_cat <- left_join(bdi_spp, iucn_cat_aphiaid) %>% 
  #filter(group != "not comprehensively assessed") %>%
  mutate(general_group = case_when(group %in% c("hagfishes", "sturgeons", "tunas_and_billfishes", "tarpons_and_ladyfishes")~ "pelagic_fishes",
                                   group %in% c("angelfishes", "blennies", "butterfly_fishes", "groupers", "pufferfishes", "seabreams_porgies_picarels", "surgeonfishes", "wrasses_and_parrotfishes") ~ "reef_fishes",
                                   group %in% c("marine_turtles", "seasnakes") ~ "marine_reptiles",
                                   TRUE ~ group)) %>% 
  arrange(desc(n_current))

bdi_exit_100 <- bdi_spp_exit_cat %>% 
  filter(aphiaid %in% spp_exit_100_list)
```

```{r}
hbi_exit_rcp85_2050 <- spp_exit_pxls %>% 
  filter(scenario == "RCP85_2050") %>% 
  filter(MRGID == 8451) 
```
```{r}
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8451 )) +
  coord_sf(expand = FALSE) +
  geom_tile(data = hbi_exit_rcp85_2050, aes(x = x, y = y,
                                               fill = n_exit_spp,
                                               color = n_exit_spp)) +
  scale_fill_viridis_c(option = "turbo",
                        limits = c(1,40),
                        breaks = c(1, 10, 20, 30, 40)) +
  scale_color_viridis_c(option = "turbo",
                        limits = c(1,40),
                        breaks = c(1, 10, 20, 30, 40)) +
  #geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5) +
  labs(title = "Howland & Baker EEZ region species exits (RCP8.5 2050)",
       fill = "Species exit count",
       color = "Species exit count") +
  theme_minimal() 
```

```{r}
ci_exit_rcp85_2050 <- spp_exit_pxls %>% 
  filter(scenario == "RCP85_2050") %>% 
  filter(MRGID == 8309) 
```
```{r}
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8309)) +
  coord_sf(expand = FALSE) +
  geom_tile(data = ci_exit_rcp85_2050, aes(x = x, y = y,
                                               fill = n_exit_spp,
                                               color = n_exit_spp)) +
  scale_fill_viridis_c(option = "turbo",
                        limits = c(1,40),
                        breaks = c(1, 10, 20, 30, 40)) +
  scale_color_viridis_c(option = "turbo",
                        limits = c(1,40),
                        breaks = c(1, 10, 20, 30, 40)) +
  #geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5) +
  labs(title = "Christmas Island EEZ region species exits (RCP8.5 2050)",
       fill = "Species exit count",
       color = "Species exit count") +
  theme_minimal() 
```

```{r}
ci_exit_rcp85_2050 <- spp_exit_pxls %>% 
  filter(scenario == "RCP85_2050") %>% 
  filter(MRGID == 8309) 
```
```{r}
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8309)) +
  coord_sf(expand = FALSE) +
  geom_tile(data = ci_exit_rcp85_2050, aes(x = x, y = y,
                                               fill = n_exit_spp,
                                               color = n_exit_spp)) +
  scale_fill_viridis_c(option = "turbo",
                        limits = c(1,40),
                        breaks = c(1, 10, 20, 30, 40)) +
  scale_color_viridis_c(option = "turbo",
                        limits = c(1,40),
                        breaks = c(1, 10, 20, 30, 40)) +
  #geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5) +
  labs(title = "Christmas Island EEZ region species exits (RCP8.5 2050)",
       fill = "Species exit count",
       color = "Species exit count") +
  theme_minimal() 
```

```{r}
bdi_exit_rcp85_2050 <- spp_exit_pxls %>% 
  filter(scenario == "RCP85_2050") %>% 
  filter(MRGID == 8340) 
```
```{r}
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8340)) +
  coord_sf(expand = FALSE) +
  geom_tile(data = bdi_exit_rcp85_2050, aes(x = x, y = y,
                                               fill = n_exit_spp,
                                               color = n_exit_spp)) +
  scale_fill_viridis_c(option = "turbo",
                        limits = c(1,40),
                        breaks = c(1, 10, 20, 30, 40)) +
  scale_color_viridis_c(option = "turbo",
                        limits = c(1,40),
                        breaks = c(1, 10, 20, 30, 40)) +
  #geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5) +
  labs(title = "Bassas da India EEZ region species exits (RCP8.5 2050)",
       fill = "Species exit count",
       color = "Species exit count") +
  theme_minimal() 
```

## Mapping specific taxons in specific EEZs
Both NK and Howland & Baker Islands have a lot of reef-builing coral entries and exits, respectively. Let's take a closer look. I suspect that the hotspots here are small area concentrations of corals. 

```{r}
# Run species entry pixels but just for corals
source(here( "common_fxns.R"))
```

# Read in AquaX files 

```{r}
# read in meta data on the aquax species
load("/home/shares/data-aquax/META_02122025.Rdata")
```

```{r}
meta_clean <- META %>% 
  mutate(worms_name = tolower(ScientificName)) %>% 
  select(!ScientificName) %>% 
  # simplify for now for easier viewing
  select(AphiaID, SPID, worms_name, GROUP, NR_CR, range_pixels, SDM)
```

```{r}
spp_fs <- list.files('/home/shares/data-aquax/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(META, by = c('aphiaid' = 'AphiaID'))
```

# NK
```{r}
# Get aphiaids for the reef building corals entering NK
nk_coral <- nk_spp_future_iucn %>% 
  filter(general_group == "reef_building_corals")

nk_coral_aphiaid <- nk_coral$aphiaid
```

```{r}
# Only include NK in species in entry df
nk_spp_entry <- eez_spp_pxls_long %>% 
  filter(spp_change == "entry into new") %>% 
  filter(MRGID == 8328)
```

```{r}
df_nk_corals <- df %>% 
  filter(aphiaid %in% nk_coral_aphiaid)
```

```{r}
# Run NK coral species (don't need parallel)
system.time({
  coral_nk_entry_rcp85_list <- lapply(nk_coral_aphiaid, FUN = extract_eez_entry_pixels_df_db, 
                                       df = df_nk_corals, 
                                       s = "RCP85_2050",
                                       eez_spp_entry = nk_spp_entry,
                                       eez_raster_df = eez_raster_df)
})
```

```{r}
coral_nk_entry_rcp85_df  <- bind_rows(coral_nk_entry_rcp85_list) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse
system.time({
coral_nk_entry_rcp85_sum_df  <- coral_nk_entry_rcp85_df  %>% 
  fgroup_by(MRGID, x, y) %>% 
  fcount() %>% 
  mutate(scenario = "RCP85_2050")
})
```

```{r}
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8328 )) +
  coord_sf(expand = FALSE) +
  geom_tile(data = coral_nk_entry_rcp85_sum_df, aes(x = x, y = y,
                                               fill = N,
                                               color = N)) +
  scale_fill_viridis_c(option = "turbo",
                        breaks = c(1, 3, 6, 9, 12)) +
  scale_color_viridis_c(option = "turbo",
                        breaks = c(1, 3, 6, 9, 12)) +
  geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5) +
  labs(title = "North Korea EEZ coral species entries (RCP8.5 2050)",
       fill = "Species entry count",
       color = "Species entry count") +
  theme_minimal() 
```

# HBI

```{r}
# Get aphiaids for the reef building corals exiting Howland and Baker Islands
hbi_coral <- hb_spp_exit_cat%>% 
  filter(general_group == "reef_building_corals")

hbi_coral_aphiaid <- hbi_coral$aphiaid
```

```{r}
# Only include HBI in species in exit df
hbi_spp_exit <- eez_spp_pxls_long %>% 
  filter(spp_change == "exit from historical") %>% 
  filter(MRGID == 8451)
```

```{r}
df_hbi_corals <- df %>% 
  filter(aphiaid %in% hbi_coral_aphiaid)
```

```{r}
# Run HBI coral species (don't need parallel)
system.time({
  coral_hbi_exit_rcp85_list <- lapply(hbi_coral_aphiaid, FUN = extract_eez_exit_pixels_df_db, 
                                       df = df_hbi_corals, 
                                       s = "RCP85_2050",
                                       eez_spp_exit = hbi_spp_exit,
                                       eez_raster_df = eez_raster_df)
})
```

```{r}
coral_hbi_exit_rcp85_df  <- bind_rows(coral_hbi_exit_rcp85_list) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse
system.time({
coral_hbi_exit_rcp85_sum_df  <- coral_hbi_exit_rcp85_df  %>% 
  fgroup_by(MRGID, x, y) %>% 
  fcount() %>% 
  mutate(scenario = "RCP85_2050")
})
```

```{r}
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8451 )) +
  coord_sf(expand = FALSE) +
  geom_tile(data = coral_hbi_exit_rcp85_sum_df, aes(x = x, y = y,
                                               fill = N,
                                               color = N)) +
  scale_fill_viridis_c(option = "turbo",
                        limits = c(1,26),
                        breaks = c(1, 5, 10, 15, 20, 26)) +
  scale_color_viridis_c(option = "turbo",
                        limits = c(1,26),
                        breaks = c(1, 5, 10, 15, 20, 26)) +
  #geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5) +
  labs(title = "Howland & Baker EEZ region coral species exits (RCP8.5 2050)",
       fill = "Species exit count",
       color = "Species exit count") +
  theme_minimal() 
```

## Izu stingray (Hemitrygon izuensis) range
```{r}
# Let's look at the range of the Izu stingrayy (Hemitrygon izuensis)
izu <- read_parquet("/home/shares/data-aquax/SDM/FINAL_EMSDM_EMMEAN_SP_1042848.parquet")
```


```{r}
# Izu cutoff is 444 so filter RCP 8.5 2050 to only include pixels with values greater than 444
izu_rcp85_2050 <- izu %>% 
  mutate(RCP85_2050_suitable = ifelse(RCP85_2050 >= 444, "suitable", "not suitable")) 
```

```{r}
izu_rcp85_2050_suitable <- rbind(izu %>%  mutate(RCP85_2050_suitable = "all"), (izu_rcp85_2050 %>% 
  filter(RCP85_2050_suitable == "suitable")))
```

```{r}
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8328)) +
  coord_sf(expand = FALSE) +
  geom_tile(data = izu_rcp85_2050_suitable, aes(x = x, y = y,
                              fill = RCP85_2050,
                              color = RCP85_2050)) +
  scale_fill_viridis_c(option = "mako") +
  scale_color_viridis_c(option = "mako") +
  geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5) +
  geom_sf(data = south_korea, fill = "cornsilk2", color = "black", size = 0.5) +
  geom_sf(data = japan, fill = "cornsilk2", color = "black", size = 0.5) +
  labs(title = "Izu stingray range (RCP8.5 2050)",
       subtitle = "Suitability threshold is 444",
       fill = "Suitability value",
       color = "Suitability value") +
  facet_wrap(~RCP85_2050_suitable) +
  theme_minimal() +
  theme(legend.position = "bottom")
```


```{r}
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8328)) +
  coord_sf(expand = FALSE) +
  geom_tile(data = izu_rcp85_2050_suitable %>% 
  filter(RCP85_2050_suitable == "suitable"), aes(x = x, y = y,
                              fill = RCP85_2050,
                              color = RCP85_2050)) +
  scale_fill_viridis_c(option = "mako") +
  scale_color_viridis_c(option = "mako") +
  scale_y_continuous(limits = c(37, 43)) +
  scale_x_continuous(limits = c(122, 134)) +
  geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5)+
  labs(title = "Izu stingray range (RCP8.5 2050)",
       subtitle = "Suitability threshold is 444",
       fill = "Suitability value",
       color = "Suitability value") +
  theme_minimal()
```

## Thrasher shark (cutoff is 560)

```{r}
ts <- read_parquet("/home/shares/data-aquax/SDM/FINAL_EMSDM_EMMEAN_SP_220019.parquet")
```

```{r}
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8328)) +
  coord_sf(expand = FALSE) +
  geom_tile(data = ts %>% 
  filter(RCP85_2050 >= 560), aes(x = x, y = y,
                              fill = RCP85_2050,
                              color = RCP85_2050)) +
  scale_fill_viridis_c(option = "mako") +
  scale_color_viridis_c(option = "mako") +
  scale_y_continuous(limits = c(37, 43)) +
  scale_x_continuous(limits = c(122, 134)) +
  geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5)+
  labs(title = "Thrasher shark range (RCP8.5 2050)",
       subtitle = "Suitability threshold is 560",
       fill = "Suitability value",
       color = "Suitability value") +
  theme_minimal()
```

## Test out function with NK data to see what's causing horizontal lines in the plot
```{r}
# From common_fxns.R
# Join the eez dataframe and filter for the spp_entry_eezs
  spp_eez <- inner_join(spp, eez_raster_df) %>% 
    filter(GEONAME %in% spp_entry_eezs) %>% 
    filter(.data[[s]] >= cutoff) %>% 
    mutate(binary = 1) %>% 
    bind_rows()
```


```{r}
# Join eez_raster_df and thrasher shark data, filter for NK EEZ and RCP8.5 2050
nk_ts <- inner_join(ts, eez_raster_df) %>%
  filter(MRGID == 8328) %>%
  filter(RCP85_2050 >= cutoff) %>% 
  mutate(binary = 1) 
```

```{r}
ts_filtered <- ts %>% 
  filter(RCP85_2050 >= cutoff) %>% 
  filter(x >= 122 & x <= 134) %>%
  filter(y >= 37 & y <= 43)
```

```{r}
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8328)) +
  coord_sf(expand = FALSE) +
  geom_tile(data = nk_ts, aes(x = x, y = y,
                              fill = RCP85_2050,
                              color = RCP85_2050)) +
  scale_fill_viridis_c(option = "mako") +
  scale_color_viridis_c(option = "mako") +
  #scale_y_continuous(limits = c(37, 43)) +
  #scale_x_continuous(limits = c(122, 134)) +
  geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5)+
  labs(title = "Thrasher shark range (RCP8.5 2050)",
       subtitle = "Suitability threshold is 560",
       fill = "Suitability value",
       color = "Suitability value") +
  theme_minimal()
```


```{r}
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8328)) +
  coord_sf(expand = FALSE) +
  geom_tile(data = eez_raster_df %>% filter(MRGID == 8328), aes(x = x, y = y)) +
  geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5)+
  theme_minimal()
```


```{r}
# Check alignment
# Check if origins match (ts_test_raster is e-15 off from 0, 0 center, so rounding X and Y should fix stripes)
origin(eez_raster_blank)
ts_test_raster <- rast(ts, type = "xyz", crs = "EPSG:4326")  # if ts has a regular grid
origin(ts_test_raster)
res(ts_test_raster)
```

```{r}
# See if rounding fixes problem
eez_raster_df_round <- eez_raster_df %>% 
  mutate(x = round(x, 3),
         y = round(y, 3))

ts_round <- ts %>% 
  mutate(x = round(x, 3),
         y = round(y, 3))
```

```{r}
nk_ts_round <- inner_join(ts_round, eez_raster_df_round, by = c("x", "y")) %>%
  filter(MRGID == 8328) %>%
  filter(RCP85_2050 >= cutoff) %>% 
  mutate(binary = 1)
```

```{r}
ggplot() +
  geom_sf(data = eez %>% filter(MRGID == 8328)) +
  coord_sf(expand = FALSE) +
  geom_tile(data = nk_ts_round, aes(x = x, y = y,
                              fill = RCP85_2050,
                              color = RCP85_2050)) +
  scale_fill_viridis_c(option = "mako") +
  scale_color_viridis_c(option = "mako") +
  #scale_y_continuous(limits = c(37, 43)) +
  #scale_x_continuous(limits = c(122, 134)) +
  geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5)+
  labs(title = "Thrasher shark range (RCP8.5 2050) - rounded",
       subtitle = "Suitability threshold is 560",
       fill = "Suitability value",
       color = "Suitability value") +
  theme_minimal()
```

```{r}
# Get counts of entry into new, exit from historical etc. by EEZ
eez_spp_change <- eez_spp_pxls_long %>% 
  filter(MRGID %in% eez_200nm_list) %>%
  group_by(GEONAME, scenario, spp_change) %>% 
  summarize(spp_change_count = n()) %>% 
  ungroup()
```