---
title: "Untitled"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
library(arrow)
library(patchwork)
library(cowplot)
library(collapse)
```

This document creates some more focused plots for the data calculated in 01_iucn_aquax_eez_overlap_db.qmd and 02_iucn_aquax_rangearea.qmd. It saves figures for RCP 4.5 2050.

## Read in threatened species data

```{r}
# Spp in EEZs
spp_eez_pxl_count <- read_csv(here("data/iucn_threatened_spp_eez_pxls_061626.csv"))
```

```{r}
# Spp EEZ entry/exit pixels
spp_entry_pxls <- read_rds(here("data/iucn_atrisk_10pxlthres_entry_pxls_072826.rds"))
spp_exit_pxls <- read_rds(here("data/iucn_atrisk_10pxlthres_exit_pxls_072926.rds"))
```

```{r}
# Species range area
area_df <- read_csv(here("data/iucn_threatened_spp_range_area_061826.csv"))
```

## EEZs

```{r}
eez <- st_read('/home/shares/usgs-cap-rangeshifts/spatial/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
eez <- st_read('/Users/abowd/World_EEZ_v12_20231025_gpkg/eez_v12.gpkg')
```

```{r}
# Get list of EEZ IDs that are not overlapping claims (i.e., POL_TYPE = 200NM)
eez_200nm_list <- eez %>% 
  filter(POL_TYPE == "200NM") %>% 
  pull(MRGID)
```

```{r}
eez_simple <- eez %>% 
  dplyr::select(MRGID, GEONAME) %>% 
  st_drop_geometry() %>% 
  filter(MRGID %in% eez_200nm_list)

eez_simple_sf <- eez %>% 
  dplyr::select(MRGID, GEONAME) %>% 
  filter(MRGID %in% eez_200nm_list)
```

```{r}
# Create a blank raster matching AquaX resolution (EEZs should be large enough to just do 0.05 degree raster)
eez_raster_blank <- rast(xmin = -180,
                  xmax = 180,
                  ymin = -90, 
                  ymax = 90,
                  resolution = 0.05,
                  crs = "EPSG:4326")
```

```{r}
eez_raster_mrgid <- rasterize(eez_simple_sf, eez_raster_blank, field = "MRGID")
```

```{r}
# Create dataframe of EEZ raster
eez_raster_df <- as.data.frame(eez_raster_mrgid, xy = TRUE) %>% 
  # Round so that the X and Y match the AquaX data (which is rounded to 3 decimal places)
  mutate(x = round(x, 3),
         y = round(y, 3))
```

# Pixel map hotspots

```{r}
library(rnaturalearth)
globe <- ne_countries(scale = "large", returnclass = "sf")
```

```{r}
# Sea of Japan
sea_of_japan <- ggplot() +
  geom_sf(data = globe, fill = "cornsilk2", size = 0.1, color = "black")+
  geom_sf(data = eez_simple_sf, size = 0.01) +
  coord_sf(expand = FALSE) +
  geom_tile(data = spp_entry_pxls_45_2050, aes(x = x, y = y,
                                               fill = n_entry_spp,
                                               color = n_entry_spp)) +
  
  scale_y_continuous(limits = c(29, 48)) +
  scale_x_continuous(limits = c(119, 148)) +
  scale_fill_viridis_c(limits = c(1, 16), 
                      breaks = c(1, 2, 4, 8, 16),
                      direction = -1,
                      trans = "sqrt", # visually stretches the lower value data
                      guide = guide_colorbar(
                      #barwidth  = unit(5, "cm"),   # wide bar for bottom placement
                      #barheight = unit(0.4, "cm"),
                      ticks = TRUE,
                      ticks.colour = "white",
                      ticks.linewidth = 0.8
  )) +
  scale_color_viridis_c(limits = c(1, 16), 
                        breaks = c(1, 2, 4, 8, 16),
                        direction = -1,
                        trans = "sqrt", # visually stretches the lower value data
                        guide = "none")  +
  geom_sf(data = eez_simple_sf, fill = NA, color = "black", size = 0.01) +
  theme_void() + 
  labs(fill = "species entry count",
       color = "species entry count",
       title = "Sea of Japan",
       #subtitle = "SSP2-4.5, 2050",
       x = "",
       y = "") +
  theme(legend.position = "bottom",
        plot.margin = margin(l = 5)) 
```

#Find the species moving into North Korea, South Korea, and Russia and plot these

```{r}
# Organize dataframe so that's it's easier to select a scenario and compare to current
eez_spp_pxls_long <- spp_eez_pxl_count %>% 
  pivot_longer(cols = starts_with("n_RCP"),
               names_to = "scenario",
               values_to = "pxl_count") %>% 
  mutate(scenario = str_remove(scenario, "n_")) %>% 
  # Replace NA with 0, otherwise will be assigned to "continued presence"
  mutate(pxl_count = replace_na(pxl_count, 0),
         n_current = replace_na(n_current, 0)) %>% 
  # Set 10 pixel threshold (at least ten pixels need in EEZ for species presence)
  mutate(spp_change = case_when(n_current < 10 &  pxl_count >= 10 ~ 'entry into new',
                            n_current >= 10  & pxl_count < 10 ~ 'exit from historical',
                            n_current < 10  & pxl_count < 10 ~ 'no presence',
                            TRUE ~ 'continued presence'))
```

```{r}
# 65 species entering the Sea of Japan EEZ (MRGID 8328, 8327, 5690) under RCP4.5 in 2050 (tricky with Russia being such a large EEZ)
sea_japan_entry_spp <- eez_spp_pxls_long %>% 
  filter(MRGID %in% c(8328, 8327, 5690)) %>%
  filter(scenario == "RCP45_2050") %>% 
  filter(spp_change == "entry into new") 

```

```{r}
# 19 species entering the Baltic Sea EEZ (MRGID 5694, 5690, 5684, 5683, 5675, 5676, 5669, 5687, 5674) under RCP4.5 in 2050 (tricky with Russia being such a large EEZ)
baltic_entry_spp <- eez_spp_pxls_long %>% 
  filter(MRGID %in% c(5694, 5690, 5684, 5683, 5675, 5676, 5669, 5687, 5674)) %>%
  filter(MRGID != 5690) %>% # remove russia (doesn't look like there are any species entries on map)
  filter(scenario == "RCP45_2050") %>% 
  filter(spp_change == "entry into new") 

```

# Read in AquaX files 

```{r}
# read in meta data on the aquax species
load("/home/shares/data-aquax/META_02122025.Rdata")
```

```{r}
meta_clean <- META %>% 
  mutate(worms_name = tolower(ScientificName)) %>% 
  select(!ScientificName) %>% 
  # simplify for now for easier viewing
  select(AphiaID, SPID, worms_name, GROUP, NR_CR, range_pixels, SDM)
```

```{r}
spp_fs <- list.files('/home/shares/data-aquax/SDM',
                     full.names = TRUE)

df <- data.frame(f = spp_fs) %>%
    mutate(fsize = file.size(f),
           aphiaid = str_extract(basename(f), '[0-9]+') %>% as.integer()) %>%
    left_join(META, by = c('aphiaid' = 'AphiaID'))
```

## Sea of Japan - entry species 
Plot the current range of the species entering the Sea of Japan region

```{r}
# Get aphiaids for the reef building corals entering Sea of Japan
sea_japan_aphiaid <- unique(sea_japan_entry_spp$aphiaid)
```

```{r}
df_sea_japan <- df %>% 
  filter(aphiaid %in% sea_japan_aphiaid)

```

We can use the least concern count function to map the ranges of all the reef building corals entering Sea of Japan.

```{r}
# Load functions for processing AquaX data
source(here( "common_fxns.R"))
```

```{r}
# Run Sea of Japan coral current species (don't need parallel)
system.time({
  sea_japan_all_current_list <- lapply(sea_japan_aphiaid, 
                                       FUN = least_concern_count, 
                                       df = df_sea_japan, 
                                       s = "Current",
                                       n_threads = 1)
})
```

```{r}
sea_japan_all_current_df <- bind_rows(sea_japan_all_current_list) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse (4 seconds)
system.time({
sea_japan_all_current_sum_df <- sea_japan_all_current_df %>% 
  fgroup_by(x, y) %>% 
  fcount()
})
```

```{r}
ggplot() +
  geom_sf(data = globe, fill = "cornsilk2", size = 0.1, color = "black")+
  geom_sf(data = eez_simple_sf, size = 0.01) +
  coord_sf(expand = FALSE) +
  geom_tile(data = sea_japan_all_current_sum_df, aes(x = x, y = y, fill = N, color = N)) +
  scale_y_continuous(limits = c(29, 48)) +
  scale_x_continuous(limits = c(119, 148)) +
  scale_fill_viridis_c(option = "viridis",
direction = -1,
                      limits = c(1, 56),
                      breaks = c(1, 6, 12, 18, 24, 30, 56)) +
  scale_color_viridis_c(option = "viridis",
                      direction = -1,
                      limits = c(1, 56),
                      breaks = c(1, 6, 12, 18, 24, 30, 56)) +
  #scale_y_continuous(limits = c(37, 43)) +
  #scale_x_continuous(limits = c(122, 134)) +
  #geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5)+
  labs(title = "Current range of species entering Sea of Japan EEZs",
       fill = "Count",
       color = "Count") +
  theme_minimal()
```

```{r}
# Run Sea of Japan coral current species (don't need parallel)
system.time({
  sea_japan_all_future_list <- lapply(sea_japan_aphiaid, 
                                       FUN = least_concern_count, 
                                       df = df_sea_japan, 
                                       s = "RCP45_2050",
                                       n_threads = 1)
})
```

```{r}
sea_japan_all_future_df <- bind_rows(sea_japan_all_future_list) %>% 
  # drop binary = 0 (no pixels above threshold)
  filter(binary == 1)
```

```{r}
# Sum by XY cell, using collapse (3 seconds)
system.time({
sea_japan_all_future_sum_df <- sea_japan_all_future_df %>% 
  fgroup_by(x, y) %>% 
  fcount()
})
```

```{r}
ggplot() +
  geom_sf(data = globe, fill = "cornsilk2", size = 0.1, color = "black")+
  geom_sf(data = eez_simple_sf, size = 0.01) +
  coord_sf(expand = FALSE) +
  geom_tile(data = sea_japan_all_future_sum_df, aes(x = x, y = y, fill = N, color = N)) +
  scale_y_continuous(limits = c(29, 48)) +
  scale_x_continuous(limits = c(119, 148)) +
  scale_fill_viridis_c(option = "viridis",
                        direction = -1,
                      limits = c(1, 56),
                      breaks = c(1, 6, 12, 18, 24, 30, 56)) +
  scale_color_viridis_c(option = "viridis",
                        direction = -1,
                      limits = c(1, 56),
                      breaks = c(1, 6, 12, 18, 24, 30, 56)) +
  #scale_y_continuous(limits = c(37, 43)) +
  #scale_x_continuous(limits = c(122, 134)) +
  #geom_sf(data = north_korea, fill = "cornsilk2", color = "black", size = 0.5)+
  labs(title = "Future range of species entering Sea of Japan EEZs",
       fill = "Count",
       color = "Count") +
  theme_minimal()
```