---
title: "IUCN threat category"
format: 
  html:
    toc: true
    number-sections: true
    embed-resources: true
execute:
    message: false
    warning: false
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(terra)
library(arrow)
library(patchwork)
library(cowplot)
library(collapse)
```

This document finds the threat categories listed by the IUCN for the at-risk species in this analysis (1,323)

## Read in threatened species data

```{r}
# Spp in EEZs
spp_eez_pxl_count <- read_csv(here("data/iucn_threatened_spp_eez_pxls_061626.csv"))
```

```{r}
# Spp EEZ entry/exit pixels
spp_entry_pxls <- read_rds(here("data/iucn_atrisk_1pxlthres_entry_pxls_091626.rds"))
spp_exit_pxls <- read_rds(here("data/iucn_atrisk_1pxlthres_exit_pxls_091626.rds"))
```

```{r}
# Species range area
area_df <- read_csv(here("data/iucn_threatened_spp_range_area_061826.csv"))
```


## IUCN info

```{r}
iucn_lookup <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/ab9cd09931a4ce5c12d0bcd2984595d6522ed158/_data/iucn_redlist_2026/iucn_worms_aphia_id_complete.csv")
```

```{r}
iucn_cat <- read_csv("https://raw.githubusercontent.com/nceas-range-shifts/chi_shifts/refs/heads/main/_data/iucn_redlist_2026/iucn_species_assessments.csv")
```

```{r}
# Add aphiaid to iucn_cat
iucn_cat_aphiaid <- left_join(iucn_cat, (iucn_lookup %>% dplyr::select(iucn_assessment_id, aphia_id)),
 by = c("assessment_id" = "iucn_assessment_id")) %>% 
  rename(aphiaid = aphia_id) 
```

```{r}
# Merge in iucn_cat to get group for each species
spp_eez_pxl_count_group <- left_join(spp_eez_pxl_count, iucn_cat_aphiaid)
```

## IUCN threats
For those species experiencing range shifts, what are the most common IUCN threats? Are there particular threats that are more common for species experiencing range loss vs gain?
https://www.iucnredlist.org/resources/threat-classification-scheme

According to the IUCN, threats marked with timing as “ongoing”, severity as “rapid” or “very rapid” declines and scopes of “majority” or “whole” are typically considered major threats to a species. IUCN stopped using scores in 2022 due to lack of sceintific method.

```{r}
library(iucnredlist) ### remotes::install_github("IUCN-UK/iucnredlist")
```

```{r}
api_key <- Sys.getenv("IUCN_API_KEY") ### usethis::edit_r_environ() to add key to R environment file
api <- init_api(api_key)
```

```{r}
# Get IUCN ID data 
iucn_id_data <- unique(spp_eez_pxl_count_group$assessment_id)
```

```{r}
# still need to run this through assesment_data_many
iucn_id_data <- assessment_data_many(api, iucn_id_data, wait_time = 0.5)
```

```{r}
# run element_names() to see what categories can be extracted
iucn_threats <- extract_element(iucn_id_data, "threats")
```

```{r}
# Group by threat code and count the unique number of assessment_ids in each
iucn_threat_codes_summary <- iucn_threats %>% 
    group_by(code, description) %>% 
    summarize(n = length(unique(assessment_id)))
```


```{r}
# Save data
write_csv(iucn_threat_codes_summary, here("data/iucn_atrisk_spp_threat_codes_100226.csv"))
```